### 2023-03-04 ms/ms.tex
ms

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers

\draftfalse

\journalname{Journal of Advances in Modeling Earth Systems (JAMES)}

\begin{document}

\title{The Community Land Model, version 5: \\ Parameter Perturbation Experiment}

\authors{D. Kennedy \affil{1}, Katherine Dagon\affil{1}, David M. Lawrence\affil{1}}
%R.A. Fisher
%B.M. Sanderson
%C.D. Koven
%A.L.S. Swann
%C.M. Zarakas
%L.R. Hawkins
%F. Hoffman
%N. Collier
%Y. Luo
%X. Lu
%S. Levis
%S.C. Swenson
%K.W. Oleson
%E. Kluzek
%J. Liu

\affiliation{1}{Climate and Global Dynamics Laboratory, NCAR, Boulder, CO, USA.}

\correspondingauthor{Daniel Kennedy}{djk2120@ucar.edu}

\begin{keypoints}
\item enter point 1 here
\item enter point 2 here
\item enter point 3 here
\end{keypoints}

\begin{abstract}
[ enter your Abstract here ]
\end{abstract}

\section{Introduction}
\section{Experiment Description}
\subsection{Model description}
The Community Terrestrial System Model (CTSM) is developed by the CESM Land Model Working Group (LMWG) and maintained at the National Center for Atmospheric Research (NCAR). This experiment utilizes the Community Land Model configuration of CTSM, version 5.1 (CLM5.1). The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.

Relative to CLM5.0, version 5.1 includes minor bug fixes, several parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}. The PPE experiment required additional code modifications to identify and programmatically vary the full suite of model parameters. The exact model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}). We utilized the full biogeochemistry version of CLM in land-only mode, with the crop model turned off. The component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP}

\subsection{Model spin-up}
Model spin-up for the equilibration of carbon and nitrogen pools within biogeochemistry-enabled land models can consume up to 98\% of computational time \cite{sun2023}.

\cite{lu2020}

\subsection{Sparsegrid}

\cite{hoffman2013}

\subsection{Experimental Design}
\subsection{Forcing Scenarios}
\label{sect:exps}
 \begin{table}[h]
 \caption{Forcing Scenarios}
 \centering
 \begin{tabular}{l c c c c}
 \hline
  Name  & Meteorology & CO$_2$ (ppmv) & N addition & Description \\
 \hline
   CTL2010  & 2005-2014 & 367 & - & control experiment\\
   C285        & 2005-2014 & 285 & - & low CO$_2$ \\
   C867        & 2005-2014 & 867 & - & high CO$_2$ \\
   AF1855    & 1851-1860 & 367 & - & pre-industrial climate \\
   AF2095    & 2091-2100 & 367 & - & late century climate (SSP3-7.0) \\
   NDEP      & 2005-2014 & 367 & 5g/m$^2$ & enhanced nitrogen deposition \\
 \hline
 \end{tabular}
 \label{tab:exps}
 \end{table}

\subsection{Analyses}
\section{Results}
\section{Discussion}
\section{Conclusion}
%Text here ===>>>

%%

%  Numbered lines in equations:
%  To add line numbers to lines in equations,
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

%% Enter Figures and Tables near as possible to where they are first mentioned:
%
% DO NOT USE \psfrag or \subfigure commands.
%
% Figure captions go below the figure.
% Acronyms used in figure captions will be spelled out in the final, published version.

% Table titles go above tables;  other caption information
%  should be placed in last line of the table, using
% \multicolumn2l{$^a$ This is a table note.}
% NOTE that there is no difference between table caption and table heading in the final, published version
%
%----------------
% EXAMPLE FIGURES
%
% \begin{figure}
% \includegraphics{example.png}
% \caption{caption}
% \end{figure}
%
% Giving latex a width will help it to scale the figure properly. A simple trick is to use \textwidth. Try this if large figures run off the side of the page.
% \begin{figure}
% \noindent\includegraphics[width=\textwidth]{anothersample.png}
%\caption{caption}
%\label{pngfiguresample}
%\end{figure}
%
%
% If you get an error about an unknown bounding box, try specifying the width and height of the figure with the natwidth and natheight options. This is common when trying to add a PDF figure without pdflatex.
% \begin{figure}
% \noindent\includegraphics[natwidth=800px,natheight=600px]{samplefigure.pdf}
%\caption{caption}
%\label{pdffiguresample}
%\end{figure}
%
%
% PDFLatex does not seem to be able to process EPS figures. You may want to try the epstopdf package.
%

%
% ---------------
% EXAMPLE TABLE
%
% \begin{table}
% \caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
% \centering
% \begin{tabular}{l c}
% \hline
%  Run  & Time (min)  \\
% \hline
%   $l1$  & 260   \\
%   $l2$  & 300   \\
%   $l3$  & 340   \\
%   $h1$  & 270   \\
%   $h2$  & 250   \\
%   $h3$  & 380   \\
%   $r1$  & 370   \\
%   $r2$  & 390   \\
% \hline
% \multicolumn{2}{l}{$^{a}$Footnote text here.}
% \end{tabular}
% \end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% SIDEWAYS FIGURES and TABLES
% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
%
% \begin{sidewaysfigure}
% \includegraphics[width=20pc]{figsamp}
% \caption{caption here}
% \label{newfig}
% \end{sidewaysfigure}
%
%  \begin{sidewaystable}
%  \caption{Caption here}
% \label{tab:signif_gap_clos}
%  \begin{tabular}{ccc}
% one&two&three\\
% four&five&six
%  \end{tabular}
%  \end{sidewaystable}

%% If using numbered lines, please surround equations with \begin{linenomath*}...\end{linenomath*}
%\begin{linenomath*}
%\begin{equation}
%y|{f} \sim g(m, \sigma),
%\end{equation}
%\end{linenomath*}

%%% End of body of article

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional Appendices go here
%
% The \appendix command resets counters and redefines section heads
%
% After typing \appendix
%
%\section{Here Is Appendix Title}
% will show
% A: Here Is Appendix Title
%
%\appendix
%\section{Here is a sample appendix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Optional Glossary, Notation or Acronym section goes here:
%
% Glossary is only allowed in Reviews of Geophysics
%  \begin{glossary}
%  \term{Term}
%   Term Definition here
%  \term{Term}
%   Term Definition here
%  \term{Term}
%   Term Definition here
%  \end{glossary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acronyms
%% NOTE that acronyms in the final published version will be spelled out when used in figure captions.
%   \begin{acronyms}
%   \acro{Acronym}
%   Definition here
%   \acro{EMOS}
%   Ensemble model output statistics
%   \acro{ECMWF}
%   Centre for Medium-Range Weather Forecasts
%   \end{acronyms}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Notation
%   \begin{notation}
%   \notation{$a+b$} Notation Definition here
%   \notation{$e=mc^2$}
%   Equation in German-born physicist Albert Einstein's theory of special
%  relativity that showed that the increased relativistic mass ($m$) of a
%  body comes from the energy of motion of the body—that is, its kinetic
%  energy ($E$)—divided by the speed of light squared ($c^2$).
%   \end{notation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% DATA SECTION and ACKNOWLEDGMENTS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Open Research Section}
This section MUST contain a statement that describes where the data supporting the conclusions can be obtained. Data cannot be listed as ''Available from authors'' or stored solely in supporting information. Citations to archived data should be included in your reference list. Wiley will publish it as a separate section on the paper’s page. Examples and complete information are here:
https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors

\acknowledgments
Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% REFERENCES and BIBLIOGRAPHY
%
% \bibliography{<name of your .bib file>} don't specify the file extension
% don't specify bibliographystyle
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliography{refs}

%Reference citation instructions and examples:
%
% Please use ONLY \cite and \citeA for reference citations.
% \cite for parenthetical references
% ...as shown in recent studies (Simpson et al., 2019)
% \citeA for in-text citations
% ...Simpson et al. (2019) have shown...
%
%
%...as shown by \citeA{jskilby}.
%...as shown by \citeA{lewin76}, \citeA{carson86}, \citeA{bartoldy02}, and \citeA{rinaldi03}.
%...has been shown \cite{jskilbye}.
%...has been shown \cite{lewin76,carson86,bartoldy02,rinaldi03}.
%... \cite <i.e.>[]{lewin76,carson86,bartoldy02,rinaldi03}.
%...has been shown by \cite <e.g.,>[and others]{lewin76}.
%
% apacite uses < > for prenotes and [ ] for postnotes
% DO NOT use other cite commands (e.g., \citet, \citep, \citeyear, \nocite, \citealp, etc.).
%

\end{document}

More Information and Advice:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  SECTION HEADS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Capitalize the first letter of each word (except for
% prepositions, conjunctions, and articles that are
% three or fewer letters).

% AGU follows standard outline style; therefore, there cannot be a section 1 without
% a section 2, or a section 2.3.1 without a section 2.3.2.
% Please make sure your section numbers are balanced.
% ---------------
% Level 1 head
%
% Use the \section{} command to identify level 1 heads;
% type the appropriate head wording between the curly
% brackets, as shown below.
%
%An example:
%\section{Level 1 Head: Introduction}
%
% ---------------
% Level 2 head
%
% Use the \subsection{} command to identify level 2 heads.
%An example:
%\subsection{Level 2 Head}
%
% ---------------
% Level 3 head
%
% Use the \subsubsection{} command to identify level 3 heads
%An example:
%\subsubsection{Level 3 Head}
%
%---------------
% Level 4 head
%
% Use the \subsubsubsection{} command to identify level 3 heads
% An example:
%\subsubsubsection{Level 4 Head} An example.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  IN-TEXT LISTS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Do not use bulleted lists; enumerated lists are okay.
% \begin{enumerate}
% \item
% \item
% \item
% \end{enumerate}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  EQUATIONS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Single-line equations are centered.
% Equation arrays will appear left-aligned.

Math coded inside display math mode \[ ...\]
 will not be numbered, e.g.,:
 \[ x^2=y^2 + z^2\]

 Math coded inside \begin{equation} and \end{equation} will
 be automatically numbered, e.g.,:
 \begin{equation}
 x^2=y^2 + z^2
 \end{equation}

% To create multiline equations, use the
% \begin{eqnarray} and \end{eqnarray} environment
% as demonstrated below.
\begin{eqnarray}
  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
        && + (y - y_{0}) \sin \Theta  \nonumber \\
  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
        && + (y - y_{0}) \cos \Theta.
\end{eqnarray}

%If you don't want an equation number, use the star form:
%\begin{eqnarray*}...\end{eqnarray*}

% Break each line at a sign of operation
% (+, -, etc.) if possible, with the sign of operation
% on the new line.

% Indent second and subsequent lines to align with
% the first character following the equal sign on the
% first line.

% Use an \hspace{} command to insert horizontal space
% into your equation if necessary. Place an appropriate
% unit of measure between the curly braces, e.g.
% \hspace{1in}; you may have to experiment to achieve
% the correct amount of space.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  EQUATION NUMBERING: COUNTER
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% You may change equation numbering by resetting
% the equation counter or by explicitly numbering
% an equation.

% To explicitly number an equation, type \eqnum{}
% (with the desired number between the brackets)
% after the \begin{equation} or \begin{eqnarray}
% command.  The \eqnum{} command will affect only
% the equation it appears with; LaTeX will number
% any equations appearing later in the manuscript
% according to the equation counter.
%

% If you have a multiline equation that needs only
% one equation number, use a \nonumber command in
% front of the double backslashes (\\) as shown in
% the multiline equation above.

% If you are using line numbers, remember to surround
% equations with \begin{linenomath*}...\end{linenomath*}

%  To add line numbers to lines in equations:
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

### 2023-03-07 ms/ms2.tex

--- 2023-03-04 ms/ms.tex
@@
 
-\authors{D. Kennedy \affil{1}, Katherine Dagon\affil{1}, David M. Lawrence\affil{1}}
-%R.A. Fisher
-%B.M. Sanderson
-%C.D. Koven
-%A.L.S. Swann
-%C.M. Zarakas
-%L.R. Hawkins
-%F. Hoffman
-%N. Collier
-%Y. Luo
-%X. Lu
-%S. Levis
-%S.C. Swenson
-%K.W. Oleson
-%E. Kluzek
-%J. Liu
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  AUTHORS AND AFFILIATIONS
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\affiliation{1}{Climate and Global Dynamics Laboratory, NCAR, Boulder, CO, USA.}
+% Authors are individuals who have significantly contributed to the
+% research and preparation of the article. Group authors are allowed, if
+% each author in the group is separately identified in an appendix.)
+
+% List authors by first name or initial followed by last name and
+% separated by commas. Use \affil{} to number affiliations, and
+% \thanks{} for author notes.
+% Additional author notes should be indicated with \thanks{} (for
+% example, for current addresses).
+
+% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
+% Author\affil{3,4}\thanks{Also funded by Monsanto.}}
+
+\authors{Daniel Kennedy\affil{1}, Katherine Dagon\affil{1}, David M. Lawrence\affil{1}}
+
+\affiliation{1}{Climate and Global Dynamics Laboratory, National Center for Atmospheric Research, Boulder, CO, USA}
 
@@
 \section{Introduction}
-\section{Experiment Description}
-\subsection{Model description}
-The Community Terrestrial System Model (CTSM) is developed by the CESM Land Model Working Group (LMWG) and maintained at the National Center for Atmospheric Research (NCAR).
-This experiment utilizes the Community Land Model configuration of CTSM, version 5.1 (CLM5.1).
-The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.
+\section{Experiment description}
 
-Relative to CLM5.0, version 5.1 includes minor bug fixes, several parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}.
-The PPE experiment required additional code modifications to identify and programmatically vary the full suite of model parameters.
-The exact model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
-We utilized the full biogeochemistry version of CLM in land-only mode, with the crop model turned off.
-The component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP}
+\subsection{Sparse grid}
 
-\subsection{Model spin-up}
-Model spin-up for the equilibration of carbon and nitrogen pools within biogeochemistry-enabled land models can consume up to 98\% of computational time \cite{sun2023}.
+Multivariate spatiotemporal clustering
 
-\cite{lu2020}
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/sparsegrid_gpp.pdf}
+\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation.
+We opted for 400 clusters (i.e.
+400 sparse gridcells) to balance computational cost against representativeness.
+For reference there are 22,648 and 5,666 land gridcells in standard 1$^{\circ}$ and 2$^{\circ}$ CLM simulations, respectively.}
+\label{fig:sparsegrid}
+\end{figure}
 
-\subsection{Sparsegrid}
+\begin{figure}[h]
+\centering
+\includegraphics[width=25pc]{../figs/sims.pdf}
+\caption{A standard 1-degree CLM-BGC simulation with 1500 years of spinup is prohibitively expensive ($\sim$400k pe-hours) for exhaustive parameter sensitivity testing.
+Utilizing the Matrix-CN spinup method and our sparsegrid configuration yields approximately 1300 simulations per million pe-hours.}
+\label{fig:timing}
+\end{figure}
 
-\cite{hoffman2013}
-
-\subsection{Experimental Design}
 \subsection{Forcing Scenarios}
@@
 
-\subsection{Analyses}
+\break
 \section{Results}
-\section{Discussion}
-\section{Conclusion}
-%Text here ===>>>
 
-%%
+\begin{figure}[h]
+\centering
+\includegraphics[width=30pc]{../figs/bar.pdf}
+\caption{206 parameters were identified and perturbed across the various domains of the land model.}
+\label{fig:params}
+\end{figure}
 
-%  Numbered lines in equations:
-%  To add line numbers to lines in equations,
-%  \begin{linenomath*}
-%  \begin{equation}
-%  \end{equation}
-%  \end{linenomath*}
+\break
 
-%% Enter Figures and Tables near as possible to where they are first mentioned:
-%
-% DO NOT USE \psfrag or \subfigure commands.
-%
-% Figure captions go below the figure.
-% Acronyms used in figure captions will be spelled out in the final, published version.
+\begin{figure}[h]
+\centering
+\includegraphics[width=25pc]{../figs/gpp_violin.pdf}
+\caption{Distributions of global annual GPP across the six forcing scenarios. See Section \ref{sect:exps} for scenario descriptions.}
+\label{fig:gpp1}
+\end{figure}
 
-% Table titles go above tables;  other caption information
-%  should be placed in last line of the table, using
-% \multicolumn2l{$^a$ This is a table note.}
-% NOTE that there is no difference between table caption and table heading in the final, published version
-%
-%----------------
-% EXAMPLE FIGURES
-%
-% \begin{figure}
-% \includegraphics{example.png}
-% \caption{caption}
-% \end{figure}
-%
-% Giving latex a width will help it to scale the figure properly. A simple trick is to use \textwidth. Try this if large figures run off the side of the page.
-% \begin{figure}
-% \noindent\includegraphics[width=\textwidth]{anothersample.png}
-%\caption{caption}
-%\label{pngfiguresample}
-%\end{figure}
-%
-%
-% If you get an error about an unknown bounding box, try specifying the width and height of the figure with the natwidth and natheight options.
-This is common when trying to add a PDF figure without pdflatex.
-% \begin{figure}
-% \noindent\includegraphics[natwidth=800px,natheight=600px]{samplefigure.pdf}
-%\caption{caption}
-%\label{pdffiguresample}
-%\end{figure}
-%
-%
-% PDFLatex does not seem to be able to process EPS figures. You may want to try the epstopdf package.
-%
+\break
 
-%
-% ---------------
-% EXAMPLE TABLE
-%
-% \begin{table}
-% \caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
-% \centering
-% \begin{tabular}{l c}
-% \hline
-%  Run  & Time (min)  \\
-% \hline
-%   $l1$  & 260   \\
-%   $l2$  & 300   \\
-%   $l3$  & 340   \\
-%   $h1$  & 270   \\
-%   $h2$  & 250   \\
-%   $h3$  & 380   \\
-%   $r1$  & 370   \\
-%   $r2$  & 390   \\
-% \hline
-% \multicolumn{2}{l}{$^{a}$Footnote text here.}
-% \end{tabular}
-% \end{table}
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/lai_biome.pdf}
+\caption{The eight most influential parameters on leaf area index within the CTL2010 ensemble, globally and within three biomes.}
+\label{fig:lai}
+\end{figure}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% SIDEWAYS FIGURES and TABLES
-% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
-%
-% \begin{sidewaysfigure}
-% \includegraphics[width=20pc]{figsamp}
-% \caption{caption here}
-% \label{newfig}
-% \end{sidewaysfigure}
-%
-%  \begin{sidewaystable}
-%  \caption{Caption here}
-% \label{tab:signif_gap_clos}
-%  \begin{tabular}{ccc}
-% one&two&three\\
-% four&five&six
-%  \end{tabular}
-%  \end{sidewaystable}
+\break
 
-%% If using numbered lines, please surround equations with \begin{linenomath*}...\end{linenomath*}
-%\begin{linenomath*}
-%\begin{equation}
-%y|{f} \sim g(m, \sigma),
-%\end{equation}
-%\end{linenomath*}
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/ET_response.pdf}
+\caption{(a) Global evapotranspiration in the warm forcing scenario vs.
+our control experiment.
+The default parameterization ET enhancement is +8.8\%, as shown with the orange line.
+The shading spans the CTL2010 default ET, plus or minus 1 W/m2.
+(b) The top 8 parameters governing global ET in the CTL2010 experiment.
+(c) The top 8 parameters governing ET response to the future climate anomaly forcing (AF2095).
+Parameters governing the response to temperature anomalies tend to differ from the parameters controlling present-day ET.}
+\label{fig:et}
+\end{figure}
 
-%%% End of body of article
+\break
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Optional Appendices go here
-%
-% The \appendix command resets counters and redefines section heads
-%
-% After typing \appendix
-%
-%\section{Here Is Appendix Title}
-% will show
-% A: Here Is Appendix Title
-%
-%\appendix
-%\section{Here is a sample appendix}
+\begin{figure}[h]
+\centering
+\includegraphics[width=25pc]{../figs/wue.pdf}
+\caption{Global photosynthesis vs.
+transpiration for the control and high CO$_2$ ensembles.
+Water use efficiency (in this case defined as GPP/T) contours are drawn for reference.
+The default parameterization increases GPP by 53\% in response to increasing CO$_2$ from 367 to 867 ppm, made possible by an even larger increase in WUE (+88\%).}
+\label{fig:wue}
+\end{figure}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Optional Glossary, Notation or Acronym section goes here:
-%
-% Glossary is only allowed in Reviews of Geophysics
-%  \begin{glossary}
-%  \term{Term}
-%   Term Definition here
-%  \term{Term}
-%   Term Definition here
-%  \term{Term}
-%   Term Definition here
-%  \end{glossary}
+\break
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Acronyms
-%% NOTE that acronyms in the final published version will be spelled out when used in figure captions.
-%   \begin{acronyms}
-%   \acro{Acronym}
-%   Definition here
-%   \acro{EMOS}
-%   Ensemble model output statistics
-%   \acro{ECMWF}
-%   Centre for Medium-Range Weather Forecasts
-%   \end{acronyms}
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/deltmap_panel.pdf}
+\caption{Screenshot of interactive diagnostic for exploring maps of parameter effects.
+In this case, the effect of medlynslope on runoff within the CTL2010 ensemble.
+Increasing medlynslope tends to increase transpiration and reduce runoff.}
+\label{fig:panel}
+\end{figure}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Notation
-%   \begin{notation}
-%   \notation{$a+b$} Notation Definition here
-%   \notation{$e=mc^2$}
-%   Equation in German-born physicist Albert Einstein's theory of special
-%  relativity that showed that the increased relativistic mass ($m$) of a
-%  body comes from the energy of motion of the body—that is, its kinetic
-%  energy ($E$)—divided by the speed of light squared ($c^2$).
-%   \end{notation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% DATA SECTION and ACKNOWLEDGMENTS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Open Research Section}
-This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
-Data cannot be listed as ''Available from authors'' or stored solely in supporting information.
-Citations to archived data should be included in your reference list.
-Wiley will publish it as a separate section on the paper’s page.
-Examples and complete information are here:
-https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
-
-\acknowledgments
-Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
+\break
 
@@
 
-\bibliography{refs}
+%\bibliography{ enter your bibtex bibliography filename here }
 

### 2023-03-13 ms/ms.tex
ms

@@
 \section{Experiment Description}
+\label{methods}
 \subsection{Model description}
+\label{sect:md}
 The Community Terrestrial System Model (CTSM) is developed by the CESM Land Model Working Group (LMWG) and maintained at the National Center for Atmospheric Research (NCAR).
@@
 Model spin-up for the equilibration of carbon and nitrogen pools within biogeochemistry-enabled land models can consume up to 98\% of computational time \cite{sun2023}.
+Depending on the evaluation criteria and model configuration, CLM5 requires between 800 and 2000 years (or more) to reach steady-state conditions \cite{lawrence2019}.
+Absent equilibrium, the drift towards steady state can obscure important model dynamics or features.
+For this reason, each member of the PPE requires independent spin-up.
 
-\cite{lu2020}
+To manage computational cost we leveraged the Matrix-CN spin-up mode recently implemented within CLM \cite{lu2020}.
+This new module leverages a linearized simplification of CLM's biogeochemistry to significantly reduce spin-up time.
+Our spin-up protocol featured 20 years in accelerated decomposition mode, followed by 80 years of Matrix-CN, followed by 40 years of `normal' mode, cycling over a ten-year forcing dataset.
+This was designed to achieve sufficiently equilibrated model states, while minimizing computational time.
+To reduce model cost, we did not require full equilibration of deep soil carbon (beyond 1 meter).
+Any inferences about deep soil carbon would therefore be subject to uncertainty due to spin-up concerns.
 
 \subsection{Sparsegrid}
+\label{sect:sg}
+One of the largest controls on model cost is resolution.
+Most CLM simulations utilize nominal 1$^\circ$ resolution, which requires over 20,000 land grid cells.
+In order to manage computational cost, parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}.
+CLM does not require standard rectilinear grids, which allows for innovative techniques to reduce model resolution
 
-\cite{hoffman2013}
+Multivariate spatio-temporal clustering (MVSC) has been utilized to extract patterns of climatological significance from climate model output \cite{hoffman2005} and applied to design a representativeness-based sampling network \cite{hoffman2013}.
+Instead of lowering resolution by increasing grid spacing, we used MVSC to strategically remove redundant grid cells, leaving only 400 grid cells that efficiently sample important model dynamics.
+Our goal was to quantify parameter effects with sufficient accuracy, while minimizing computational cost.
+
+We used k-means clustering to identify groups of grid cells with similar climatologies based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
+Instead of running every grid cell from a given cluster, we only run one representative grid cell, whichever is located nearest the cluster centroid in climate space.
+We refer to the set of representative grid cells as the `sparsegrid'.
+A paint-by-number approach is used to recompose mapped output and compute global means, where the output from the representative grid cell is substituted for all members of the cluster cohort.
+
+Clustering was based on a subset of 18 meaningful CLM variables (Table \ref{tab:sg}).
+The clustering algorithm received 12 observations of each variable per grid cell, namely the mean and interannual variability computed for six 30-year climatology windows (1865-1894, 1895-1924, ...
+, 1985-2014).
+Clusters were delineated to equalize the multi-dimensional variance across the user-specified number of groups, $k$.
+We tested 15 values of $k$, ranging from 10 to 800.
+Based on ILAMB2.5 benchmarking \cite{collier2018} against the full grid output, we opted for a 400-cluster sparsegrid, to balance computational cost against model fidelity (Supp Figure \ref{supp:ilamb}).
+Because our emphasis is on vegetated regions, we masked out Antarctica within the clustering algorithm, whereby we do not provide any output below 60$^\circ$S.
+
+\begin{table}[h]
+\caption{Clustering inputs}
+\centering
+\begin{tabular}{l c c c c}
+ \hline
+ Climate forcing variables & Ecosystem state variables &Ecosystem flux variables \\
+ \hline
+ 2m air temperature (TSA) & Leaf area index (TLAI) & Gross primary production (GPP) \\
+Atmospheric rain (RAIN) & Ecosystem carbon (TOTECOSYSC) &Heterotrophic respiration (HR) \\
+Atmospheric snow (SNOW) &  Ecosystem nitrogen (TOTECOSYSN) &Autotrophic respiration (AR) \\
+2m specific humidity (Q2M) & Soil ice (TOTSOILICE) &Net biome production (NBP) \\
+Solar radiation (FSDS) & Soil liquid water (TOTSOILLIQ) & Total liquid runoff (QRUNOFF) \\
+& Snow cover fraction (FSNO) & Sensible heat  (FSH) \\
+&&Latent heat (EFLX\_LH\_TOT ) \\
+ \hline
+ \end{tabular}
+ \label{tab:sg}
+ \end{table}
 
 \subsection{Experimental Design}
+493 simulations.
+Nine simulations failed.
+
 \subsection{Forcing Scenarios}
@@
 \section{Results}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=30pc]{../figs/sims.pdf}
+\caption{The approximate number of simulations afforded by 1 million core-hours for a range of CLM configurations.
+Configurations are labeled according to spin-up procedure and resolution, with `sg' signifying sparsegrid.
+The inset map shows the locations of the 400 sparse grid cells as the red dots.
+See Section \ref{methods} for spin-up and sparsegrid details.
+My speed calculations here are a bit janky, assume linear scaling of cost for increasing time or space (zqz).
+}
+\label{fig:sims}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/sparsegrid_gpp.pdf}
+\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation.
+We opted for 400 clusters to balance computational cost against representativeness.
+For reference there are 22648, 5666, and 1764 land gridcells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively.}
+\label{}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=30pc]{../figs/bar.pdf}
+\caption{206 parameters were identified and perturbed across the various domains of the land model.}
+\label{fig:params}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=35pc]{../figs/violins.pdf}
+\caption{Distributions of global annual gross primary production (GPP) and net biome production (NBP) across the six forcing scenarios.
+See Section \ref{sect:exps} for scenario descriptions.
+The distributions are concentrated around the default parameterization with long tails, indicating that a small proportion of parameters have large effects on these two variables.
+In the case of NBP, parameter effects appear larger than scenario effects.}
+\label{}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/lai_biome.pdf}
+\caption{The eight most influential parameters on leaf area index within the CTL2010 ensemble, globally and within three biomes.}
+\label{fig:lai}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/ET_response.pdf}
+\caption{(a) Global evapotranspiration in the warm forcing scenario vs.
+our control experiment.
+The default parameterization ET enhancement is +8.8\%, as shown with the orange line.
+The shading spans the CTL2010 default ET, plus or minus 1 W/m2.
+(b) The top 8 parameters governing global ET in the CTL2010 experiment.
+(c) The top 8 parameters governing ET response to the future climate anomaly forcing (AF2095).
+Parameters governing the response to temperature anomalies tend to differ from the parameters controlling present-day ET.}
+\label{fig:et}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/deltmap_panel.pdf}
+\caption{Screenshot of interactive diagnostic for exploring maps of parameter effects.
+In this case, the effect of medlynslope on runoff within the CTL2010 ensemble.
+Increasing medlynslope tends to increase transpiration and reduce runoff.}
+\label{fig:panel}
+\end{figure}
+
 \section{Discussion}
 \section{Conclusion}
-%Text here ===>>>
-
-%%
-
-%  Numbered lines in equations:
-%  To add line numbers to lines in equations,
-%  \begin{linenomath*}
-%  \begin{equation}
-%  \end{equation}
-%  \end{linenomath*}
-
-%% Enter Figures and Tables near as possible to where they are first mentioned:
-%
-% DO NOT USE \psfrag or \subfigure commands.
-%
-% Figure captions go below the figure.
-% Acronyms used in figure captions will be spelled out in the final, published version.
-
-% Table titles go above tables;  other caption information
-%  should be placed in last line of the table, using
-% \multicolumn2l{$^a$ This is a table note.}
-% NOTE that there is no difference between table caption and table heading in the final, published version
-%
-%----------------
-% EXAMPLE FIGURES
-%
-% \begin{figure}
-% \includegraphics{example.png}
-% \caption{caption}
-% \end{figure}
-%
-% Giving latex a width will help it to scale the figure properly. A simple trick is to use \textwidth. Try this if large figures run off the side of the page.
-% \begin{figure}
-% \noindent\includegraphics[width=\textwidth]{anothersample.png}
-%\caption{caption}
-%\label{pngfiguresample}
-%\end{figure}
-%
-%
-% If you get an error about an unknown bounding box, try specifying the width and height of the figure with the natwidth and natheight options.
-This is common when trying to add a PDF figure without pdflatex.
-% \begin{figure}
-% \noindent\includegraphics[natwidth=800px,natheight=600px]{samplefigure.pdf}
-%\caption{caption}
-%\label{pdffiguresample}
-%\end{figure}
-%
-%
-% PDFLatex does not seem to be able to process EPS figures. You may want to try the epstopdf package.
-%
-
-%
-% ---------------
-% EXAMPLE TABLE
-%
-% \begin{table}
-% \caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
-% \centering
-% \begin{tabular}{l c}
-% \hline
-%  Run  & Time (min)  \\
-% \hline
-%   $l1$  & 260   \\
-%   $l2$  & 300   \\
-%   $l3$  & 340   \\
-%   $h1$  & 270   \\
-%   $h2$  & 250   \\
-%   $h3$  & 380   \\
-%   $r1$  & 370   \\
-%   $r2$  & 390   \\
-% \hline
-% \multicolumn{2}{l}{$^{a}$Footnote text here.}
-% \end{tabular}
-% \end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% SIDEWAYS FIGURES and TABLES
-% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
-%
-% \begin{sidewaysfigure}
-% \includegraphics[width=20pc]{figsamp}
-% \caption{caption here}
-% \label{newfig}
-% \end{sidewaysfigure}
-%
-%  \begin{sidewaystable}
-%  \caption{Caption here}
-% \label{tab:signif_gap_clos}
-%  \begin{tabular}{ccc}
-% one&two&three\\
-% four&five&six
-%  \end{tabular}
-%  \end{sidewaystable}
-
-%% If using numbered lines, please surround equations with \begin{linenomath*}...\end{linenomath*}
-%\begin{linenomath*}
-%\begin{equation}
-%y|{f} \sim g(m, \sigma),
-%\end{equation}
-%\end{linenomath*}
-
-%%% End of body of article
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Optional Appendices go here
-%
-% The \appendix command resets counters and redefines section heads
-%
-% After typing \appendix
-%
-%\section{Here Is Appendix Title}
-% will show
-% A: Here Is Appendix Title
-%
-%\appendix
-%\section{Here is a sample appendix}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Optional Glossary, Notation or Acronym section goes here:
-%
-% Glossary is only allowed in Reviews of Geophysics
-%  \begin{glossary}
-%  \term{Term}
-%   Term Definition here
-%  \term{Term}
-%   Term Definition here
-%  \term{Term}
-%   Term Definition here
-%  \end{glossary}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Acronyms
-%% NOTE that acronyms in the final published version will be spelled out when used in figure captions.
-%   \begin{acronyms}
-%   \acro{Acronym}
-%   Definition here
-%   \acro{EMOS}
-%   Ensemble model output statistics
-%   \acro{ECMWF}
-%   Centre for Medium-Range Weather Forecasts
-%   \end{acronyms}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Notation
-%   \begin{notation}
-%   \notation{$a+b$} Notation Definition here
-%   \notation{$e=mc^2$}
-%   Equation in German-born physicist Albert Einstein's theory of special
-%  relativity that showed that the increased relativistic mass ($m$) of a
-%  body comes from the energy of motion of the body—that is, its kinetic
-%  energy ($E$)—divided by the speed of light squared ($c^2$).
-%   \end{notation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% DATA SECTION and ACKNOWLEDGMENTS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section*{Open Research Section}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% REFERENCES and BIBLIOGRAPHY
-%
-% \bibliography{<name of your .bib file>} don't specify the file extension
-% don't specify bibliographystyle
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \bibliography{refs}
 
-%Reference citation instructions and examples:
-%
-% Please use ONLY \cite and \citeA for reference citations.
-% \cite for parenthetical references
-% ...as shown in recent studies (Simpson et al., 2019)
-% \citeA for in-text citations
-% ...Simpson et al. (2019) have shown...
-%
-%
-%...as shown by \citeA{jskilby}.
-%...as shown by \citeA{lewin76}, \citeA{carson86}, \citeA{bartoldy02}, and \citeA{rinaldi03}.
-%...has been shown \cite{jskilbye}.
-%...has been shown \cite{lewin76,carson86,bartoldy02,rinaldi03}.
-%... \cite <i.e.>[]{lewin76,carson86,bartoldy02,rinaldi03}.
-%...has been shown by \cite <e.g.,>[and others]{lewin76}.
-%
-% apacite uses < > for prenotes and [ ] for postnotes
-% DO NOT use other cite commands (e.g., \citet, \citep, \citeyear, \nocite, \citealp, etc.).
-%
+\appendix
+\section{Supplementary Figures}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=42pc]{../figs/supp/ilamb.pdf}
+\caption{ILAMB 2.5 overall scores comparing remapped sparsegrid output to the full grid 2$^{\circ}$ model output.
+Column headings indicate the number of clusters, and whether annual means or annual means and standard deviations were used as input to the clustering algorithm.
+We should try to remove the gray bits (zqz).
+Full, interactive results are available at \url{www.ilamb.org/PPE/CLM/2021-02}.
+See main text Section \ref{sect:sg} for clustering details.}
+\label{supp:ilamb}
+\end{figure}
 
 \end{document}
-
-More Information and Advice:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  SECTION HEADS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Capitalize the first letter of each word (except for
-% prepositions, conjunctions, and articles that are
-% three or fewer letters).
-
-% AGU follows standard outline style; therefore, there cannot be a section 1 without
-% a section 2, or a section 2.3.1 without a section 2.3.2.
-% Please make sure your section numbers are balanced.
-% ---------------
-% Level 1 head
-%
-% Use the \section{} command to identify level 1 heads;
-% type the appropriate head wording between the curly
-% brackets, as shown below.
-%
-%An example:
-%\section{Level 1 Head: Introduction}
-%
-% ---------------
-% Level 2 head
-%
-% Use the \subsection{} command to identify level 2 heads.
-%An example:
-%\subsection{Level 2 Head}
-%
-% ---------------
-% Level 3 head
-%
-% Use the \subsubsection{} command to identify level 3 heads
-%An example:
-%\subsubsection{Level 3 Head}
-%
-%---------------
-% Level 4 head
-%
-% Use the \subsubsubsection{} command to identify level 3 heads
-% An example:
-%\subsubsubsection{Level 4 Head} An example.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  IN-TEXT LISTS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Do not use bulleted lists; enumerated lists are okay.
-% \begin{enumerate}
-% \item
-% \item
-% \item
-% \end{enumerate}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  EQUATIONS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Single-line equations are centered.
-% Equation arrays will appear left-aligned.
-
-Math coded inside display math mode \[ ...\]
- will not be numbered, e.g.,:
- \[ x^2=y^2 + z^2\]
-
- Math coded inside \begin{equation} and \end{equation} will
- be automatically numbered, e.g.,:
- \begin{equation}
- x^2=y^2 + z^2
- \end{equation}
-
-% To create multiline equations, use the
-% \begin{eqnarray} and \end{eqnarray} environment
-% as demonstrated below.
-\begin{eqnarray}
-  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
-        && + (y - y_{0}) \sin \Theta  \nonumber \\
-  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
-        && + (y - y_{0}) \cos \Theta.
-\end{eqnarray}
-
-%If you don't want an equation number, use the star form:
-%\begin{eqnarray*}...\end{eqnarray*}
-
-% Break each line at a sign of operation
-% (+, -, etc.) if possible, with the sign of operation
-% on the new line.
-
-% Indent second and subsequent lines to align with
-% the first character following the equal sign on the
-% first line.
-
-% Use an \hspace{} command to insert horizontal space
-% into your equation if necessary. Place an appropriate
-% unit of measure between the curly braces, e.g.
-% \hspace{1in}; you may have to experiment to achieve
-% the correct amount of space.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  EQUATION NUMBERING: COUNTER
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% You may change equation numbering by resetting
-% the equation counter or by explicitly numbering
-% an equation.
-
-% To explicitly number an equation, type \eqnum{}
-% (with the desired number between the brackets)
-% after the \begin{equation} or \begin{eqnarray}
-% command.  The \eqnum{} command will affect only
-% the equation it appears with; LaTeX will number
-% any equations appearing later in the manuscript
-% according to the equation counter.
-%
-
-% If you have a multiline equation that needs only
-% one equation number, use a \nonumber command in
-% front of the double backslashes (\\) as shown in
-% the multiline equation above.
-
-% If you are using line numbers, remember to surround
-% equations with \begin{linenomath*}...\end{linenomath*}
-
-%  To add line numbers to lines in equations:
-%  \begin{linenomath*}
-%  \begin{equation}
-%  \end{equation}
-%  \end{linenomath*}

### 2023-03-29 ms/ms.tex
ms

@@
 \section{Introduction}
+
+Water availability, land temperature extremes, fire risk, crop productivity, and the strength of the terrestrial carbon sink.
+Our certainty in model projections varies by domain and generally decreases as the time horizon is extended.
+Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations (Fried, Arora). 
+A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate (Bonan and Doney?).
+But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns we should expect predictions to become more skillful over time.
+
+While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations.
+Within the model development process, there is a larger emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
+At this point, complexity of land models has increased to the point
+
+Probabilistic forecasts
+
+Models have parameters.
+Some can be measured, but many cannot.
+
+Parametric uncertainty vs. structural uncertainty
+
 \section{Experiment Description}
@@
 
-Relative to CLM5.0, version 5.1 includes minor bug fixes, several parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}.
-The PPE experiment required additional code modifications to identify and programmatically vary the full suite of model parameters.
+Relative to CLM5.0, version 5.1 includes minor bug fixes, parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}.
+The PPE experiment required additional code modifications to programmatically vary the full suite of model parameters.
 The exact model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
-We utilized the full biogeochemistry version of CLM in land-only mode, with the crop model turned off.
+We utilized the biogeochemistry version of CLM in land-only mode, with the crop model turned off.
 The component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP}
@@
 To manage computational cost we leveraged the Matrix-CN spin-up mode recently implemented within CLM \cite{lu2020}.
-This new module leverages a linearized simplification of CLM's biogeochemistry to significantly reduce spin-up time.
+This new module utilizes a linearized simplification of CLM's biogeochemistry to significantly reduce spin-up time.
 Our spin-up protocol featured 20 years in accelerated decomposition mode, followed by 80 years of Matrix-CN, followed by 40 years of `normal' mode, cycling over a ten-year forcing dataset.
 This was designed to achieve sufficiently equilibrated model states, while minimizing computational time.
-To reduce model cost, we did not require full equilibration of deep soil carbon (beyond 1 meter).
-Any inferences about deep soil carbon would therefore be subject to uncertainty due to spin-up concerns.
+This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter).
+Certain inferences about deep soil carbon would therefore be subject to uncertainty due to spin-up concerns.
 
@@
 \label{sect:sg}
-One of the largest controls on model cost is resolution.
+Another control on model cost is resolution.
 Most CLM simulations utilize nominal 1$^\circ$ resolution, which requires over 20,000 land grid cells.
 In order to manage computational cost, parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}.
-CLM does not require standard rectilinear grids, which allows for innovative techniques to reduce model resolution
+We used a clustering algorithm to achieve an alternative low resolution configuration.
 
 Multivariate spatio-temporal clustering (MVSC) has been utilized to extract patterns of climatological significance from climate model output \cite{hoffman2005} and applied to design a representativeness-based sampling network \cite{hoffman2013}.
-Instead of lowering resolution by increasing grid spacing, we used MVSC to strategically remove redundant grid cells, leaving only 400 grid cells that efficiently sample important model dynamics.
-Our goal was to quantify parameter effects with sufficient accuracy, while minimizing computational cost.
+Instead of lowering resolution by coarsening a rectilinear grid, we used MVSC to strategically remove redundant grid cells, leaving only 400 grid cells that efficiently sample important model dynamics.
 
 We used k-means clustering to identify groups of grid cells with similar climatologies based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
-Instead of running every grid cell from a given cluster, we only run one representative grid cell, whichever is located nearest the cluster centroid in climate space.
-We refer to the set of representative grid cells as the `sparsegrid'.
+We selected one representative grid cell from each cluster, whichever is located nearest the cluster centroid in climate space.
+The set of representative grid cells comprise a `sparsegrid', which are used in lieu of a coarse grid.
 A paint-by-number approach is used to recompose mapped output and compute global means, where the output from the representative grid cell is substituted for all members of the cluster cohort.
@@
 \subsection{Experimental Design}
-493 simulations.
-Nine simulations failed.
+In the preparation for this experiment, 206 CLM-BGC parameters were identified.
+We decided to vary each parameter once at a time, to a minimum and maximum values.
+To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter.
+In some cases literature values were available, but in the vast majority expert judgment was used.
+The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.
 
-\subsection{Forcing Scenarios}
+Each simulation lasted 150 years, with the first 140 for spin-up, followed by a 10-year period for analysis purposes.
+We opted for six different forcing scenarios to understand the intersection of parameter effects and climate change (Table \ref{tab:exps}).
+The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and is the default forcing data for CLM5 \cite{lawrence2019}.
+We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
+Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the mean climate change signal.
+We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.
+
 \label{sect:exps}
@@
 \subsection{Analyses}
+
+Biome analysis
+
 \section{Results}

### 2023-04-10 ms/ms2.tex
ms

@@
 \documentclass[draft]{agujournal2019}
-\usepackage{url} %this package should fix any errors with URLs in refs.
+\usepackage{url}
 \usepackage{lineno}
-\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
+\usepackage[inline]{trackchanges} 
 \usepackage{soul}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  AUTHORS AND AFFILIATIONS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\authors{D. Kennedy \affil{1}, Katherine Dagon\affil{1}, David M. Lawrence\affil{1}}
 
-% Authors are individuals who have significantly contributed to the
-% research and preparation of the article. Group authors are allowed, if
-% each author in the group is separately identified in an appendix.)
-
-% List authors by first name or initial followed by last name and
-% separated by commas. Use \affil{} to number affiliations, and
-% \thanks{} for author notes.
-% Additional author notes should be indicated with \thanks{} (for
-% example, for current addresses).
-
-% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
-% Author\affil{3,4}\thanks{Also funded by Monsanto.}}
-
-\authors{Daniel Kennedy\affil{1}, Katherine Dagon\affil{1}, David M. Lawrence\affil{1}}
-
-\affiliation{1}{Climate and Global Dynamics Laboratory, National Center for Atmospheric Research, Boulder, CO, USA}
+\affiliation{1}{Climate and Global Dynamics Laboratory, NCAR, Boulder, CO, USA.}
 
@@
 \begin{keypoints}
-\item enter point 1 here
-\item enter point 2 here
-\item enter point 3 here
+\item Parameter perturbation experiment
 \end{keypoints}
@@
 \section{Introduction}
-\section{Experiment description}
 
-\subsection{Sparse grid}
+Water availability, land temperature extremes, fire risk, crop productivity, and the strength of the terrestrial carbon sink.
+Our certainty in model projections varies by domain and generally decreases as the time horizon is extended.
+Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations (Fried, Arora). 
+A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate (Bonan and Doney?).
+But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns we should expect predictions to become more skillful over time.
 
-Multivariate spatiotemporal clustering
+While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations.
+Within the model development process, there is a larger emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
+At this point, complexity of land models has increased to the point
 
-\begin{figure}[h]
+Probabilistic forecasts
+
+Models have parameters.
+Some can be measured, but many cannot.
+
+Parametric uncertainty vs. structural uncertainty
+
+\section{Experiment Description}
+\label{methods}
+\subsection{Model description}
+\label{sect:md}
+The Community Terrestrial System Model (CTSM) is developed by the CESM Land Model Working Group (LMWG) and maintained at the National Center for Atmospheric Research (NCAR).
+This experiment utilizes the Community Land Model configuration of CTSM, version 5.1 (CLM5.1).
+The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.
+
+Relative to CLM5.0, version 5.1 includes minor bug fixes, parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}.
+The PPE experiment required additional code modifications to programmatically vary the full suite of model parameters.
+The exact model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
+We utilized the biogeochemistry version of CLM in land-only mode, with the crop model turned off.
+The component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP}
+
+\subsection{Model spin-up}
+Model spin-up for the equilibration of carbon and nitrogen pools within biogeochemistry-enabled land models can consume up to 98\% of computational time \cite{sun2023}.
+Depending on the evaluation criteria and model configuration, CLM5 requires between 800 and 2000 years (or more) to reach steady-state conditions \cite{lawrence2019}.
+Absent equilibrium, the drift towards steady state can obscure important model dynamics or features.
+For this reason, each member of the PPE requires independent spin-up.
+
+To manage computational cost we leveraged the Matrix-CN spin-up mode recently implemented within CLM \cite{lu2020}.
+This new module utilizes a linearized simplification of CLM's biogeochemistry to significantly reduce spin-up time.
+Our spin-up protocol featured 20 years in accelerated decomposition mode, followed by 80 years of Matrix-CN, followed by 40 years of `normal' mode, cycling over a ten-year forcing dataset.
+This was designed to achieve sufficiently equilibrated model states, while minimizing computational time.
+This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter).
+Certain inferences about deep soil carbon would therefore be subject to uncertainty due to spin-up concerns.
+
+\subsection{Sparsegrid}
+\label{sect:sg}
+Another control on model cost is resolution.
+Most CLM simulations utilize nominal 1$^\circ$ resolution, which requires over 20,000 land grid cells.
+In order to manage computational cost, parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}.
+We used a clustering algorithm to achieve an alternative low resolution configuration.
+
+Multivariate spatio-temporal clustering (MVSC) has been utilized to extract patterns of climatological significance from climate model output \cite{hoffman2005} and applied to design a representativeness-based sampling network \cite{hoffman2013}.
+Instead of lowering resolution by coarsening a rectilinear grid, we used MVSC to strategically remove redundant grid cells, leaving only 400 grid cells that efficiently sample important model dynamics.
+
+We used k-means clustering to identify groups of grid cells with similar climatologies based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
+We selected one representative grid cell from each cluster, whichever is located nearest the cluster centroid in climate space.
+The set of representative grid cells comprise a `sparsegrid', which are used in lieu of a coarse grid.
+A paint-by-number approach is used to recompose mapped output and compute global means, where the output from the representative grid cell is substituted for all members of the cluster cohort.
+
+Clustering was based on a subset of 18 meaningful CLM variables (Table \ref{tab:sg}).
+The clustering algorithm received 12 observations of each variable per grid cell, namely the mean and interannual variability computed for six 30-year climatology windows (1865-1894, 1895-1924, ...
+, 1985-2014).
+Clusters were delineated to equalize the multi-dimensional variance across the user-specified number of groups, $k$.
+We tested 15 values of $k$, ranging from 10 to 800.
+Based on ILAMB2.5 benchmarking \cite{collier2018} against the full grid output, we opted for a 400-cluster sparsegrid, to balance computational cost against model fidelity (Supp Figure \ref{supp:ilamb}).
+Because our emphasis is on vegetated regions, we masked out Antarctica within the clustering algorithm, whereby we do not provide any output below 60$^\circ$S.
+
+\begin{table}[h]
+\caption{Clustering inputs}
 \centering
-\includegraphics[width=40pc]{../figs/sparsegrid_gpp.pdf}
-\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation.
-We opted for 400 clusters (i.e.
-400 sparse gridcells) to balance computational cost against representativeness.
-For reference there are 22,648 and 5,666 land gridcells in standard 1$^{\circ}$ and 2$^{\circ}$ CLM simulations, respectively.}
-\label{fig:sparsegrid}
-\end{figure}
+\begin{tabular}{l c c c c}
+ \hline
+ Climate forcing variables & Ecosystem state variables &Ecosystem flux variables \\
+ \hline
+ 2m air temperature (TSA) & Leaf area index (TLAI) & Gross primary production (GPP) \\
+Atmospheric rain (RAIN) & Ecosystem carbon (TOTECOSYSC) &Heterotrophic respiration (HR) \\
+Atmospheric snow (SNOW) &  Ecosystem nitrogen (TOTECOSYSN) &Autotrophic respiration (AR) \\
+2m specific humidity (Q2M) & Soil ice (TOTSOILICE) &Net biome production (NBP) \\
+Solar radiation (FSDS) & Soil liquid water (TOTSOILLIQ) & Total liquid runoff (QRUNOFF) \\
+& Snow cover fraction (FSNO) & Sensible heat  (FSH) \\
+&&Latent heat (EFLX\_LH\_TOT ) \\
+ \hline
+ \end{tabular}
+ \label{tab:sg}
+ \end{table}
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=25pc]{../figs/sims.pdf}
-\caption{A standard 1-degree CLM-BGC simulation with 1500 years of spinup is prohibitively expensive ($\sim$400k pe-hours) for exhaustive parameter sensitivity testing.
-Utilizing the Matrix-CN spinup method and our sparsegrid configuration yields approximately 1300 simulations per million pe-hours.}
-\label{fig:timing}
-\end{figure}
+\subsection{Experimental Design}
+In the preparation for this experiment, 206 CLM-BGC parameters were identified.
+We decided to vary each parameter once at a time, to a minimum and maximum values.
+To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter.
+In some cases literature values were available, but in the vast majority expert judgment was used.
+The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.
 
-\subsection{Forcing Scenarios}
+Each simulation lasted 150 years, with the first 140 for spin-up, followed by a 10-year period for analysis purposes.
+We opted for six different forcing scenarios to understand the intersection of parameter effects and climate change (Table \ref{tab:exps}).
+The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and is the default forcing data for CLM5 \cite{lawrence2019}.
+We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
+Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the mean climate change signal.
+We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.
+
 \label{sect:exps}
@@
 
-\break
+\subsection{Analyses}
+
+Biome analysis
+
 \section{Results}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=30pc]{../figs/sims.pdf}
+\caption{The approximate number of simulations afforded by 1 million core-hours for a range of CLM configurations.
+Configurations are labeled according to spin-up procedure and resolution, with `sg' signifying sparsegrid.
+The inset map shows the locations of the 400 sparse grid cells as the red dots.
+See Section \ref{methods} for spin-up and sparsegrid details.
+My speed calculations here are a bit janky, assume linear scaling of cost for increasing time or space (zqz).
+}
+\label{fig:sims}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/sparsegrid_gpp.pdf}
+\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation.
+We opted for 400 clusters to balance computational cost against representativeness.
+For reference there are 22648, 5666, and 1764 land gridcells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively.}
+\label{}
+\end{figure}
 
@@
 
-\break
-
 \begin{figure}[h]
 \centering
-\includegraphics[width=25pc]{../figs/gpp_violin.pdf}
-\caption{Distributions of global annual GPP across the six forcing scenarios. See Section \ref{sect:exps} for scenario descriptions.}
-\label{fig:gpp1}
+\includegraphics[width=35pc]{../figs/violins.pdf}
+\caption{Distributions of global annual gross primary production (GPP) and net biome production (NBP) across the six forcing scenarios.
+See Section \ref{sect:exps} for scenario descriptions.
+The distributions are concentrated around the default parameterization with long tails, indicating that a small proportion of parameters have large effects on these two variables.
+In the case of NBP, parameter effects appear larger than scenario effects.}
+\label{}
 \end{figure}
-
-\break
 
@@
 \end{figure}
-
-\break
 
@@
 
-\break
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=25pc]{../figs/wue.pdf}
-\caption{Global photosynthesis vs.
-transpiration for the control and high CO$_2$ ensembles.
-Water use efficiency (in this case defined as GPP/T) contours are drawn for reference.
-The default parameterization increases GPP by 53\% in response to increasing CO$_2$ from 367 to 867 ppm, made possible by an even larger increase in WUE (+88\%).}
-\label{fig:wue}
-\end{figure}
-
-\break
-
 \begin{figure}[h]
@@
 
-\break
+\section{Discussion}
+\section{Conclusion}
+\section*{Open Research Section}
+This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
+Data cannot be listed as ''Available from authors'' or stored solely in supporting information.
+Citations to archived data should be included in your reference list.
+Wiley will publish it as a separate section on the paper’s page.
+Examples and complete information are here:
+https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% REFERENCES and BIBLIOGRAPHY
-%
-% \bibliography{<name of your .bib file>} don't specify the file extension
-% don't specify bibliographystyle
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\acknowledgments
+Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
 
-%\bibliography{ enter your bibtex bibliography filename here }
+\bibliography{refs}
 
-%Reference citation instructions and examples:
-%
-% Please use ONLY \cite and \citeA for reference citations.
-% \cite for parenthetical references
-% ...as shown in recent studies (Simpson et al., 2019)
-% \citeA for in-text citations
-% ...Simpson et al. (2019) have shown...
-%
-%
-%...as shown by \citeA{jskilby}.
-%...as shown by \citeA{lewin76}, \citeA{carson86}, \citeA{bartoldy02}, and \citeA{rinaldi03}.
-%...has been shown \cite{jskilbye}.
-%...has been shown \cite{lewin76,carson86,bartoldy02,rinaldi03}.
-%... \cite <i.e.>[]{lewin76,carson86,bartoldy02,rinaldi03}.
-%...has been shown by \cite <e.g.,>[and others]{lewin76}.
-%
-% apacite uses < > for prenotes and [ ] for postnotes
-% DO NOT use other cite commands (e.g., \citet, \citep, \citeyear, \nocite, \citealp, etc.).
-%
+\appendix
+\section{Supplementary Figures}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=42pc]{../figs/supp/ilamb.pdf}
+\caption{ILAMB 2.5 overall scores comparing remapped sparsegrid output to the full grid 2$^{\circ}$ model output.
+Column headings indicate the number of clusters, and whether annual means or annual means and standard deviations were used as input to the clustering algorithm.
+We should try to remove the gray bits (zqz).
+Full, interactive results are available at \url{www.ilamb.org/PPE/CLM/2021-02}.
+See main text Section \ref{sect:sg} for clustering details.}
+\label{supp:ilamb}
+\end{figure}
 
 \end{document}
-
-More Information and Advice:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  SECTION HEADS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Capitalize the first letter of each word (except for
-% prepositions, conjunctions, and articles that are
-% three or fewer letters).
-
-% AGU follows standard outline style; therefore, there cannot be a section 1 without
-% a section 2, or a section 2.3.1 without a section 2.3.2.
-% Please make sure your section numbers are balanced.
-% ---------------
-% Level 1 head
-%
-% Use the \section{} command to identify level 1 heads;
-% type the appropriate head wording between the curly
-% brackets, as shown below.
-%
-%An example:
-%\section{Level 1 Head: Introduction}
-%
-% ---------------
-% Level 2 head
-%
-% Use the \subsection{} command to identify level 2 heads.
-%An example:
-%\subsection{Level 2 Head}
-%
-% ---------------
-% Level 3 head
-%
-% Use the \subsubsection{} command to identify level 3 heads
-%An example:
-%\subsubsection{Level 3 Head}
-%
-%---------------
-% Level 4 head
-%
-% Use the \subsubsubsection{} command to identify level 3 heads
-% An example:
-%\subsubsubsection{Level 4 Head} An example.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  IN-TEXT LISTS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Do not use bulleted lists; enumerated lists are okay.
-% \begin{enumerate}
-% \item
-% \item
-% \item
-% \end{enumerate}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  EQUATIONS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Single-line equations are centered.
-% Equation arrays will appear left-aligned.
-
-Math coded inside display math mode \[ ...\]
- will not be numbered, e.g.,:
- \[ x^2=y^2 + z^2\]
-
- Math coded inside \begin{equation} and \end{equation} will
- be automatically numbered, e.g.,:
- \begin{equation}
- x^2=y^2 + z^2
- \end{equation}
-
-% To create multiline equations, use the
-% \begin{eqnarray} and \end{eqnarray} environment
-% as demonstrated below.
-\begin{eqnarray}
-  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
-        && + (y - y_{0}) \sin \Theta  \nonumber \\
-  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
-        && + (y - y_{0}) \cos \Theta.
-\end{eqnarray}
-
-%If you don't want an equation number, use the star form:
-%\begin{eqnarray*}...\end{eqnarray*}
-
-% Break each line at a sign of operation
-% (+, -, etc.) if possible, with the sign of operation
-% on the new line.
-
-% Indent second and subsequent lines to align with
-% the first character following the equal sign on the
-% first line.
-
-% Use an \hspace{} command to insert horizontal space
-% into your equation if necessary. Place an appropriate
-% unit of measure between the curly braces, e.g.
-% \hspace{1in}; you may have to experiment to achieve
-% the correct amount of space.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  EQUATION NUMBERING: COUNTER
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% You may change equation numbering by resetting
-% the equation counter or by explicitly numbering
-% an equation.
-
-% To explicitly number an equation, type \eqnum{}
-% (with the desired number between the brackets)
-% after the \begin{equation} or \begin{eqnarray}
-% command.  The \eqnum{} command will affect only
-% the equation it appears with; LaTeX will number
-% any equations appearing later in the manuscript
-% according to the equation counter.
-%
-
-% If you have a multiline equation that needs only
-% one equation number, use a \nonumber command in
-% front of the double backslashes (\\) as shown in
-% the multiline equation above.
-
-% If you are using line numbers, remember to surround
-% equations with \begin{linenomath*}...\end{linenomath*}
-
-%  To add line numbers to lines in equations:
-%  \begin{linenomath*}
-%  \begin{equation}
-%  \end{equation}
-%  \end{linenomath*}

### 2023-04-26 ms/ms.tex
update
restore ms

@@
 
-Water availability, land temperature extremes, fire risk, crop productivity, and the strength of the terrestrial carbon sink.
-Our certainty in model projections varies by domain and generally decreases as the time horizon is extended.
-Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations (Fried, Arora). 
-A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate (Bonan and Doney?).
-But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns we should expect predictions to become more skillful over time.
+Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many variables represented within the land component of Earth System Models.
+Understanding how such states respond to carbon dioxide concentration, and how we expect carbon dioxide concentration to evolve over time, is at the center of climate change research.  
+Our certainty in climate model projections varies by domain and generally decreases with extended time horizons.
+Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. 
+A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{bonan2018}.
+But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, we would expect predictions to become more skillful over time.
 
 While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations.
-Within the model development process, there is a larger emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
-At this point, complexity of land models has increased to the point
+Within the day-to-day model development process, there is a greater emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
+However, the complexity of land models has increased to the point where understanding the full span of model function has become challenging \cite{fisher2020}, and manual model calibration is becoming intractable \cite{dagon2020}.
+Maintaining, improving, and interpreting complex land models benefits from thoughtful investment in software to automate and routinize important components of the development process, e.g.
+\citeA{collier2018}.
 
-Probabilistic forecasts
+One such fundamental model development step is parameter sensitivity testing and/or parameter optimization.
+Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017}.
+When applied systematically, the result is a parameter perturbation ensemble (PPE), alternatively termed a perturbed physics ensemble.
+PPEs have played an important role in assessing projection uncertainty in climate models.
+They also form a basis for automated model calibration.
+Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.
+In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
+In this paper we present that dataset, how it was produced, and its potential applications.
 
-Models have parameters.
-Some can be measured, but many cannot.
-
-Parametric uncertainty vs. structural uncertainty
+We are working with the Community Land Model (CLM), the land component of the Community Earth System Model, which is developed and maintained at the National Center for Atmospheric Research (NCAR).
+This work follows up on \citeA{dagon2020}, which began the process of systematically varying CLM parameters in service of automated model calibration.
+We extend that work by sampling a more exhaustive set of parameters across a broader range of land model processes.
+A set of 206 parameters were varied independently (one-at-a-time) across a range of climate scenarios.
+This dataset has already demonstrated great utility for diagnosing parameter effects, while the software and modeling infrastructure has greatly our expanded our capability to generate insights about CLM dynamics and uncertainties.
 
@@
 \label{sect:md}
-The Community Terrestrial System Model (CTSM) is developed by the CESM Land Model Working Group (LMWG) and maintained at the National Center for Atmospheric Research (NCAR).
+The Community Terrestrial System Model (CTSM) is developed by the CESM Land Model Working Group and maintained at the National Center for Atmospheric Research (NCAR).
 This experiment utilizes the Community Land Model configuration of CTSM, version 5.1 (CLM5.1).
@@
 
-We used k-means clustering to identify groups of grid cells with similar climatologies based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
-We selected one representative grid cell from each cluster, whichever is located nearest the cluster centroid in climate space.
-The set of representative grid cells comprise a `sparsegrid', which are used in lieu of a coarse grid.
+We used k-means clustering to identify groups of grid cells with similar climates based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
+We selected one representative grid cell from each cluster to stand in for the entire cluster.
+The representative grid cell is whichever is located nearest the cluster centroid in climate space.
+The set of representative grid cells comprise a `sparsegrid', which are used in lieu of a `coarse' grid.
 A paint-by-number approach is used to recompose mapped output and compute global means, where the output from the representative grid cell is substituted for all members of the cluster cohort.
@@
 Clustering was based on a subset of 18 meaningful CLM variables (Table \ref{tab:sg}).
-The clustering algorithm received 12 observations of each variable per grid cell, namely the mean and interannual variability computed for six 30-year climatology windows (1865-1894, 1895-1924, ...
+The clustering algorithm analyzed 12 observations of each variable per grid cell, namely the mean and interannual variability computed for six 30-year climatology windows (1865-1894, 1895-1924, ...
 , 1985-2014).
@@
 In the preparation for this experiment, 206 CLM-BGC parameters were identified.
-We decided to vary each parameter once at a time, to a minimum and maximum values.
+We decided to vary each parameter independently, to a low and high value.
 To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter.
-In some cases literature values were available, but in the vast majority expert judgment was used.
+In some cases literature values were directly utilized, but in the majority of cases, expert judgment was used.
 The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.
 
-Each simulation lasted 150 years, with the first 140 for spin-up, followed by a 10-year period for analysis purposes.
+Each simulation comprised 150 years, with the first 140 for spin-up, followed by a 10-year period for analysis.
 We opted for six different forcing scenarios to understand the intersection of parameter effects and climate change (Table \ref{tab:exps}).
@@
 We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
-Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the mean climate change signal.
+Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to a derived climate change signal.
 We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.
@@
 \section{Results}
+A major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
+Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times faster than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
+There are 22648, 5666, and 1764 land gridcells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
+Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibration criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.
 
@@
 Configurations are labeled according to spin-up procedure and resolution, with `sg' signifying sparsegrid.
-The inset map shows the locations of the 400 sparse grid cells as the red dots.
+The inset map shows the locations of the 400 sparse grid cells.
 See Section \ref{methods} for spin-up and sparsegrid details.
-My speed calculations here are a bit janky, assume linear scaling of cost for increasing time or space (zqz).
-}
+(My speed calculations here are a bit janky, assume linear scaling of cost for increasing time or space zqz) }
 \label{fig:sims}
@@
 
+Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
+stronger the relationship between output from the full grid and the sparsegrid.
+Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sggpp}).
+We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
 \begin{figure}[h]
@@
 \caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation.
-We opted for 400 clusters to balance computational cost against representativeness.
-For reference there are 22648, 5666, and 1764 land gridcells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively.}
-\label{}
+We opted for 400 clusters to balance computational cost against representativeness.}
+\label{fig:sggpp}
 \end{figure}
+
+Parameters were varied across a wide variety of CLM subcomponents (Figure \ref{fig:params}).
+A total of 206 parameters were targeted for sensitivity testing.
+Many parameters were hard-coded, but have now been extracted to the CLM parameter file for easier manipulation.
+In some cases new parameters were introduced to allow perturbation of important processes.
+For example a sand perturbation factor ($sand\_pf$) was introduced in order to perturb the percent sand in the soil column without creating a new surface dataset.
+A spreadsheet detailing all of the parameters, and their ranges, is provided in the supplementary material and available online (zqz link).
 
@@
 See Section \ref{sect:exps} for scenario descriptions.
-The distributions are concentrated around the default parameterization with long tails, indicating that a small proportion of parameters have large effects on these two variables.
-In the case of NBP, parameter effects appear larger than scenario effects.}
+The distributions are concentrated around the default parameterization with long tails, indicating that a small proportion of parameters have large effects.
+Especially in the case of NBP, parameter effects can be larger than scenario effects.}
 \label{}
@@
 \section{Discussion}
+
+Major discussion points:
+\begin{itemize}
+\item need for speed
+\item parameter effects can be quite large
+\item parameter priors are difficult to establish
+\item mean and climate response variables may vary
+\item community resource
+\item invest in infrastructure
+\end{itemize}
+
 \section{Conclusion}

### 2023-05-10 ms/ms.tex
hi dan

@@
 \usepackage{soul}
+\usepackage{pdflscape}
 \linenumbers
@@
 
-\title{The Community Land Model, version 5: \\ Parameter Perturbation Experiment}
+\title{The Community Land Model, version 5: \\ One-at-a-time Parameter Perturbation Ensemble}
 
@@
 Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many variables represented within the land component of Earth System Models.
-Understanding how such states respond to carbon dioxide concentration, and how we expect carbon dioxide concentration to evolve over time, is at the center of climate change research.  
+Understanding how these variables respond to carbon dioxide concentration, and how we expect carbon dioxide concentration to evolve over time, is at the center of climate change research.  
 Our certainty in climate model projections varies by domain and generally decreases with extended time horizons.
@@
 We extend that work by sampling a more exhaustive set of parameters across a broader range of land model processes.
-A set of 206 parameters were varied independently (one-at-a-time) across a range of climate scenarios.
+A set of approximately 200 parameters were varied across a range of climate scenarios.
 This dataset has already demonstrated great utility for diagnosing parameter effects, while the software and modeling infrastructure has greatly our expanded our capability to generate insights about CLM dynamics and uncertainties.
@@
 The PPE experiment required additional code modifications to programmatically vary the full suite of model parameters.
-The exact model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
+The model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
 We utilized the biogeochemistry version of CLM in land-only mode, with the crop model turned off.
@@
 \subsection{Experimental Design}
-In the preparation for this experiment, 206 CLM-BGC parameters were identified.
+In the preparation for this experiment, 188 CLM-BGC parameters were identified.
 We decided to vary each parameter independently, to a low and high value.
@@
 
-Each simulation comprised 150 years, with the first 140 for spin-up, followed by a 10-year period for analysis.
+Each simulation ran for 150 years, with the first 140 for spin-up, followed by a 10-year period for analysis.
 We opted for six different forcing scenarios to understand the intersection of parameter effects and climate change (Table \ref{tab:exps}).
@@
 We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
-Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to a derived climate change signal.
+Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the mean climate change signal.
 We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.
@@
 
+\subsection{Parameters} 
+\begin{landscape}
+
+ \begin{table}[h]
+ \caption{Some key parameters \\
+ Right now I'm sorting them by domain, should I sort them alphabetically by parameter name instead?}
+ \centering
+ \begin{tabular}{l l c}
+ \hline
+  Parameter  & Description & Model Domain \\
+ \hline
+d\_max & Dry surface layer parameter & Sensible, latent heat and momentum fluxes \\
+frac\_sat\_soil\_dsl\_init & Fraction of saturated soil at which DSL initiates & Sensible, latent heat and momentum fluxes \\
+fff & Decay factor for fractional saturated area & Hydrology \\
+liq\_canopy\_storage\_scalar & Canopy-storage-of-liquid-water parameter & Hydrology \\
+maximum\_leaf\_wetted\_fraction & Maximum leaf wetted fraction & Hydrology \\
+medlynintercept & Medlyn intercept of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
+medlynslope & Medlyn slope of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
+tpu25ratio & Ratio of tpu25top to vcmax25top & Stomatal resistance and photosynthesis \\
+jmaxb0 & Baseline proportion of nitrogen allocated for electron transport & Photosynthetic capacity (LUNA) \\
+jmaxb1 & Response of electron transport rate to light & Photosynthetic capacity (LUNA) \\
+slatop & Specific leaf area at top of canopy & Photosynthetic capacity (LUNA) \\
+wc2wjb0 & Baseline ratio of wc:wj & Photosynthetic capacity (LUNA) \\
+kmax & Plant segment max conductance & Plant hydraulics \\
+krmax & Root segment max conductance & Plant hydraulics \\
+psi50 & Water potential at 50\% loss of conductance & Plant hydraulics \\
+nstem & Stem number & Biomass heat storage \\
+lmr\_intercept\_atkin & Intercept in the calculation of leaf maintenance respiration& Plant respiration \\
+froot\_leaf & Allocation parameter: new fine root C per new leaf C & Carbon and nitrogen allocation \\
+leafcn & Leaf C:N & Carbon and nitrogen allocation \\
+leaf\_long & Leaf longevity & Vegetation phenology and turnover \\
+cpha & Activation energy for cp & Acclimation parameters \\
+jmaxhd & Deactivation energy for jmax & Acclimation parameters \\
+kcha & Activation energy for kc & Acclimation parameters \\
+lmrha & Activation energy for lmr & Acclimation parameters \\
+lmrhd & Deactivation energy for lmr & Acclimation parameters \\
+tpuha & Activation energy for tpu & Acclimation parameters \\
+tpuse\_sf & Scale factor for tpu entropy term & Acclimation parameters \\
+vcmaxha & Activation energy for vcmax & Acclimation parameters \\
+vcmaxhd & Deactivation energy for vcmax & Acclimation parameters \\
+ \hline
+ \end{tabular}
+ \end{table}
+\end{landscape}
+
 \subsection{Analyses}
@@
 \section{Results}
-A major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
-Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times faster than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
-There are 22648, 5666, and 1764 land gridcells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
-Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibration criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.
 
@@
 \caption{The approximate number of simulations afforded by 1 million core-hours for a range of CLM configurations.
-Configurations are labeled according to spin-up procedure and resolution, with `sg' signifying sparsegrid.
+Configurations are labeled according to spin-up procedure (CN-matrix or the standard 1500-year spinup) and horizontal resolution (`sg' signifies sparsegrid).
 The inset map shows the locations of the 400 sparse grid cells.
-See Section \ref{methods} for spin-up and sparsegrid details.
-(My speed calculations here are a bit janky, assume linear scaling of cost for increasing time or space zqz) }
+See Section \ref{methods} for spin-up and sparsegrid details.}
 \label{fig:sims}
 \end{figure}
+
+A major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
+Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times faster than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
+There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
+Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibration criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.
 
@@
 \centering
-\includegraphics[width=40pc]{../figs/sparsegrid_gpp.pdf}
+\includegraphics[width=35pc]{../figs/sparsegrid_gpp.pdf}
 \caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation.
@@
 Parameters were varied across a wide variety of CLM subcomponents (Figure \ref{fig:params}).
-A total of 206 parameters were targeted for sensitivity testing.
+A total of 188 parameters were targeted for sensitivity testing.
 Many parameters were hard-coded, but have now been extracted to the CLM parameter file for easier manipulation.
@@
 
+Most of the parameter perturbations had a relatively small impact on any given output variable.
+For example, the distributions of GPP and NBP are concentrated around the default parameterization with long tails, indicating that a small proportion of parameters have large effects (Figure \ref{fig:violins}).
+One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death.
+This simulation is likely unrealistic, indicating that the perturbation range was too large, but possibly also that the model struggles to acclimate to warmer soils.
+This represents one of the valuable outcomes of an extensive PPE, exposing unexpected model behavior, which may belie brittle parameterizations or bugs.
+
 \begin{figure}[h]
@@
 \includegraphics[width=35pc]{../figs/violins.pdf}
-\caption{Distributions of global annual gross primary production (GPP) and net biome production (NBP) across the six forcing scenarios.
-See Section \ref{sect:exps} for scenario descriptions.
-The distributions are concentrated around the default parameterization with long tails, indicating that a small proportion of parameters have large effects.
-Especially in the case of NBP, parameter effects can be larger than scenario effects.}
-\label{}
+\caption{Distributions of global annual gross primary production (GPP) and net ecosystem production (NEP) across the PPE within each of our six forcing scenarios.
+See Section \ref{sect:exps} for scenario descriptions.}
+\label{fig:violins}
 \end{figure}
+
+The PPE is likewise useful for determining which parameters influence which model processes and where.
+For example the parameters controlling leaf area index varies significantly by biome (Figure \ref{fig:lai}).
+Plant hydraulics were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
+The most influential parameters globally includes parameters from each of these three biomes.
+We have similar plots delineated by plant functional type, and for many other output variables in our extended diagnostics set (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT})
 
@@
 
+Repeating our perturbations across the six forcing scenarios ensures that we can identify important parameters not just under present-day conditions, but also in the future.
+In the case of evaporation (ET), the parameters that control the increase in ET due to warming tend to be different from the parameters that control present-day ET (Figure \ref{fig:et}).
+ET response can differ significantly from the default case (i.e. stray from the orange line), even when ET is near the default under present-day conditions (i.e. within the shaded region).
+Many of the new parameters that appear in the ET response parameter rankings govern plant acclimation to temperature. 
+These parameters, while not especially influential on present-day ET, will be important in governing the trend in ET going forward.
+
 \begin{figure}[h]
@@
 our control experiment.
-The default parameterization ET enhancement is +8.8\%, as shown with the orange line.
-The shading spans the CTL2010 default ET, plus or minus 1 W/m2.
-(b) The top 8 parameters governing global ET in the CTL2010 experiment.
+The response to warming with default parameters is ET+8.8\% (highlighted with the orange line).
+The shading spans the default ET, plus or minus 1 W/m2.
+(b) The top 8 parameters governing global ET in the control experiment.
 (c) The top 8 parameters governing ET response to the future climate anomaly forcing (AF2095).
@@
 \end{figure}
+
+The sheer volume of data generated by a large PPE can be overwhelming - our full dataset exceeds 2TB.
+Our diagnostics package, which operates on only a subset of output variables, generated 1720 figures (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
+In many cases we have found it easier to utilize interactive plotting utilities to toggle through the various parameters, output variables, and forcing scenarios (Figure \ref{fig:panel}).
+Interactive widgets are relatively easy to generate while performing data analysis, but more difficult to host online, hence the screenshot.
+Effectively leveraging the latest data science tools will greatly enhance the utility of PPE datasets.
 

### 2023-05-10 ms/makebib.sh

@@ -1,0 +1,14 @@
+grep citation ms.aux | sed 's/citation//g' | tr -d '\\{}' | sort | uniq > cites.txt
+
+d='/Users/kennedy/projects/references/refs/'
+:> refs.bib
+while read c; do
+    file=$d$c".bib"
+    if [ -f $file ]; then
+	cat $file >> refs.bib
+	echo -en '\n' >> refs.bib
+    else
+	echo "missing bib: "$c
+    fi
+done < cites.txt
+    

### 2023-05-10 ms/ms.tex
discussion

@@
 \subsection{Model spin-up}
+\label{sect:mcn}
 Model spin-up for the equilibration of carbon and nitrogen pools within biogeochemistry-enabled land models can consume up to 98\% of computational time \cite{sun2023}.
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=30pc]{../figs/sims.pdf}
-\caption{The approximate number of simulations afforded by 1 million core-hours for a range of CLM configurations.
-Configurations are labeled according to spin-up procedure (CN-matrix or the standard 1500-year spinup) and horizontal resolution (`sg' signifies sparsegrid).
-The inset map shows the locations of the 400 sparse grid cells.
-See Section \ref{methods} for spin-up and sparsegrid details.}
-\label{fig:sims}
-\end{figure}
-
-A major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
-Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times faster than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
-There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
-Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibration criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.
-
-Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
-stronger the relationship between output from the full grid and the sparsegrid.
-Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sggpp}).
-We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
-\begin{figure}[h]
-\centering
-\includegraphics[width=35pc]{../figs/sparsegrid_gpp.pdf}
-\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation.
-We opted for 400 clusters to balance computational cost against representativeness.}
-\label{fig:sggpp}
-\end{figure}
-
 Parameters were varied across a wide variety of CLM subcomponents (Figure \ref{fig:params}).
@@
 
+A major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
+Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times faster than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
+There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
+Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibration criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=30pc]{../figs/sims.pdf}
+\caption{The approximate number of simulations afforded by 1 million core-hours for a range of CLM configurations.
+Configurations are labeled according to spin-up procedure (CN-matrix or the standard 1500-year spinup) and horizontal resolution (`sg' signifies sparsegrid).
+The inset map shows the locations of the 400 sparse grid cells.
+See Section \ref{methods} for spin-up and sparsegrid details.}
+\label{fig:sims}
+\end{figure}
+
+Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
+stronger the relationship between output from the full grid and the sparsegrid.
+Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sggpp}).
+We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
+\begin{figure}[h]
+\centering
+\includegraphics[width=35pc]{../figs/sparsegrid_gpp.pdf}
+\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation.
+We opted for 400 clusters to balance computational cost against representativeness.}
+\label{fig:sg}
+\end{figure}
+
 Most of the parameter perturbations had a relatively small impact on any given output variable.
@@
 \label{fig:violins}
-\end{figure}
-
-The PPE is likewise useful for determining which parameters influence which model processes and where.
-For example the parameters controlling leaf area index varies significantly by biome (Figure \ref{fig:lai}).
-Plant hydraulics were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
-The most influential parameters globally includes parameters from each of these three biomes.
-We have similar plots delineated by plant functional type, and for many other output variables in our extended diagnostics set (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT})
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=40pc]{../figs/lai_biome.pdf}
-\caption{The eight most influential parameters on leaf area index within the CTL2010 ensemble, globally and within three biomes.}
-\label{fig:lai}
 \end{figure}
@@
 
-The sheer volume of data generated by a large PPE can be overwhelming - our full dataset exceeds 2TB.
+The PPE is likewise useful for determining which parameters influence which model processes and where.
+For example the parameters controlling leaf area index varies significantly by biome (Figure \ref{fig:lai}).
+Plant hydraulics were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
+The most influential parameters globally includes parameters from each of these three biomes.
+We have similar plots delineated by plant functional type, and for many other output variables in our extended diagnostics set (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT})
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/lai_biome.pdf}
+\caption{The eight most influential parameters on leaf area index within the CTL2010 ensemble, globally and within three biomes.}
+\label{fig:lai}
+\end{figure}
+
+The volume of data generated by a large PPE can be overwhelming - our full dataset exceeds 2TB.
 Our diagnostics package, which operates on only a subset of output variables, generated 1720 figures (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
@@
 Interactive widgets are relatively easy to generate while performing data analysis, but more difficult to host online, hence the screenshot.
-Effectively leveraging the latest data science tools will greatly enhance the utility of PPE datasets.
 
@@
 
-Major discussion points:
-\begin{itemize}
-\item need for speed
-\item parameter effects can be quite large
-\item parameter priors are difficult to establish
-\item mean and climate response variables may vary
-\item community resource
-\item invest in infrastructure
-\end{itemize}
+In this project, we have identified and perturbed 188 CLM parameters to create a large one-at-a-time parameter perturbation ensemble.
+This ensemble is useful for understanding parametric controls on CLM dynamics and has already been leveraged as a starting point for parameter optimization efforts.
+We made significant investments in software infrastructure to enable this experiment and reduce the burden of generating future PPEs.
+
+There were several barriers to perturbing the full set of CLM5-BGC parameters.
+Firstly, many parameters had not been officially identified as such.
+In such cases, we identified hard-coded values, established an appropriate parameter name, and extracted that parameter to the CLM parameter file for easier manipulation.
+In the end, we perturbed 188 parameters across a variety of CLM processes (Figure \ref{fig:params}).
+The second challenge involved defining a perturbation range for each parameter.
+We solicited expert judgment to set a minimum and maximum reasonable value for each parameter.
+In some cases, literature values were explicitly used (e.g.
+medlynslope), but the most common range was $\pm20\%$.
+The third challenge involved computational cost.
+
+Running CLM-BGC with standard spin-up at 1$^\circ$ resolution, which is the most common configuration, would be prohibitively expensive, yielding only 2.3 simulations per million core-hours (for reference the CLM annual computing allocation is ~40 million core-hours).
+We sought a 2400-simulation ensemble, requiring a much faster model configuration.
+We sped up the model approximately 500x (Figure \ref{fig:sims}) by strategically reducing the number of model grid cells and by leveraging a linearized spin-up solver (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
+We found that most of the important model dynamics could be effectively resolved using only 400 grid cells and a 140-year spin-up sequence (Figure \ref{fig:sg}, Supp Figures \ref{supp:ilamb} and zqz spinup figure).
+
+We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:violins}).
+That being said, the majority of parameter perturbations had very small effect, leading to output distributions strongly concentrated around the default simulation with long tails representing a small subset of parameters.
+In the case of gross primary production, only xx parameters had a parameter effect of more than 2\% under the control forcing scenario (Supp Figure \ref{supp:gpp}).
+As such, we would expect the the effective parameter space dimensionality for any given process would be significantly smaller than implied by the 188 parameters.
+
+We opted for six forcing scenarios to identify parameter effects not just in present-day but in response to pre-industrial or future climate conditions.
+The parameters that were most influential under present-day conditions did not necessarily match the set of parameters controlling the response to future forcing (Figure \ref{fig:et}).
+We found that many acclimation parameters, which were not as important for determining present-day evapotranspiration (ET), were among the most influential on the response of ET to future climate.
+Similarly, the parameters controlling leaf area index varied significantly depending on biome (Figure \ref{fig:lai}).
+
+A one-at-a-time PPE cannot capture parameter interactions, and our min/max sampling protocol precludes diagnosing non-linearities.
+As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty.
+The advantage of our dataset is that we can diagnose parameter effects without the uncertainty of an intermediating emulator or regression model.
+As such, it is easy to diagnose which parameters are most influential on a given process, which has been leveraged for parameter filtering.
+We have published a large set of ensemble diagnostic plots online, which serves as a valuable enhancement to our model technical documentation.
+Now in addition to seeing the definition of a given parameter, and the relevant equations, you can see the magnitude and spatial patterns of its effects (e.g.
+Figure \ref{fig:panel}).
+
+We have already begun a follow-on experiment that perturbs a subset of parameters with the latin hypercube sampling strategy to work towards calibration of leaf area index.
+By investing in the infrastructure underlying this first experiment, all of our subsequent experiments have required much less time and effort.
+We expect to continue to extend our efforts in this domain, as well as repeat these types of experiments with subsequent model releases.
 

### 2023-05-11 ms/ms.tex
edits
from overleaf

@@
 
-\title{The Community Land Model, version 5: \\ One-at-a-time Parameter Perturbation Ensemble}
+\title{The Community Land Model, version 5.1 \\ One-at-a-time Parameter Perturbation Ensemble}
 
-\authors{D. Kennedy \affil{1}, Katherine Dagon\affil{1}, David M. Lawrence\affil{1}}
-%R.A. Fisher
-%B.M. Sanderson
-%C.D. Koven
-%A.L.S. Swann
-%C.M. Zarakas
-%L.R. Hawkins
-%F. Hoffman
-%N. Collier
-%Y. Luo
-%X. Lu
-%S. Levis
-%S.C. Swenson
-%K.W. Oleson
-%E. Kluzek
-%J. Liu
+\authors{D. Kennedy\affil{1} and many more}
 
@@
 Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many variables represented within the land component of Earth System Models.
-Understanding how these variables respond to carbon dioxide concentration, and how we expect carbon dioxide concentration to evolve over time, is at the center of climate change research.  
+Understanding how these variables respond to CO$_2$ concentration, and how we expect CO$_2$ concentration to evolve over time, is at the center of climate change research.  
 Our certainty in climate model projections varies by domain and generally decreases with extended time horizons.
@@
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{bonan2018}.
-But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, we would expect predictions to become more skillful over time.
+But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, we might expect predictions to become more skillful across subsequent model generations.
 
@@
 When applied systematically, the result is a parameter perturbation ensemble (PPE), alternatively termed a perturbed physics ensemble.
-PPEs have played an important role in assessing projection uncertainty in climate models.
+PPEs have played an important role in assessing projection uncertainty in climate models (Hawkins 2019 JAMES).
 They also form a basis for automated model calibration.
@@
 
-We used k-means clustering to identify groups of grid cells with similar climates based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
+We used k-means clustering to identify groups of grid cells with similar dynamics based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
 We selected one representative grid cell from each cluster to stand in for the entire cluster.
@@
 The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.
+In some cases parameters could not be perturbed independently, because they are required to sum to 1.
+In the case of nitrogen fixation costs, we opted to perturbed the parameters independently, but also as a group (`KCN'), to enforce nitrogen limitation in lieu of switching from one uptake pathway to another.
 
@@
 \begin{landscape}
-
  \begin{table}[h]
- \caption{Some key parameters \\
- Right now I'm sorting them by domain, should I sort them alphabetically by parameter name instead?}
+ \caption{Some key parameters}
  \centering
@@
 stronger the relationship between output from the full grid and the sparsegrid.
-Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sggpp}).
+Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sg}).
 We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
@@
 Most of the parameter perturbations had a relatively small impact on any given output variable.
-For example, the distributions of GPP and NBP are concentrated around the default parameterization with long tails, indicating that a small proportion of parameters have large effects (Figure \ref{fig:violins}).
+For example, the distributions of GPP and NEP are concentrated around the default simulation with long tails, indicating that a small proportion of parameters have large effects (Figure \ref{fig:violins}).
 One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death.

### 2023-05-11 ms/ms.tex

@@
 
-Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many variables represented within the land component of Earth System Models.
-Understanding how these variables respond to CO$_2$ concentration, and how we expect CO$_2$ concentration to evolve over time, is at the center of climate change research.  
-Our certainty in climate model projections varies by domain and generally decreases with extended time horizons.
+Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many variables represented within land components of Earth System Models.
+Understanding how these variables respond directly and indirectly to CO$_2$ concentration, and how they themselves impact CO$_2$ concentrations, is at the center of climate change research.  
+Our certainty in climate model projections varies by domain and generally decreases with extended time horizons \cite{esd-13-885-2022}.
 Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. 
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{bonan2018}.
-But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, we might expect predictions to become more skillful across subsequent model generations.
+But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, combined with increasingly comprehensive land surface modelling platforms, we might expect predictions to become more skillful across subsequent model generations.
 
@@
 Within the day-to-day model development process, there is a greater emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
-However, the complexity of land models has increased to the point where understanding the full span of model function has become challenging \cite{fisher2020}, and manual model calibration is becoming intractable \cite{dagon2020}.
+However, the complexity of land models has increased to the point where understanding the full span of model function has become challenging \cite{fisher2020}, and manual model calibration appears intractable \cite{dagon2020}.
 Maintaining, improving, and interpreting complex land models benefits from thoughtful investment in software to automate and routinize important components of the development process, e.g.
@@
 Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017}.
-When applied systematically, the result is a parameter perturbation ensemble (PPE), alternatively termed a perturbed physics ensemble.
+When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.
 PPEs have played an important role in assessing projection uncertainty in climate models (Hawkins 2019 JAMES).
-They also form a basis for automated model calibration.
+They also form a basis for automated model calibration \cite{williamson2013history}.
 Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.
@@
 Relative to CLM5.0, version 5.1 includes minor bug fixes, parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}.
-The PPE experiment required additional code modifications to programmatically vary the full suite of model parameters.
+The PPE experiment required additional code modifications to systematically vary the full suite of model parameters.
 The model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
@@
 \label{sect:mcn}
-Model spin-up for the equilibration of carbon and nitrogen pools within biogeochemistry-enabled land models can consume up to 98\% of computational time \cite{sun2023}.
+Model spin-up for the equilibration of carbon and nitrogen pools within biogeochemistry-enabled land models can consume up to 98\% of computational time needed for a simulation \cite{sun2023}.
 Depending on the evaluation criteria and model configuration, CLM5 requires between 800 and 2000 years (or more) to reach steady-state conditions \cite{lawrence2019}.
-Absent equilibrium, the drift towards steady state can obscure important model dynamics or features.
-For this reason, each member of the PPE requires independent spin-up.
+In the absence of equilibrium, the drift towards steady state can obscure important model dynamics or features.
+For this reason, each member of the PPE has a unique steady state and thus requires an independent spin-up.
 
@@
 This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter).
-Certain inferences about deep soil carbon would therefore be subject to uncertainty due to spin-up concerns.
+Certain inferences about deep soil carbon are therefore subject to uncertainty due to spin-up concerns.
 
@@
 In order to manage computational cost, parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}.
-We used a clustering algorithm to achieve an alternative low resolution configuration.
+Here, we use a clustering algorithm to achieve an alternative low resolution configuration.
 
@@
 In the preparation for this experiment, 188 CLM-BGC parameters were identified.
-We decided to vary each parameter independently, to a low and high value.
+We decided to vary each parameter independently, to a low and high values.
 To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter.
@@
 We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
-Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the mean climate change signal.
+Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the corresponding mean climate change signal (see Table \ref{tab:exps}).
 We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.
@@
 A total of 188 parameters were targeted for sensitivity testing.
-Many parameters were hard-coded, but have now been extracted to the CLM parameter file for easier manipulation.
+Many parameters were originally hard-coded, but have now been extracted to the CLM parameter file for easier manipulation.
 In some cases new parameters were introduced to allow perturbation of important processes.
@@
 Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
-stronger the relationship between output from the full grid and the sparsegrid.
+better the approximation oif output from the full grid can be acheived with the sparsegrid.
 Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sg}).
@@
 Most of the parameter perturbations had a relatively small impact on any given output variable.
-For example, the distributions of GPP and NEP are concentrated around the default simulation with long tails, indicating that a small proportion of parameters have large effects (Figure \ref{fig:violins}).
+For example, the distributions of GPP and NEP are concentrated around the default simulation with long tails, indicating that only a small proportion of parameters have large effects (Figure \ref{fig:violins}).
 One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death.

### 2023-05-16 ms/ms.tex
overleaf

@@
 
-Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many variables represented within land components of Earth System Models.
-Understanding how these variables respond directly and indirectly to CO$_2$ concentration, and how they themselves impact CO$_2$ concentrations, is at the center of climate change research.  
+Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within land components of Earth System Models.
+Understanding how these processes respond directly and indirectly to CO$_2$ concentrations, and how they themselves impact CO$_2$ concentrations, is at the center of climate change research.  
 Our certainty in climate model projections varies by domain and generally decreases with extended time horizons \cite{esd-13-885-2022}.
@@
 
-While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations.
-Within the day-to-day model development process, there is a greater emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
+While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations (McNeall et al.
+2016 ESD).
+Within the model development process, there is a significant emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
 However, the complexity of land models has increased to the point where understanding the full span of model function has become challenging \cite{fisher2020}, and manual model calibration appears intractable \cite{dagon2020}.
@@
 
-One such fundamental model development step is parameter sensitivity testing and/or parameter optimization.
-Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017}.
+One such fundamental model development step is parameter sensitivity testing and/or parameter optimization (Qian et al. 2018 JGR Atmos).
+Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017} (Balaji et al. 2022 PNAS).
 When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.
-PPEs have played an important role in assessing projection uncertainty in climate models (Hawkins 2019 JAMES).
-They also form a basis for automated model calibration \cite{williamson2013history}.
+PPEs have played an important role in assessing projection uncertainty in climate models (Hawkins 2019 JAMES), (Peatier et al.
+2022 GRL, Sanderson et al.
+2008 JClim, Tett et al.
+2022 JClim, Yamazaki et al.
+2021 Clim Dyn).
+PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013history} or Bayesian uncertainty quantification (Cleary et al. 2021 JCP).
 Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.
 In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
-In this paper we present that dataset, how it was produced, and its potential applications.
+In this paper we present that dataset, describe how it was produced, and survey some potential applications.
 
@@
 \label{sect:md}
-The Community Terrestrial System Model (CTSM) is developed by the CESM Land Model Working Group and maintained at the National Center for Atmospheric Research (NCAR).
+The Community Terrestrial Systems Model (CTSM) is developed by the CESM Land Model Working Group and maintained at the National Center for Atmospheric Research (NCAR).
 This experiment utilizes the Community Land Model configuration of CTSM, version 5.1 (CLM5.1).
@@
 This new module utilizes a linearized simplification of CLM's biogeochemistry to significantly reduce spin-up time.
-Our spin-up protocol featured 20 years in accelerated decomposition mode, followed by 80 years of Matrix-CN, followed by 40 years of `normal' mode, cycling over a ten-year forcing dataset.
-This was designed to achieve sufficiently equilibrated model states, while minimizing computational time.
-This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter).
+Our spin-up protocol featured 20 years in accelerated decomposition mode, followed by 80 years of Matrix-CN, followed by 40 years of `normal' mode, cycling over a ten-year forcing dataset (described below).
+This protocol was designed to achieve sufficiently equilibrated model states, while minimizing computational time.
+This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter depth).
 Certain inferences about deep soil carbon are therefore subject to uncertainty due to spin-up concerns.
@@
 \begin{table}[h]
-\caption{Clustering inputs}
+\caption{Clustering inputs categorized into three groups, with the CLM variable name in parentheses}
 \centering
@@
  \hline
-d\_max & Dry surface layer parameter & Sensible, latent heat and momentum fluxes \\
+d\_max & Dry surface layer (DSL) parameter & Sensible, latent heat and momentum fluxes \\
 frac\_sat\_soil\_dsl\_init & Fraction of saturated soil at which DSL initiates & Sensible, latent heat and momentum fluxes \\
@@
 There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
-Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibration criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.
+Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.
 
@@
 \centering
-\includegraphics[width=35pc]{../figs/sparsegrid_gpp.pdf}
-\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation.
+\includegraphics[width=25pc]{../figs/sparsegrid_gpp.pdf}
+\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last thirty years of a transient CLM5.1 simulation.
 We opted for 400 clusters to balance computational cost against representativeness.}
@@
 The PPE is likewise useful for determining which parameters influence which model processes and where.
-For example the parameters controlling leaf area index varies significantly by biome (Figure \ref{fig:lai}).
-Plant hydraulics were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
-The most influential parameters globally includes parameters from each of these three biomes.
-We have similar plots delineated by plant functional type, and for many other output variables in our extended diagnostics set (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT})
+For example the parameters controlling leaf area index vary significantly by biome (Figure \ref{fig:lai}).
+Plant hydraulics parameters were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
+The most influential parameters globally include parameters from each of these three biomes.
+We have similar plots delineated by plant functional type, and for many other output variables in our extended diagnostics set (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
 
@@
 In this project, we have identified and perturbed 188 CLM parameters to create a large one-at-a-time parameter perturbation ensemble.
-This ensemble is useful for understanding parametric controls on CLM dynamics and has already been leveraged as a starting point for parameter optimization efforts.
+This ensemble is useful for understanding parametric controls on CLM processes and has already been leveraged as a starting point for parameter optimization efforts.
 We made significant investments in software infrastructure to enable this experiment and reduce the burden of generating future PPEs.
@@
 In the case of gross primary production, only xx parameters had a parameter effect of more than 2\% under the control forcing scenario (Supp Figure \ref{supp:gpp}).
-As such, we would expect the the effective parameter space dimensionality for any given process would be significantly smaller than implied by the 188 parameters.
+As such, we would expect that the dimensionality of the effective parameter space for any given process would be significantly smaller than implied by the 188 parameters.
 
@@
 As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty.
-The advantage of our dataset is that we can diagnose parameter effects without the uncertainty of an intermediating emulator or regression model.
+The advantage of our dataset is that we can diagnose parameter effects without the uncertainty of an approximating emulator or regression model.
 As such, it is easy to diagnose which parameters are most influential on a given process, which has been leveraged for parameter filtering.
 We have published a large set of ensemble diagnostic plots online, which serves as a valuable enhancement to our model technical documentation.
-Now in addition to seeing the definition of a given parameter, and the relevant equations, you can see the magnitude and spatial patterns of its effects (e.g.
+Now in addition to seeing the definition of a given parameter, and the relevant equations, a model user can see the magnitude and spatial patterns of its effects (e.g.
 Figure \ref{fig:panel}).
 
-We have already begun a follow-on experiment that perturbs a subset of parameters with the latin hypercube sampling strategy to work towards calibration of leaf area index.
+We have already begun a follow-on experiment that perturbs a subset of parameters with the Latin hypercube sampling strategy to work towards calibrating leaf area index.
 By investing in the infrastructure underlying this first experiment, all of our subsequent experiments have required much less time and effort.

### 2023-05-16 ms/ms.tex
cites

@@
 Understanding how these processes respond directly and indirectly to CO$_2$ concentrations, and how they themselves impact CO$_2$ concentrations, is at the center of climate change research.  
-Our certainty in climate model projections varies by domain and generally decreases with extended time horizons \cite{esd-13-885-2022}.
+Our certainty in climate model projections varies by domain and generally decreases with extended time horizons \cite{koven2022}.
 Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. 
@@
 
-While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations (McNeall et al.
-2016 ESD).
+While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
 Within the model development process, there is a significant emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
@@
 2021 Clim Dyn).
-PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013history} or Bayesian uncertainty quantification (Cleary et al. 2021 JCP).
+PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013} or Bayesian uncertainty quantification (Cleary et al. 2021 JCP).
 Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.

### 2023-05-16 ms/ms.tex
overleaf

@@
 
-One such fundamental model development step is parameter sensitivity testing and/or parameter optimization (Qian et al. 2018 JGR Atmos).
-Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017} (Balaji et al. 2022 PNAS).
+One such fundamental model development step is parameter sensitivity testing and/or parameter optimization \cite{qian2018}.
+Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017,balaji2022}.
 When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.

### 2023-05-16 ms/ms.tex
overleaf

@@
 When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.
-PPEs have played an important role in assessing projection uncertainty in climate models (Hawkins 2019 JAMES), (Peatier et al.
-2022 GRL, Sanderson et al.
-2008 JClim, Tett et al.
-2022 JClim, Yamazaki et al.
-2021 Clim Dyn).
-PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013} or Bayesian uncertainty quantification (Cleary et al. 2021 JCP).
+PPEs have played an important role in assessing projection uncertainty in climate models \cite{sanderson2008,hawkins2019,yamazaki2021,peatier2022,tett2022}.
+PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013} or Bayesian uncertainty quantification \cite{cleary2021}.
 Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.

### 2023-05-16 ms/ms.tex
overleaf

@@
 \journalname{Journal of Advances in Modeling Earth Systems (JAMES)}
-
 \begin{document}
-
 \title{The Community Land Model, version 5.1 \\ One-at-a-time Parameter Perturbation Ensemble}
 
-\authors{D. Kennedy\affil{1} and many more}
+\authors{D. Kennedy\affil{1} and ...}
 
 \affiliation{1}{Climate and Global Dynamics Laboratory, NCAR, Boulder, CO, USA.}
-
 \correspondingauthor{Daniel Kennedy}{djk2120@ucar.edu}
@@
 \item enter point 1 here
-\item enter point 2 here
-\item enter point 3 here
 \end{keypoints}
@@
 \section{Introduction}
-
 Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within land components of Earth System Models.
-Understanding how these processes respond directly and indirectly to CO$_2$ concentrations, and how they themselves impact CO$_2$ concentrations, is at the center of climate change research.  
+Understanding how these processes respond directly and indirectly to CO$_2$ concentrations, and how they themselves impact CO$_2$ concentrations, is a critical facet of climate change research.  
 Our certainty in climate model projections varies by domain and generally decreases with extended time horizons \cite{koven2022}.
 Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. 
-A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{bonan2018}.
+A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
 But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, combined with increasingly comprehensive land surface modelling platforms, we might expect predictions to become more skillful across subsequent model generations.

### 2023-05-17 ms/ms.tex
Updates from Overleaf
overleaf

@@
 \centering
-\includegraphics[width=35pc]{../figs/violins.pdf}
-\caption{Distributions of global annual gross primary production (GPP) and net ecosystem production (NEP) across the PPE within each of our six forcing scenarios.
-See Section \ref{sect:exps} for scenario descriptions.}
+\includegraphics[width=\textwidth]{../figs/violins.png}
+\caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across the PPE within each of our six forcing scenarios.
+Each point represents a simulation within the PPE.
+Points are spaced horizontally to minimize overlap.}
+\label{fig:violins}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/violins.png}
+\caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across the PPE within each of our six forcing scenarios.
+Each point represents a simulation within the PPE.
+Points are spaced horizontally to minimize overlap.}
 \label{fig:violins}

### 2023-05-17 ms/ms.tex
Updates from Overleaf

@@
 \caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across the PPE within each of our six forcing scenarios.
-Each point represents a simulation within the PPE.
-Points are spaced horizontally to minimize overlap.}
+Circles are drawn for 5 and 0.1 PgC bins respectively, with the circle size proportional to the number of simulations falling in that bin.
+In each case the largest circle is centered at the default simulation.}
 \label{fig:violins}
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/violins.png}
-\caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across the PPE within each of our six forcing scenarios.
-Each point represents a simulation within the PPE.
-Points are spaced horizontally to minimize overlap.}
+\includegraphics[width=\textwidth]{../figs/variance.pdf}
+\caption{needs caption}
 \label{fig:violins}

### 2023-05-22 ms/ms.tex
Updates from Overleaf

--- 2023-05-17 ms/ms.tex
@@
 When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.
-PPEs have played an important role in assessing projection uncertainty in climate models \cite{sanderson2008,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013} or Bayesian uncertainty quantification \cite{cleary2021}.
+PPEs have played an important role in assessing projection uncertainty in climate models (Booth 2012; ERL, Murphy 2004; Nature) \cite{sanderson2008,hawkins2019,yamazaki2021,peatier2022,tett2022}.
+PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013} (Williamson2017; GMD-NEMO calibration; Couvreux2020 JAMES; Hourdin 2020 JAMES) or Bayesian uncertainty quantification \cite{cleary2021}.
 Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.
@@
 \centering
-\includegraphics[width=30pc]{../figs/bar.pdf}
-\caption{206 parameters were identified and perturbed across the various domains of the land model.}
+\includegraphics[width=\textwidth]{../figs/bar.pdf}
+\caption{193 parameters were perturbed across the various domains of the land model.}
 \label{fig:params}
@@
 \caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across the PPE within each of our six forcing scenarios.
-Each point represents a simulation within the PPE.
-Points are spaced horizontally to minimize overlap.}
+Markers are drawn for 10 and 0.28 PgC bins for GPP and NEP, respectively.
+The marker area is proportional to the number of simulations falling in that bin.
+In each case the ellipse is centered about the default simulation.}
 \label{fig:violins}
@@
 \centering
-\includegraphics[width=40pc]{../figs/lai_biome.pdf}
+\includegraphics[width=\textwidth]{../figs/lai_biome.pdf}
 \caption{The eight most influential parameters on leaf area index within the CTL2010 ensemble, globally and within three biomes.}

### 2023-05-22 ms/ms.tex
overleaf

@@
 When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.
-PPEs have played an important role in assessing projection uncertainty in climate models (Booth 2012; ERL, Murphy 2004; Nature) \cite{sanderson2008,hawkins2019,yamazaki2021,peatier2022,tett2022}.
+PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
 PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013} (Williamson2017; GMD-NEMO calibration; Couvreux2020 JAMES; Hourdin 2020 JAMES) or Bayesian uncertainty quantification \cite{cleary2021}.
@@
 \centering
-\includegraphics[width=40pc]{../figs/ET_response.pdf}
+\includegraphics[width=\textwidth]{../figs/ET_response.pdf}
 \caption{(a) Global evapotranspiration in the warm forcing scenario vs.

### 2023-05-22 ms/ms.tex
overleaf

@@
 When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.
-PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013} (Williamson2017; GMD-NEMO calibration; Couvreux2020 JAMES; Hourdin 2020 JAMES) or Bayesian uncertainty quantification \cite{cleary2021}.
+PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy_2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
+PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.
 Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.

### 2023-05-22 ms/ms.tex
refs

@@
 When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.
-PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy_2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
+PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
 PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.

### 2023-05-22 ms/ms.tex

@@
 We extend that work by sampling a more exhaustive set of parameters across a broader range of land model processes.
-A set of approximately 200 parameters were varied across a range of climate scenarios.
-This dataset has already demonstrated great utility for diagnosing parameter effects, while the software and modeling infrastructure has greatly our expanded our capability to generate insights about CLM dynamics and uncertainties.
+A set of 193 parameters were varied across a range of climate scenarios.
+This dataset has already demonstrated great utility for diagnosing parameter effects, while the software and modeling infrastructure has greatly our expanded our capability to generate insights about parameter effects and uncertainties within CLM.
 
@@
 The set of representative grid cells comprise a `sparsegrid', which are used in lieu of a `coarse' grid.
-A paint-by-number approach is used to recompose mapped output and compute global means, where the output from the representative grid cell is substituted for all members of the cluster cohort.
+To recompose mapped output and compute global means, the output from the representative grid cell is substituted for all members of the cluster cohort.
 
@@
 
-Parameters were varied across a wide variety of CLM subcomponents (Figure \ref{fig:params}).
-A total of 188 parameters were targeted for sensitivity testing.
-Many parameters were originally hard-coded, but have now been extracted to the CLM parameter file for easier manipulation.
+A total of 193 parameters were targeted for sensitivity testing across a wide variety of CLM subcomponents (Figure \ref{fig:params}).
 In some cases new parameters were introduced to allow perturbation of important processes.
@@
 Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
-better the approximation oif output from the full grid can be acheived with the sparsegrid.
-Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sg}).
+better the approximation of output from the full grid can be achieved with the sparsegrid (Figure \ref{fig:sg}).
+Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Supp Figure zqz).
 We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
@@
 \includegraphics[width=25pc]{../figs/sparsegrid_gpp.pdf}
-\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last thirty years of a transient CLM5.1 simulation.
+\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last forty years of a transient CLM5.1 simulation.
 We opted for 400 clusters to balance computational cost against representativeness.}

### 2023-05-22 ms/ms.tex

@@
 \centering
-\includegraphics[width=\textwidth]{../figs/ET_response.pdf}
-\caption{(a) Global evapotranspiration in the warm forcing scenario vs.
+\includegraphics[width=\textwidth]{../figs/ET_response.png}
+\caption{(a) Global annual average evapotranspiration (ET) in the warm forcing scenario vs.
 our control experiment.
-The response to warming with default parameters is ET+8.8\% (highlighted with the orange line).
-The shading spans the default ET, plus or minus 1 W/m2.
+The shading spans the default ET in the control experiment, $\pm$1 W/m2.
+The response of ET to future climate with default parameters is +8.8\%.
+The orange line shows the expected future ET should the model conform to the default ET response.
 (b) The top 8 parameters governing global ET in the control experiment.
-(c) The top 8 parameters governing ET response to the future climate anomaly forcing (AF2095).
-Parameters governing the response to temperature anomalies tend to differ from the parameters controlling present-day ET.}
+(c) The top 8 parameters governing ET response to future climate, i.e. the ratio of future ET to control ET.
+Parameters governing the change in ET expected in the future tend to differ from the parameters controlling present-day ET.}
 \label{fig:et}

### 2023-05-22 ms/ms.tex
Updates from Overleaf

@@
 ET response can differ significantly from the default case (i.e. stray from the orange line), even when ET is near the default under present-day conditions (i.e. within the shaded region).
-Many of the new parameters that appear in the ET response parameter rankings govern plant acclimation to temperature. 
-These parameters, while not especially influential on present-day ET, will be important in governing the trend in ET going forward.
+All of the new parameters that appear in the ET response parameter rankings govern plant acclimation to temperature. 
+These parameters, while not especially influential on present-day ET, are among the most important in governing the trend in ET going forward.
 
@@
 \includegraphics[width=\textwidth]{../figs/lai_biome.pdf}
-\caption{The eight most influential parameters on leaf area index within the CTL2010 ensemble, globally and within three biomes.}
+\caption{The eight most influential parameters on leaf area index within the control ensemble, globally and within three biomes.}
 \label{fig:lai}
@@
 In many cases we have found it easier to utilize interactive plotting utilities to toggle through the various parameters, output variables, and forcing scenarios (Figure \ref{fig:panel}).
-Interactive widgets are relatively easy to generate while performing data analysis, but more difficult to host online, hence the screenshot.
+Interactive widgets are relatively easy to generate within a data analysis session, but more difficult to host online, hence the screenshot.
 
@@
 \centering
-\includegraphics[width=40pc]{../figs/deltmap_panel.pdf}
+\includegraphics[width=\textwidth]{../figs/deltmap_panel.pdf}
 \caption{Screenshot of interactive diagnostic for exploring maps of parameter effects.
-In this case, the effect of medlynslope on runoff within the CTL2010 ensemble.
-Increasing medlynslope tends to increase transpiration and reduce runoff.}
+In this case, the effect of medlynslope on runoff within the control ensemble.
+Increasing medlynslope tends to reduce runoff, but only in regions with sufficient vegetation activity.}
 \label{fig:panel}

### 2023-05-22 ms/ms.tex

@@
 \caption{needs caption}
-\label{fig:violins}
+\label{fig:variance}
 \end{figure}
@@
 
-In this project, we have identified and perturbed 188 CLM parameters to create a large one-at-a-time parameter perturbation ensemble.
+In this project, we have identified and perturbed 193 CLM parameters to create a large one-at-a-time PPE.
 This ensemble is useful for understanding parametric controls on CLM processes and has already been leveraged as a starting point for parameter optimization efforts.
@@
 In such cases, we identified hard-coded values, established an appropriate parameter name, and extracted that parameter to the CLM parameter file for easier manipulation.
-In the end, we perturbed 188 parameters across a variety of CLM processes (Figure \ref{fig:params}).
+In the end, we perturbed 193 parameters across a variety of CLM processes (Figure \ref{fig:params}).
 The second challenge involved defining a perturbation range for each parameter.
@@
 medlynslope), but the most common range was $\pm20\%$.
-The third challenge involved computational cost.
 
-Running CLM-BGC with standard spin-up at 1$^\circ$ resolution, which is the most common configuration, would be prohibitively expensive, yielding only 2.3 simulations per million core-hours (for reference the CLM annual computing allocation is ~40 million core-hours).
-We sought a 2400-simulation ensemble, requiring a much faster model configuration.
-We sped up the model approximately 500x (Figure \ref{fig:sims}) by strategically reducing the number of model grid cells and by leveraging a linearized spin-up solver (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
-We found that most of the important model dynamics could be effectively resolved using only 400 grid cells and a 140-year spin-up sequence (Figure \ref{fig:sg}, Supp Figures \ref{supp:ilamb} and zqz spinup figure).
+The third challenge involved managing computational cost.
+Quantifying parameter effects is a necessary prerequisite to automated calibration and uncertainty quantification.
+Because the parameter space of CLM is quite large and the model response potentially non-linear, large ensembles are required to adequately resolve response surfaces.
+With standard CLM configurations, such ensembles would far exceed our computing allocation.
+As such, the computational cost of inferring parameter effects is a major constraint.
+We were able to speed up the model approximately 500x by strategically reducing the number of model grid cells (Figure \ref{fig:sg}) and by leveraging a linearized spin-up solver (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
+Designing faster model configurations and more efficient sampling strategies will greatly aid in model calibration and uncertainty quantification.
 
+For this experiment we opted for a one-at-a-time perturbation strategy, testing a minimum and maximum value for each of 193 parameters.
 We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:violins}).
-That being said, the majority of parameter perturbations had very small effect, leading to output distributions strongly concentrated around the default simulation with long tails representing a small subset of parameters.
-In the case of gross primary production, only xx parameters had a parameter effect of more than 2\% under the control forcing scenario (Supp Figure \ref{supp:gpp}).
-As such, we would expect that the dimensionality of the effective parameter space for any given process would be significantly smaller than implied by the 188 parameters.
+That being said, the majority of parameter perturbations had very small effects, such that a majority of simulations were clustered around the default simulation.
+In the case of canopy evaporation, more than 90\% of the ensemble variance could be explained by the ten most influential parameters (Figure \ref{fig:variance}).
+As such, we would expect that the dimensionality of the effective parameter space for any given process would be significantly smaller than implied by the 193 parameters.
 
@@
 As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty.
-The advantage of our dataset is that we can diagnose parameter effects without the uncertainty of an approximating emulator or regression model.
-As such, it is easy to diagnose which parameters are most influential on a given process, which has been leveraged for parameter filtering.
+The advantage of our dataset is that we can diagnose parameter effects without the uncertainty contributed by an  emulator or regression model.
+As such, it is easy to diagnose which parameters are most influential on a given process, which has already been leveraged for parameter filtering.
 We have published a large set of ensemble diagnostic plots online, which serves as a valuable enhancement to our model technical documentation.

### 2023-05-23 ms/ms.tex
overleaf

@@
 \centering
-\includegraphics[width=\textwidth]{../figs/violins.png}
-\caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across the PPE within each of our six forcing scenarios.
-Markers are drawn for 10 and 0.28 PgC bins for GPP and NEP, respectively.
-The marker area is proportional to the number of simulations falling in that bin.
-In each case the ellipse is centered about the default simulation.}
-\label{fig:violins}
+\includegraphics[width=\textwidth]{../figs/ranges.png}
+\caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across six forcing scenarios.
+Circles mark the CLM5.1 default simulation, and the bars span the ensemble ranges.}
+\label{fig:ranges}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{../figs/variance.pdf}
-\caption{needs caption}
+\caption{Cumulative fraction of variance explained by the most influential parameters on gross primary production and canopy evaporation.
+For example approximately 96\% of canopy evaporation variance is explained by the first ten parameters regardless of forcing, whereas between 74 and 92\% of GPP variance is explained by the first ten parameters.
+Only the top 50 parameters are shown (of 193) to improve the visualization.
+Note that parameter rankings differ by variable.}
 \label{fig:variance}

### 2023-05-23 ms/ms.tex
Updates from Overleaf

@@
 
-Most of the parameter perturbations had a relatively small impact on any given output variable.
-For example, the distributions of GPP and NEP are concentrated around the default simulation with long tails, indicating that only a small proportion of parameters have large effects (Figure \ref{fig:violins}).
+Parameter effects could be quite large, in some cases exceeding the magnitude of switching scenarios (Figure \ref{fig:ranges}).
 One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death.
-This simulation is likely unrealistic, indicating that the perturbation range was too large, but possibly also that the model struggles to acclimate to warmer soils.
-This represents one of the valuable outcomes of an extensive PPE, exposing unexpected model behavior, which may belie brittle parameterizations or bugs.
+A perturbation of $\pm$20\% likely exceeds the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils.
+Carrying out an extensive PPE increases the possibility exposing unexpected model behavior, which may belie unforeseen tipping points, brittle parameterizations,  and/or bugs.
+
+A small fraction of parameters tend to explain a large amount of the ensemble variance (Figure \ref{fig:variance}).
+For example the same four parameters explain up to 90\% of the canopy evaporation variance across all the forcing scenarios.
+Conversely, most of the parameter perturbations had a relatively small impact on any given output variable.
+As such, the effective parameter space dimensionality for any given CLM process is expected to be much smaller than would be expected from having 193 independent parameters.
 

### 2023-08-02 ms/ms.tex
resolving comments

@@
 Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within land components of Earth System Models.
-Understanding how these processes respond directly and indirectly to CO$_2$ concentrations, and how they themselves impact CO$_2$ concentrations, is a critical facet of climate change research.  
+Understanding how these processes respond to CO$_2$ concentrations, and how they themselves impact CO$_2$ concentrations, is a critical facet of climate change research.  
 Our certainty in climate model projections varies by domain and generally decreases with extended time horizons \cite{koven2022}.
@@
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
-But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, combined with increasingly comprehensive land surface modelling platforms, we might expect predictions to become more skillful across subsequent model generations.
+Effectively synthesizing the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns can help to parameterize increasingly comprehensive land surface modelling platforms
 
-While model inter-comparison projects have had tremendous utility, it can be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
-Within the model development process, there is a significant emphasis on hypothesis testing, substituting one parameterization at a time to understand the various consequences.
-However, the complexity of land models has increased to the point where understanding the full span of model function has become challenging \cite{fisher2020}, and manual model calibration appears intractable \cite{dagon2020}.
+Uncertainty in the model projections of the carbon cycle is manifest in both uncertainty around the appropriate algorithms and model structures used, and also in the parameterizations of these models.
+Model inter-comparison projects (MIPs) are the primary means by which structural uncertainty is assessed (refs).
+While these have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
+To allow for more tractable experimental design each model is typically allowed only a single parameterization of many plausible parameter sets.
+Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land surface trajectories.
+
+The process of parameter selection for land surface models is challenging on account of the intrinsic  complexity of the heterogenous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryology, biogeochemistry, physiology, land management, micrometeorology, etc.).
+The number of model parameters required to represent all the relevant processes is large, and model speeds are relatively slow, presenting a major barrier to robust objective calibration, model interpretation and development \cite{fisher2020,dagon2020}.
+Thorough assessment of the parameter sensitivity of land surfaces models in a manner that is repeatable, open and integrated with ongoing code development is desirable but very challenging \cite{hourdin2017,balaji2022}.
+It typically requires extensive computational time,  software engineering, and domain-specific expert scientific knowledge.
+
+Several promising avenues for model calibration are in development (lavendar, musa, CES), many of which require the construction of a large parameter perturbation ensembles (PPE) \cite{qian2018}. 
+PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
+PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.
+A major goal of this project is to enable automated calibration of the Community Land Model (CLM) \cite{lawrence2019}.
+This extends the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
+Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization.
+
 Maintaining, improving, and interpreting complex land models benefits from thoughtful investment in software to automate and routinize important components of the development process, e.g.
 \citeA{collier2018}.
-
-One such fundamental model development step is parameter sensitivity testing and/or parameter optimization \cite{qian2018}.
-Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017,balaji2022}.
-When applied systematically, the result is a Parameter Perturbation Ensemble (PPE), alternatively termed a Perturbed Physics Ensemble.
-PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.
 Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.
@@
 In this paper we present that dataset, describe how it was produced, and survey some potential applications.
-
-We are working with the Community Land Model (CLM), the land component of the Community Earth System Model, which is developed and maintained at the National Center for Atmospheric Research (NCAR).
-This work follows up on \citeA{dagon2020}, which began the process of systematically varying CLM parameters in service of automated model calibration.
-We extend that work by sampling a more exhaustive set of parameters across a broader range of land model processes.
-A set of 193 parameters were varied across a range of climate scenarios.
-This dataset has already demonstrated great utility for diagnosing parameter effects, while the software and modeling infrastructure has greatly our expanded our capability to generate insights about parameter effects and uncertainties within CLM.
+The dataset has already demonstrated great utility for diagnosing parameter effects (cheng,yan), while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CLM.
 
@@
 Relative to CLM5.0, version 5.1 includes minor bug fixes, parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}.
-The PPE experiment required additional code modifications to systematically vary the full suite of model parameters.
-The model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
-We utilized the biogeochemistry version of CLM in land-only mode, with the crop model turned off.
-The component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP}
+The PPE experiment required additional code modifications to systematically vary the full suite of model parameters (many of which were previously `hard-coded').
+We utilized the biogeochemical cycling version of CLM, which includes active carbon and nitrogen cycles.
+This is  as opposed to the satellite phenology mode, which is driven by observed canopy properties, and was explored in \citeA{dagon2020}.
+We used the model in land-only mode driven by observed meteorological forcing from the GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}), with the crop model turned off.
 
@@
 In the absence of equilibrium, the drift towards steady state can obscure important model dynamics or features.
-For this reason, each member of the PPE has a unique steady state and thus requires an independent spin-up.
+Because each member of the PPE can have a unique steady state, we performed an independent spin-up for each member.
 
@@
 Multivariate spatio-temporal clustering (MVSC) has been utilized to extract patterns of climatological significance from climate model output \cite{hoffman2005} and applied to design a representativeness-based sampling network \cite{hoffman2013}.
-Instead of lowering resolution by coarsening a rectilinear grid, we used MVSC to strategically remove redundant grid cells, leaving only 400 grid cells that efficiently sample important model dynamics.
+Instead of lowering resolution by coarsening a rectilinear grid, we here used MVSC to strategically remove redundant grid cells, leaving only 400 grid cells that efficiently sample important model dynamics.
 
@@
 We tested 15 values of $k$, ranging from 10 to 800.
-Based on ILAMB2.5 benchmarking \cite{collier2018} against the full grid output, we opted for a 400-cluster sparsegrid, to balance computational cost against model fidelity (Supp Figure \ref{supp:ilamb}).
+Utilizing the ILAMB2.5 benchmarking software \cite{collier2018}, we calculated skill scores to quantify how well each candidate sparsegrid mirrored the full grid output (Supp Figure \ref{supp:ilamb}).
+We opted for a 400-cluster sparsegrid, to balance computational cost against model fidelity.
 Because our emphasis is on vegetated regions, we masked out Antarctica within the clustering algorithm, whereby we do not provide any output below 60$^\circ$S.
@@
 \subsection{Experimental Design}
-In the preparation for this experiment, 188 CLM-BGC parameters were identified.
-We decided to vary each parameter independently, to a low and high values.
+Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants previously considered as `internal' are hard-coded and not amenable to systematic perturbation.
+For the purposes of this activity, we identified what we consider the full set of 188 CLM-BGC parameters, and enabled their modification via the model input files.
+
+Given the very large parameter space, to conduct an initial analysis of the response of the model to parametric uncertainty, we decided to vary each parameter independently, exploring low and high values.
 To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter.
@@
 The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.
-In some cases parameters could not be perturbed independently, because they are required to sum to 1.
-In the case of nitrogen fixation costs, we opted to perturbed the parameters independently, but also as a group (`KCN'), to enforce nitrogen limitation in lieu of switching from one uptake pathway to another.
+In some cases parameters could not be perturbed independently, because they feature inherent co-dependence.
+In the case of nitrogen fixation costs, we opted to perturb the parameters independently, but also as a group (`KCN'), to enforce nitrogen limitation in lieu of switching from one uptake pathway to another.
+We likewise opted to perturb soil hydraulic parameters directly (e.g.
+saturated hydraulic conductivity) and also via perturbations to the soil texture (percent sand, clay, and silt).
 
-Each simulation ran for 150 years, with the first 140 for spin-up, followed by a 10-year period for analysis.
-We opted for six different forcing scenarios to understand the intersection of parameter effects and climate change (Table \ref{tab:exps}).
+Each simulation ran for 150 years, with the first 140 for spin-up (as described above in Section \ref{sect:mcn}), followed by a 10-year period for analysis.
+We opted for six different forcing scenarios to understand the intersection of parameter effects and the different forcing agents associated with climate change (Table \ref{tab:exps}).
 The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and is the default forcing data for CLM5 \cite{lawrence2019}.
@@
 We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.
+Our future climate scenario utilizes the SSP3-7.0 forcing in order to allow comparison with the existing CESM2 large ensemble.
 
@@
 A total of 193 parameters were targeted for sensitivity testing across a wide variety of CLM subcomponents (Figure \ref{fig:params}).
-In some cases new parameters were introduced to allow perturbation of important processes.
-For example a sand perturbation factor ($sand\_pf$) was introduced in order to perturb the percent sand in the soil column without creating a new surface dataset.
-A spreadsheet detailing all of the parameters, and their ranges, is provided in the supplementary material and available online (zqz link).
 
@@
 
-A major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
+Given the challenging size of the CLM5-BGC parameter space, and the need to probe and document parametric uncertainties within operational land surface models, a  major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
 Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times faster than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
@@
 
-Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
-better the approximation of output from the full grid can be achieved with the sparsegrid (Figure \ref{fig:sg}).
-Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Supp Figure zqz).
+Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity.
+Generally the more clusters, the better the approximation of output from the full grid that can be achieved with the sparsegrid (Figure \ref{fig:sg}).
+Accurate reconstruction of of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Supp Figure zqz).
 We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
@@
 
-Parameter effects could be quite large, in some cases exceeding the magnitude of switching scenarios (Figure \ref{fig:ranges}).
+Overall, we found substantial impacts of parametric uncertainty on model behaviour, in some cases exceeding the magnitude of climate scenario effects (Figure \ref{fig:ranges}).
 One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death.
 A perturbation of $\pm$20\% likely exceeds the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils.
-Carrying out an extensive PPE increases the possibility exposing unexpected model behavior, which may belie unforeseen tipping points, brittle parameterizations,  and/or bugs.
+Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, which may belie unforeseen tipping points, brittle parameterizations,  and/or bugs.
 
 A small fraction of parameters tend to explain a large amount of the ensemble variance (Figure \ref{fig:variance}).
-For example the same four parameters explain up to 90\% of the canopy evaporation variance across all the forcing scenarios.
+For example the same four parameters explain up to 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig \ref{supp:fcev}).
 Conversely, most of the parameter perturbations had a relatively small impact on any given output variable.
-As such, the effective parameter space dimensionality for any given CLM process is expected to be much smaller than would be expected from having 193 independent parameters.
+As such, the effective parameter space dimensionality for any given CLM process is expected to be much smaller than would be expected from having 193 independent parameters, reflecting the fact that land surface models like CLM comprise representations of numerous interconnected domains.
+While the principle of LSMs is that these are all interlinked via complex feedback processes, many parameters nonetheless have impacts that are mostly limited to their own domain.
 
@@
 
-Repeating our perturbations across the six forcing scenarios ensures that we can identify important parameters not just under present-day conditions, but also in the future.
-In the case of evaporation (ET), the parameters that control the increase in ET due to warming tend to be different from the parameters that control present-day ET (Figure \ref{fig:et}).
+We repeated the full set of parameter perturbations across the six forcing scenarios in  see Table \ref{tab:exps}.
+This ensures that we can identify important parameters not just under present-day conditions, but also those that control responses to individual forcing variables (CO2, nitrogen deposition) and climate  In the case of evaporation (ET), for example, the parameters that control the increase in ET due to warming tend to be different from the parameters that control present-day ET (Figure \ref{fig:et}).
 ET response can differ significantly from the default case (i.e. stray from the orange line), even when ET is near the default under present-day conditions (i.e. within the shaded region).
-All of the new parameters that appear in the ET response parameter rankings govern plant acclimation to temperature. 
-These parameters, while not especially influential on present-day ET, are among the most important in governing the trend in ET going forward.
+Besides \textit{kmax}, all of the parameters that appear in the top eight parameters for ET response govern plant acclimation to temperature. 
+These parameters, while not especially influential on present-day ET, would be among the most influential in governing the trend in ET going forward, illustrating the potential limitations of calibration to ambient conditions.
 
@@
 The most influential parameters globally include parameters from each of these three biomes.
-We have similar plots delineated by plant functional type, and for many other output variables in our extended diagnostics set (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
+Because the number of potential variables, parameters, and geographical ranges of interest to the wider CLM community is larger than we can document here,  we provide a tool to explore an extended diagnostics set which summarises the $>$2TB of output data and contains 1720 figures (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
 
@@
 
-The volume of data generated by a large PPE can be overwhelming - our full dataset exceeds 2TB.
-Our diagnostics package, which operates on only a subset of output variables, generated 1720 figures (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
-In many cases we have found it easier to utilize interactive plotting utilities to toggle through the various parameters, output variables, and forcing scenarios (Figure \ref{fig:panel}).
+In many cases we found it easier to utilize interactive plotting utilities to toggle through the various parameters, output variables, and forcing scenarios (Figure \ref{fig:panel}).
 Interactive widgets are relatively easy to generate within a data analysis session, but more difficult to host online, hence the screenshot.
@@
 
-In this project, we have identified and perturbed 193 CLM parameters to create a large one-at-a-time PPE.
+In this project, we identified and perturbed 193 CLM parameters to create a large one-at-a-time PPE.
 This ensemble is useful for understanding parametric controls on CLM processes and has already been leveraged as a starting point for parameter optimization efforts.
-We made significant investments in software infrastructure to enable this experiment and reduce the burden of generating future PPEs.
+The software and analysis tools generated to enable this experiment will reduce the burden of generating future PPEs.
 
@@
 For this experiment we opted for a one-at-a-time perturbation strategy, testing a minimum and maximum value for each of 193 parameters.
-We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:violins}).
+We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:ranges}).
 That being said, the majority of parameter perturbations had very small effects, such that a majority of simulations were clustered around the default simulation.
-In the case of canopy evaporation, more than 90\% of the ensemble variance could be explained by the ten most influential parameters (Figure \ref{fig:variance}).
-As such, we would expect that the dimensionality of the effective parameter space for any given process would be significantly smaller than implied by the 193 parameters.
+In the case of canopy evaporation, more than 95\% of the ensemble variance could be explained by the ten most influential parameters (Figure \ref{fig:variance}).
+This indicates that there may be very tractable parameter estimation sub-problems if global calibration of 193 parameters proves challenging.
 
@@
 We have published a large set of ensemble diagnostic plots online, which serves as a valuable enhancement to our model technical documentation.
-Now in addition to seeing the definition of a given parameter, and the relevant equations, a model user can see the magnitude and spatial patterns of its effects (e.g.
+Now in addition to seeing the definition of a given parameter, and the relevant equations, a model user can easily investigate the magnitude and spatial patterns of its effects (e.g.
 Figure \ref{fig:panel}).
@@
 \section*{Open Research Section}
+
+The model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
+
+ (CTSM component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP})
+
 This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/supp/FCEV_global_mean.png}
+\caption{The top ten parameters controlling global annual canopy evaporation (CLM variable: FCEV) across the six forcing scenarios.
+Note that this plot is one of many from our more comprehensive diagnostics set \url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT/}.}
+\label{supp:fcev}
+\end{figure}
+
 \end{document}

### 2023-08-02 ms/ms.tex
add refs

@@
 Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants previously considered as `internal' are hard-coded and not amenable to systematic perturbation.
-For the purposes of this activity, we identified what we consider the full set of 188 CLM-BGC parameters, and enabled their modification via the model input files.
+For the purposes of this activity, we identified what we consider a very broad set of 193 CLM-BGC parameters, and enabled their modification via the model input files.
+We opted not to perturb crop parameters within this ensemble, as our focus is on unmanaged vegetation.
 
@@
 In this project, we identified and perturbed 193 CLM parameters to create a large one-at-a-time PPE.
-This ensemble is useful for understanding parametric controls on CLM processes and has already been leveraged as a starting point for parameter optimization efforts.
-The software and analysis tools generated to enable this experiment will reduce the burden of generating future PPEs.
+This ensemble is useful for understanding parametric controls on CLM processes.
+The software and analysis tools generated to enable this experiment will greatly reduce the burden of generating future PPEs and parameter sensitivity experiments.
 
@@
 In such cases, we identified hard-coded values, established an appropriate parameter name, and extracted that parameter to the CLM parameter file for easier manipulation.
-In the end, we perturbed 193 parameters across a variety of CLM processes (Figure \ref{fig:params}).
+In this experiment we perturbed 193 parameters across a variety of CLM processes (Figure \ref{fig:params}).
+This is not the full set of CLM parameters, as some processes were not included, such as crops.
+In carrying out this experiment, we have likewise realized that the epistemology of climate model `parameters' is highly nuanced, and that defining the parameters within a given model structure can be somewhat subjective, such that it would be very difficult to collate a comprehensive or definitive set of parameters for a model like CLM.
+
 The second challenge involved defining a perturbation range for each parameter.
@@
 In some cases, literature values were explicitly used (e.g.
-medlynslope), but the most common range was $\pm20\%$.
+medlynslope, \citeA{lin2015}), but the most common range was $\pm20\%$.
+It is exceedingly difficult to set parameter ranges that sample comparable probability density, even in a univariate experiment, like this one.
+The number of parameters is very large, many of which lack sufficient empirical backing for robust range evaluation.
+Even with parameters that have an empirical basis, they may behave differently at the coarse climate modeling spatial resolution, complicating the process of uncertainty quantification.
+As such, we cannot claim that the parameters are equivalently sampled, whereby parameter effects and parameter rankings will be subject to sampling asymmetries.
 
@@
 We were able to speed up the model approximately 500x by strategically reducing the number of model grid cells (Figure \ref{fig:sg}) and by leveraging a linearized spin-up solver (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
-Designing faster model configurations and more efficient sampling strategies will greatly aid in model calibration and uncertainty quantification.
+As long as computational resources remain constrained, designing faster model configurations and more efficient sampling strategies will be important for effective model calibration and uncertainty quantification.
 
-For this experiment we opted for a one-at-a-time perturbation strategy, testing a minimum and maximum value for each of 193 parameters.
+For this experiment we opted for a one-at-a-time perturbation strategy, testing a minimum and maximum value for each parameter.
 We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:ranges}).
@@
 In the case of canopy evaporation, more than 95\% of the ensemble variance could be explained by the ten most influential parameters (Figure \ref{fig:variance}).
-This indicates that there may be very tractable parameter estimation sub-problems if global calibration of 193 parameters proves challenging.
+This indicates that there may be very tractable parameter estimation sub-problems if global calibration of 193 parameters proves unattainable.
 
@@
 The advantage of our dataset is that we can diagnose parameter effects without the uncertainty contributed by an emulator or regression model.
-As such, it is easy to diagnose which parameters are most influential on a given process, which has already been leveraged for parameter filtering.
+As such, it is easy to diagnose which parameters are most influential on a given process.
 We have published a large set of ensemble diagnostic plots online, which serves as a valuable enhancement to our model technical documentation.
@@
 
-We have already begun a follow-on experiment that perturbs a subset of parameters with the Latin hypercube sampling strategy to work towards calibrating leaf area index.
+Several spin-off projects have leveraged the work presented here, including some that have already reached publication status \cite{cheng2023,yan2023a,yan2023b}.
+These projects utilized the output from this ensemble to filter for parameters that influence the relevant study domains and used the parameter ranges collated in Section zqz.
+Our project has accelerated parameter exploration work within our collaborator network by providing:
+
+\begin{itemize}
+\item Parameter ranges for nearly 200 parameters
+\item Machine learning smart low resolution unstructured grid (sparsegrid)
+\item Accelerated spinup procedure
+\item Ensemble generation scripting toolchain
+\item Parameter sensitivity diagnostics for nearly 200 parameters across 6 forcing scenarios
+\end{itemize}
+
+We have likewise begun a follow-on experiment that perturbs a subset of parameters with the Latin hypercube sampling strategy to work towards calibrating leaf area index.
 By investing in the infrastructure underlying this first experiment, all of our subsequent experiments have required much less time and effort.

### 2023-08-02 ms/makebib.sh

@@ -1,9 +1,10 @@
 grep citation ms.aux | sed 's/citation//g' | tr -d '\\{}' | sort | uniq > cites.txt
 
-d='/Users/kennedy/projects/references/refs/'
+d='/Users/djk2120/projects/references/refs/'
 :> refs.bib
 while read c; do
     file=$d$c".bib"
+    echo $file
     if [ -f $file ]; then
 	cat $file >> refs.bib
 	echo -en '\n' >> refs.bib

### 2023-08-09 ms/ms.tex
dave comments

@@
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
-Effectively synthesizing the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns can help to parameterize increasingly comprehensive land surface modelling platforms
+Effectively synthesizing the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns can help to parameterize increasingly comprehensive land modeling platforms.
 
@@
 To allow for more tractable experimental design each model is typically allowed only a single parameterization of many plausible parameter sets.
-Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land surface trajectories.
+Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land state trajectories.
 
-The process of parameter selection for land surface models is challenging on account of the intrinsic  complexity of the heterogenous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryology, biogeochemistry, physiology, land management, micrometeorology, etc.).
+The process of parameter selection for land  models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryology, biogeochemistry, physiology, land management, micrometeorology, etc.).
 The number of model parameters required to represent all the relevant processes is large, and model speeds are relatively slow, presenting a major barrier to robust objective calibration, model interpretation and development \cite{fisher2020,dagon2020}.
-Thorough assessment of the parameter sensitivity of land surfaces models in a manner that is repeatable, open and integrated with ongoing code development is desirable but very challenging \cite{hourdin2017,balaji2022}.
+Thorough assessment of the parameter sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}.
 It typically requires extensive computational time,  software engineering, and domain-specific expert scientific knowledge.
 
-Several promising avenues for model calibration are in development (lavendar, musa, CES), many of which require the construction of a large parameter perturbation ensembles (PPE) \cite{qian2018}. 
+Several promising avenues for model calibration are in development (lavendar, musa, CES), many of which require the construction of large parameter perturbation ensembles (PPEs) \cite{qian2018}. 
 PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
@@
 \citeA{collier2018}.
-Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.
+Our goal in this project is to systematize the parameter perturbation process within the CLM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.
 In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
 In this paper we present that dataset, describe how it was produced, and survey some potential applications.
-The dataset has already demonstrated great utility for diagnosing parameter effects (cheng,yan), while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CLM.
+The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CLM.
 
@@
 \label{sect:md}
-The Community Terrestrial Systems Model (CTSM) is developed by the CESM Land Model Working Group and maintained at the National Center for Atmospheric Research (NCAR).
-This experiment utilizes the Community Land Model configuration of CTSM, version 5.1 (CLM5.1).
+The is developed by the CESM Land Model Working Group and maintained at the National Center for Atmospheric Research (NCAR).
+This experiment utilizes the Community Land Model configuration version 5.1 (CLM5.1) of the Community Terrestrial Systems Model.
 The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.
@@
 Relative to CLM5.0, version 5.1 includes minor bug fixes, parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}.
-The PPE experiment required additional code modifications to systematically vary the full suite of model parameters (many of which were previously `hard-coded').
+The PPE experiment required additional code modifications to enable systematic variation of the full suite of model parameters (many of which were previously `hard-coded').
 We utilized the biogeochemical cycling version of CLM, which includes active carbon and nitrogen cycles.
@@
 This new module utilizes a linearized simplification of CLM's biogeochemistry to significantly reduce spin-up time.
-Our spin-up protocol featured 20 years in accelerated decomposition mode, followed by 80 years of Matrix-CN, followed by 40 years of `normal' mode, cycling over a ten-year forcing dataset (described below).
+Our spin-up protocol featured 20 years in accelerated decomposition mode (see \citeA{lawrence2019} for details), followed by 80 years of Matrix-CN, followed by 40 years of `normal' mode, cycling over a ten-year forcing dataset (described below).
 This protocol was designed to achieve sufficiently equilibrated model states, while minimizing computational time.
 This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter depth).
-Certain inferences about deep soil carbon are therefore subject to uncertainty due to spin-up concerns.
+Certain inferences about deep soil carbon from this ensemble are therefore subject to uncertainty due to spin-up concerns.
 
@@
 Another control on model cost is resolution.
-Most CLM simulations utilize nominal 1$^\circ$ resolution, which requires over 20,000 land grid cells.
+Most CLM simulations utilize nominal 1$^\circ$ resolution, which equates to about 20,000 land grid cells.
 In order to manage computational cost, parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}.
@@
 Multivariate spatio-temporal clustering (MVSC) has been utilized to extract patterns of climatological significance from climate model output \cite{hoffman2005} and applied to design a representativeness-based sampling network \cite{hoffman2013}.
-Instead of lowering resolution by coarsening a rectilinear grid, we here used MVSC to strategically remove redundant grid cells, leaving only 400 grid cells that efficiently sample important model dynamics.
+Instead of lowering resolution by coarsening a rectilinear grid, we here used MVSC to strategically remove effectively redundant grid cells, leaving only 400 grid cells that efficiently sample important model dynamics across the globe.
 
@@
 For the purposes of this activity, we identified what we consider a very broad set of 193 CLM-BGC parameters, and enabled their modification via the model input files.
-We opted not to perturb crop parameters within this ensemble, as our focus is on unmanaged vegetation.
+We opted not to perturb crop parameters within this ensemble, as our initial focus is on unmanaged land.
 
-Given the very large parameter space, to conduct an initial analysis of the response of the model to parametric uncertainty, we decided to vary each parameter independently, exploring low and high values.
+Given the large parameter space, to conduct an initial analysis of the response of the model to parametric uncertainty, we decided to vary each parameter independently, exploring the impact of low and high values.
 To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter.
-In some cases literature values were directly utilized, but in the majority of cases, expert judgment was used.
+In some cases literature values were directly utilized, but in the many cases, expert judgment was used.
 The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.
-In some cases parameters could not be perturbed independently, because they feature inherent co-dependence.
-In the case of nitrogen fixation costs, we opted to perturb the parameters independently, but also as a group (`KCN'), to enforce nitrogen limitation in lieu of switching from one uptake pathway to another.
+In some cases parameters could not or should not be perturbed independently, because they feature inherent co-dependence.
+In the case of nitrogen fixation costs, we opted to perturb the parameters independently, but also as a group (`KCN'), to enforce a change in overall nitrogen limitation in lieu of switching from one uptake pathway to another.
 We likewise opted to perturb soil hydraulic parameters directly (e.g.
@@
 Each simulation ran for 150 years, with the first 140 for spin-up (as described above in Section \ref{sect:mcn}), followed by a 10-year period for analysis.
-We opted for six different forcing scenarios to understand the intersection of parameter effects and the different forcing agents associated with climate change (Table \ref{tab:exps}).
+We opted for six different forcing scenarios to understand the intersection of parameter effects with different forcing agents associated with climate change (Table \ref{tab:exps}).
 The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and is the default forcing data for CLM5 \cite{lawrence2019}.
@@
 Our future climate scenario utilizes the SSP3-7.0 forcing in order to allow comparison with the existing CESM2 large ensemble.
+Our software infrastructure has greatly reduced the effort and computational cost associated with ensemble generation, such that it would be easy to add additional ensembles, to explore other emission scenarios or climate regimes (e.g.
+Last Glacial Maximum).
 
@@
 
+\subsection{Analyses}
+
+Biome analysis
+
+\section{Results}
+
+A total of 193 parameters were targeted for sensitivity testing across a wide variety of CLM subcomponents (Figure \ref{fig:params}).
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/bar.pdf}
+\caption{193 parameters were perturbed across the various domains of the land model.}
+\label{fig:params}
+\end{figure}
+
+Given the challenging size of the CLM5-BGC parameter space, and the need to probe and document parametric uncertainties within operational land surface models, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
+Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times cheaper than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
+There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
+Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=30pc]{../figs/sims.pdf}
+\caption{The approximate number of simulations afforded by 1 million core-hours on the Cheyenne supercomputer for a range of CLM configurations.
+Configurations are labeled according to spin-up procedure (CN-matrix or the standard 1500-year spinup) and horizontal resolution (`sg' signifies sparsegrid).
+The inset map shows the locations of the 400 sparse grid cells.
+See Section \ref{methods} for spin-up and sparsegrid details.}
+\label{fig:sims}
+\end{figure}
+
+Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity.
+Given more clusters, the sparsegrid provides a better approximation of output from the full grid (Figure \ref{fig:sg}).
+Accurate reconstruction of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Supp Figure zqz).
+We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
+\begin{figure}[h]
+\centering
+\includegraphics[width=25pc]{../figs/sparsegrid_gpp.pdf}
+\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last forty years of a transient CLM5.1 simulation.
+We opted for 400 clusters to balance computational cost against representativeness.}
+\label{fig:sg}
+\end{figure}
+
+Overall, we found substantial impacts of parametric uncertainty on model behaviour, in some cases exceeding the magnitude of climate scenario effects (Figure \ref{fig:ranges}).
+One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death.
+A perturbation of $\pm$20\% likely exceeds the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils.
+Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, which may belie unforeseen tipping points, brittle parameterizations,  and/or bugs.
+
+A small fraction of parameters tend to explain a large amount of the ensemble variance (Figure \ref{fig:variance}) for any given variable or metric.
+For example the same four parameters explain up to 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig \ref{supp:fcev}).
+Most of the parameter perturbations had a relatively small impact on any given output variable.
+As such, the effective parameter space dimensionality for any given CLM process is likely much smaller than would be expected from having 193 independent parameters, reflecting the fact that land surface models like CLM comprise representations of numerous interconnected domains.
+While the principle of land models is that these are all interlinked via complex feedback processes, many parameters nonetheless have impacts that are mostly limited to their own domain.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/ranges.png}
+\caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across six forcing scenarios.
+Circles mark the CLM5.1 default simulation, and the bars span the ensemble ranges.}
+\label{fig:ranges}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/variance.pdf}
+\caption{Cumulative fraction of variance explained by the most influential parameters on gross primary production and canopy evaporation.
+For example approximately 96\% of canopy evaporation variance is explained by the first ten parameters regardless of forcing, whereas between 74 and 92\% of GPP variance is explained by the first ten parameters.
+Only the top 50 parameters are shown (of 193) to improve the visualization.
+Note that parameter rankings differ by variable.}
+\label{fig:variance}
+\end{figure}
+
+We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps}.
+This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control responses to individual forcing variables (CO2, nitrogen deposition) and climate  In the case of evapotranspiration (ET), for example, the parameters that control the increase in ET due to warming tend to be different from the parameters that control present-day ET (Figure \ref{fig:et}).
+ET response can differ significantly from the default case (i.e. stray from the orange line), even when ET is near the default under present-day conditions (i.e. within the shaded region).
+Besides \textit{kmax}, all of the parameters that appear in the top eight parameters for ET response govern plant acclimation to temperature. 
+These parameters, while not especially influential on present-day ET, would be among the most influential in governing the trend in ET going forward, illustrating the potential limitations of calibration to ambient conditions.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/ET_response.png}
+\caption{(a) Global annual average evapotranspiration (ET) in the warm forcing scenario vs.
+our control experiment.
+The shading spans the default ET in the control experiment, $\pm$1 W/m2.
+The response of ET to future climate with default parameters is +8.8\%.
+The orange line shows the expected future ET should the model conform to the default ET response.
+(b) The top 8 parameters governing global ET in the control experiment.
+(c) The top 8 parameters governing ET response to future climate, i.e. the ratio of future ET to control ET.
+Parameters governing the change in ET expected in the future tend to differ from the parameters controlling present-day ET.}
+\label{fig:et}
+\end{figure}
+
+The PPE is likewise useful for determining which parameters influence which model processes and where.
+For example the parameters controlling leaf area index vary significantly by biome (Figure \ref{fig:lai}).
+Plant hydraulics parameters were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
+The most influential parameters globally include parameters that were important in each of these three biomes.
+Because the number of potential variables, parameters, and geographical ranges of interest to the wider CLM community is larger than we can document here,  we provide a tool that can be used to explore an extended diagnostics set which summarises the $>$2TB of output data and contains 1720 figures (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/lai_biome.pdf}
+\caption{The eight most influential parameters on leaf area index within the control ensemble, globally and within three biomes.}
+\label{fig:lai}
+\end{figure}
+
+In many cases we found it easier to utilize interactive plotting utilities to toggle through the various parameters, output variables, and forcing scenarios (Figure \ref{fig:panel}).
+Interactive widgets are relatively easy to generate within a data analysis session, but more difficult to host online, hence the screenshot.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/deltmap_panel.pdf}
+\caption{Screenshot of interactive diagnostic for exploring maps of parameter effects.
+In this case, the effect of medlynslope on runoff within the control ensemble.
+Increasing medlynslope tends to reduce runoff, but only in regions with sufficient vegetation activity.}
+\label{fig:panel}
+\end{figure}
+
+\section{Discussion}
+
+In this project, we identified and perturbed 193 CLM parameters to create a large one-at-a-time PPE.
+This ensemble is useful for understanding parametric controls on CLM processes.
+The software and analysis tools generated to enable and utilize this experiment will greatly reduce the burden of generating future PPEs and parameter sensitivity experiments.
+
+There were several barriers to perturbing the full set of CLM5-BGC parameters.
+Firstly, many parameters had not been officially identified as such.
+In such cases, we identified hard-coded values, established an appropriate parameter name, and extracted that parameter to the CLM parameter file for easier manipulation.
+In this experiment we perturbed 193 parameters across a variety of CLM processes (Figure \ref{fig:params}).
+This is not the full set of CLM parameters, as some processes were not included, such as crops.
+In carrying out this experiment, we have likewise realized that the epistemology of climate model `parameters' is highly nuanced, and that defining the parameters within a given model structure can be somewhat subjective, such that it would be difficult to collate a comprehensive or definitive set of parameters for a model like CLM.
+
+The second challenge involved defining a perturbation range for each parameter.
+We solicited expert judgment to set a minimum and maximum reasonable value for each parameter.
+In some cases, literature values were explicitly used (e.g.
+medlynslope, \citeA{lin2015}), but the most common range was $\pm20\%$.
+It is exceedingly difficult to set parameter ranges that sample comparable probability density, even in a univariate experiment, like this one.
+The number of parameters is large, with many lacking sufficient empirical backing for robust range evaluation.
+Even with parameters that have an empirical basis, they may behave differently at the coarse climate modeling spatial resolution (zqz), complicating the process of uncertainty quantification.
+As such, we cannot claim that the parameters are equivalently sampled, whereby parameter effects and parameter rankings could be subject to sampling asymmetries.
+
+The third challenge involved managing computational cost.
+Quantifying parameter effects is a necessary prerequisite to automated calibration and uncertainty quantification.
+Because the parameter space of CLM is quite large and the model response potentially non-linear, large ensembles are required to adequately resolve response surfaces.
+With standard CLM configurations, such ensembles would far exceed our computing allocation.
+As such, the computational cost of inferring parameter effects is a major constraint.
+We were able to reduce ensemble generation computational cost 500x by strategically reducing the number of model grid cells (Figure \ref{fig:sg}) and by leveraging a linearized spin-up solver (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
+As long as computational resources remain constrained, designing faster model configurations and more efficient sampling strategies will be important for effective model calibration and uncertainty quantification.
+
+For this experiment we opted for a one-at-a-time perturbation strategy, testing a minimum and maximum value for each parameter.
+We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:ranges}).
+That being said, the majority of parameter perturbations had small effects for any specific model variable or metric, such that a majority of simulations were clustered around the default simulation.
+In the case of canopy evaporation, for example, more than 95\% of the ensemble variance could be explained by the ten most influential parameters (Figure \ref{fig:variance}).
+This indicates that there may be tractable parameter estimation sub-problems if global calibration of 193 parameters proves unattainable.
+
+We opted for six forcing scenarios to identify parameter effects not just in present-day but in response to pre-industrial or future climate conditions.
+The parameters that were most influential under present-day conditions did not necessarily match the set of parameters controlling the response to future forcing (Figure \ref{fig:et}).
+We found that many acclimation parameters, which were not as important for determining present-day evapotranspiration (ET), were among the most influential on the response of ET to future climate.
+Similarly, the parameters controlling leaf area index varied significantly depending on biome (Figure \ref{fig:lai}).
+
+A one-at-a-time PPE cannot capture parameter interactions, and our min/max sampling protocol precludes diagnosing non-linearities.
+As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty.
+A primary utility of our dataset is that we can diagnose parameter effects without the uncertainty contributed by an emulator or a regression model.
+As such, it is easy to diagnose which parameters are most influential on a given process.
+We have published a large set of ensemble diagnostic plots online, which serves as a valuable enhancement to our model technical documentation.
+Now, in addition to seeing the definition of a given parameter, and the relevant equations, a model user can easily investigate the magnitude and spatial patterns of its effects (e.g.
+Figure \ref{fig:panel}).
+
+Several spin-off projects have leveraged the work presented here, including some that have already reached publication status \cite{cheng2023,yan2023a,yan2023b}.
+These projects utilized the output from this ensemble to filter for parameters that influence the relevant study domains and used the parameter ranges collated in Section zqz.
+Our project has accelerated parameter exploration work within our collaborator network by providing:
+
+\begin{itemize}
+\item Parameter ranges for nearly 200 parameters
+\item Machine learning smart low resolution unstructured grid (sparsegrid)
+\item Accelerated spinup procedure
+\item Ensemble generation scripting toolchain
+\item Parameter sensitivity diagnostics for nearly 200 parameters across 6 forcing scenarios
+\end{itemize}
+
+We have likewise begun a follow-on activity that perturbs a subset of important parameters with the Latin hypercube sampling strategy to work towards routine model calibration.
+Our first foray into exploring the potential for global calibration is with leaf area index due to controls that this variable has on biogeophysical and biogeochemical fluxes.
+By investing in the infrastructure that we introduce here, all of our subsequent paramter perturbation experiments have required much less time and effort.
+We expect to continue to extend our efforts in this domain towards model calibration, as well as repeat the foundational one-at-a-time experiments with subsequent model releases.
+
+\section{Conclusion}
+Do we want / need an explicit conclusion?
+
+\section*{Open Research Section}
+
+The model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
+
+ (CTSM component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP})
+
+This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
+Data cannot be listed as ''Available from authors'' or stored solely in supporting information.
+Citations to archived data should be included in your reference list.
+Wiley will publish it as a separate section on the paper’s page.
+Examples and complete information are here:
+https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
+
+\acknowledgments
+Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
+
+\bibliography{refs}
+
+\appendix
+\section{Supplementary Figures}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=42pc]{../figs/supp/ilamb.pdf}
+\caption{ILAMB 2.5 overall scores comparing remapped sparsegrid output to the full grid 2$^{\circ}$ model output.
+Column headings indicate the number of clusters, and whether annual means or annual means and standard deviations were used as input to the clustering algorithm.
+We should try to remove the gray bits (zqz).
+Full, interactive results are available at \url{www.ilamb.org/PPE/CLM/2021-02}.
+See main text Section \ref{sect:sg} for clustering details.}
+\label{supp:ilamb}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/supp/FCEV_global_mean.png}
+\caption{The top ten parameters controlling global annual canopy evaporation (CLM variable: FCEV) across the six forcing scenarios.
+Note that this plot is one of many from our more comprehensive diagnostics set \url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT/}.}
+\label{supp:fcev}
+\end{figure}
+
+Will be deleted:
 \subsection{Parameters} 
@@
 
-\subsection{Analyses}
-
-Biome analysis
-
-\section{Results}
-
-A total of 193 parameters were targeted for sensitivity testing across a wide variety of CLM subcomponents (Figure \ref{fig:params}).
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\textwidth]{../figs/bar.pdf}
-\caption{193 parameters were perturbed across the various domains of the land model.}
-\label{fig:params}
-\end{figure}
-
-Given the challenging size of the CLM5-BGC parameter space, and the need to probe and document parametric uncertainties within operational land surface models, a  major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
-Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times faster than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
-There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
-Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=30pc]{../figs/sims.pdf}
-\caption{The approximate number of simulations afforded by 1 million core-hours for a range of CLM configurations.
-Configurations are labeled according to spin-up procedure (CN-matrix or the standard 1500-year spinup) and horizontal resolution (`sg' signifies sparsegrid).
-The inset map shows the locations of the 400 sparse grid cells.
-See Section \ref{methods} for spin-up and sparsegrid details.}
-\label{fig:sims}
-\end{figure}
-
-Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity.
-Generally the more clusters, the better the approximation of output from the full grid that can be achieved with the sparsegrid (Figure \ref{fig:sg}).
-Accurate reconstruction of of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Supp Figure zqz).
-We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
-\begin{figure}[h]
-\centering
-\includegraphics[width=25pc]{../figs/sparsegrid_gpp.pdf}
-\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last forty years of a transient CLM5.1 simulation.
-We opted for 400 clusters to balance computational cost against representativeness.}
-\label{fig:sg}
-\end{figure}
-
-Overall, we found substantial impacts of parametric uncertainty on model behaviour, in some cases exceeding the magnitude of climate scenario effects (Figure \ref{fig:ranges}).
-One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death.
-A perturbation of $\pm$20\% likely exceeds the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils.
-Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, which may belie unforeseen tipping points, brittle parameterizations,  and/or bugs.
-
-A small fraction of parameters tend to explain a large amount of the ensemble variance (Figure \ref{fig:variance}).
-For example the same four parameters explain up to 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig \ref{supp:fcev}).
-Conversely, most of the parameter perturbations had a relatively small impact on any given output variable.
-As such, the effective parameter space dimensionality for any given CLM process is expected to be much smaller than would be expected from having 193 independent parameters, reflecting the fact that land surface models like CLM comprise representations of numerous interconnected domains.
-While the principle of LSMs is that these are all interlinked via complex feedback processes, many parameters nonetheless have impacts that are mostly limited to their own domain.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\textwidth]{../figs/ranges.png}
-\caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across six forcing scenarios.
-Circles mark the CLM5.1 default simulation, and the bars span the ensemble ranges.}
-\label{fig:ranges}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\textwidth]{../figs/variance.pdf}
-\caption{Cumulative fraction of variance explained by the most influential parameters on gross primary production and canopy evaporation.
-For example approximately 96\% of canopy evaporation variance is explained by the first ten parameters regardless of forcing, whereas between 74 and 92\% of GPP variance is explained by the first ten parameters.
-Only the top 50 parameters are shown (of 193) to improve the visualization.
-Note that parameter rankings differ by variable.}
-\label{fig:variance}
-\end{figure}
-
-We repeated the full set of parameter perturbations across the six forcing scenarios in  see Table \ref{tab:exps}.
-This ensures that we can identify important parameters not just under present-day conditions, but also those that control responses to individual forcing variables (CO2, nitrogen deposition) and climate  In the case of evaporation (ET), for example, the parameters that control the increase in ET due to warming tend to be different from the parameters that control present-day ET (Figure \ref{fig:et}).
-ET response can differ significantly from the default case (i.e. stray from the orange line), even when ET is near the default under present-day conditions (i.e. within the shaded region).
-Besides \textit{kmax}, all of the parameters that appear in the top eight parameters for ET response govern plant acclimation to temperature. 
-These parameters, while not especially influential on present-day ET, would be among the most influential in governing the trend in ET going forward, illustrating the potential limitations of calibration to ambient conditions.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\textwidth]{../figs/ET_response.png}
-\caption{(a) Global annual average evapotranspiration (ET) in the warm forcing scenario vs.
-our control experiment.
-The shading spans the default ET in the control experiment, $\pm$1 W/m2.
-The response of ET to future climate with default parameters is +8.8\%.
-The orange line shows the expected future ET should the model conform to the default ET response.
-(b) The top 8 parameters governing global ET in the control experiment.
-(c) The top 8 parameters governing ET response to future climate, i.e. the ratio of future ET to control ET.
-Parameters governing the change in ET expected in the future tend to differ from the parameters controlling present-day ET.}
-\label{fig:et}
-\end{figure}
-
-The PPE is likewise useful for determining which parameters influence which model processes and where.
-For example the parameters controlling leaf area index vary significantly by biome (Figure \ref{fig:lai}).
-Plant hydraulics parameters were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
-The most influential parameters globally include parameters from each of these three biomes.
-Because the number of potential variables, parameters, and geographical ranges of interest to the wider CLM community is larger than we can document here,  we provide a tool to explore an extended diagnostics set which summarises the $>$2TB of output data and contains 1720 figures (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\textwidth]{../figs/lai_biome.pdf}
-\caption{The eight most influential parameters on leaf area index within the control ensemble, globally and within three biomes.}
-\label{fig:lai}
-\end{figure}
-
-In many cases we found it easier to utilize interactive plotting utilities to toggle through the various parameters, output variables, and forcing scenarios (Figure \ref{fig:panel}).
-Interactive widgets are relatively easy to generate within a data analysis session, but more difficult to host online, hence the screenshot.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\textwidth]{../figs/deltmap_panel.pdf}
-\caption{Screenshot of interactive diagnostic for exploring maps of parameter effects.
-In this case, the effect of medlynslope on runoff within the control ensemble.
-Increasing medlynslope tends to reduce runoff, but only in regions with sufficient vegetation activity.}
-\label{fig:panel}
-\end{figure}
-
-\section{Discussion}
-
-In this project, we identified and perturbed 193 CLM parameters to create a large one-at-a-time PPE.
-This ensemble is useful for understanding parametric controls on CLM processes.
-The software and analysis tools generated to enable this experiment will greatly reduce the burden of generating future PPEs and parameter sensitivity experiments.
-
-There were several barriers to perturbing the full set of CLM5-BGC parameters.
-Firstly, many parameters had not been officially identified as such.
-In such cases, we identified hard-coded values, established an appropriate parameter name, and extracted that parameter to the CLM parameter file for easier manipulation.
-In this experiment we perturbed 193 parameters across a variety of CLM processes (Figure \ref{fig:params}).
-This is not the full set of CLM parameters, as some processes were not included, such as crops.
-In carrying out this experiment, we have likewise realized that the epistemology of climate model `parameters' is highly nuanced, and that defining the parameters within a given model structure can be somewhat subjective, such that it would be very difficult to collate a comprehensive or definitive set of parameters for a model like CLM.
-
-The second challenge involved defining a perturbation range for each parameter.
-We solicited expert judgment to set a minimum and maximum reasonable value for each parameter.
-In some cases, literature values were explicitly used (e.g.
-medlynslope, \citeA{lin2015}), but the most common range was $\pm20\%$.
-It is exceedingly difficult to set parameter ranges that sample comparable probability density, even in a univariate experiment, like this one.
-The number of parameters is very large, many of which lack sufficient empirical backing for robust range evaluation.
-Even with parameters that have an empirical basis, they may behave differently at the coarse climate modeling spatial resolution, complicating the process of uncertainty quantification.
-As such, we cannot claim that the parameters are equivalently sampled, whereby parameter effects and parameter rankings will be subject to sampling asymmetries.
-
-The third challenge involved managing computational cost.
-Quantifying parameter effects is a necessary prerequisite to automated calibration and uncertainty quantification.
-Because the parameter space of CLM is quite large and the model response potentially non-linear, large ensembles are required to adequately resolve response surfaces.
-With standard CLM configurations, such ensembles would far exceed our computing allocation.
-As such, the computational cost of inferring parameter effects is a major constraint.
-We were able to speed up the model approximately 500x by strategically reducing the number of model grid cells (Figure \ref{fig:sg}) and by leveraging a linearized spin-up solver (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
-As long as computational resources remain constrained, designing faster model configurations and more efficient sampling strategies will be important for effective model calibration and uncertainty quantification.
-
-For this experiment we opted for a one-at-a-time perturbation strategy, testing a minimum and maximum value for each parameter.
-We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:ranges}).
-That being said, the majority of parameter perturbations had very small effects, such that a majority of simulations were clustered around the default simulation.
-In the case of canopy evaporation, more than 95\% of the ensemble variance could be explained by the ten most influential parameters (Figure \ref{fig:variance}).
-This indicates that there may be very tractable parameter estimation sub-problems if global calibration of 193 parameters proves unattainable.
-
-We opted for six forcing scenarios to identify parameter effects not just in present-day but in response to pre-industrial or future climate conditions.
-The parameters that were most influential under present-day conditions did not necessarily match the set of parameters controlling the response to future forcing (Figure \ref{fig:et}).
-We found that many acclimation parameters, which were not as important for determining present-day evapotranspiration (ET), were among the most influential on the response of ET to future climate.
-Similarly, the parameters controlling leaf area index varied significantly depending on biome (Figure \ref{fig:lai}).
-
-A one-at-a-time PPE cannot capture parameter interactions, and our min/max sampling protocol precludes diagnosing non-linearities.
-As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty.
-The advantage of our dataset is that we can diagnose parameter effects without the uncertainty contributed by an emulator or regression model.
-As such, it is easy to diagnose which parameters are most influential on a given process.
-We have published a large set of ensemble diagnostic plots online, which serves as a valuable enhancement to our model technical documentation.
-Now in addition to seeing the definition of a given parameter, and the relevant equations, a model user can easily investigate the magnitude and spatial patterns of its effects (e.g.
-Figure \ref{fig:panel}).
-
-Several spin-off projects have leveraged the work presented here, including some that have already reached publication status \cite{cheng2023,yan2023a,yan2023b}.
-These projects utilized the output from this ensemble to filter for parameters that influence the relevant study domains and used the parameter ranges collated in Section zqz.
-Our project has accelerated parameter exploration work within our collaborator network by providing:
-
-\begin{itemize}
-\item Parameter ranges for nearly 200 parameters
-\item Machine learning smart low resolution unstructured grid (sparsegrid)
-\item Accelerated spinup procedure
-\item Ensemble generation scripting toolchain
-\item Parameter sensitivity diagnostics for nearly 200 parameters across 6 forcing scenarios
-\end{itemize}
-
-We have likewise begun a follow-on experiment that perturbs a subset of parameters with the Latin hypercube sampling strategy to work towards calibrating leaf area index.
-By investing in the infrastructure underlying this first experiment, all of our subsequent experiments have required much less time and effort.
-We expect to continue to extend our efforts in this domain, as well as repeat these types of experiments with subsequent model releases.
-
-\section{Conclusion}
-\section*{Open Research Section}
-
-The model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
-
- (CTSM component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP})
-
-This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
-Data cannot be listed as ''Available from authors'' or stored solely in supporting information.
-Citations to archived data should be included in your reference list.
-Wiley will publish it as a separate section on the paper’s page.
-Examples and complete information are here:
-https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
-
-\acknowledgments
-Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
-
-\bibliography{refs}
-
-\appendix
-\section{Supplementary Figures}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=42pc]{../figs/supp/ilamb.pdf}
-\caption{ILAMB 2.5 overall scores comparing remapped sparsegrid output to the full grid 2$^{\circ}$ model output.
-Column headings indicate the number of clusters, and whether annual means or annual means and standard deviations were used as input to the clustering algorithm.
-We should try to remove the gray bits (zqz).
-Full, interactive results are available at \url{www.ilamb.org/PPE/CLM/2021-02}.
-See main text Section \ref{sect:sg} for clustering details.}
-\label{supp:ilamb}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\textwidth]{../figs/supp/FCEV_global_mean.png}
-\caption{The top ten parameters controlling global annual canopy evaporation (CLM variable: FCEV) across the six forcing scenarios.
-Note that this plot is one of many from our more comprehensive diagnostics set \url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT/}.}
-\label{supp:fcev}
-\end{figure}
-
 \end{document}

### 2023-08-14 ms/ms.tex

@@
 We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps}.
-This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control responses to individual forcing variables (CO2, nitrogen deposition) and climate  In the case of evapotranspiration (ET), for example, the parameters that control the increase in ET due to warming tend to be different from the parameters that control present-day ET (Figure \ref{fig:et}).
+This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control responses to individual forcing variables (CO2, nitrogen deposition) and climate.
+In the case of evapotranspiration (ET), for example, the parameters that control the increase in ET due to warming tend to be different from the parameters that control present-day ET (Figure \ref{fig:et}).
 ET response can differ significantly from the default case (i.e. stray from the orange line), even when ET is near the default under present-day conditions (i.e. within the shaded region).
-Besides \textit{kmax}, all of the parameters that appear in the top eight parameters for ET response govern plant acclimation to temperature. 
+Besides \textit{kmax}, all of the parameters that appear in the top eight parameters for ET response to future climate govern plant acclimation to temperature. 
 These parameters, while not especially influential on present-day ET, would be among the most influential in governing the trend in ET going forward, illustrating the potential limitations of calibration to ambient conditions.
@@
 \includegraphics[width=\textwidth]{../figs/ET_response.png}
-\caption{(a) Global annual average evapotranspiration (ET) in the warm forcing scenario vs.
+\caption{(a) Global annual average evapotranspiration (ET) in the future climate scenario vs.
 our control experiment.
@@
 The response of ET to future climate with default parameters is +8.8\%.
-The orange line shows the expected future ET should the model conform to the default ET response.
+The orange line shows the expected future ET assuming a linear default ET response.
 (b) The top 8 parameters governing global ET in the control experiment.
@@
 Several spin-off projects have leveraged the work presented here, including some that have already reached publication status \cite{cheng2023,yan2023a,yan2023b}.
-These projects utilized the output from this ensemble to filter for parameters that influence the relevant study domains and used the parameter ranges collated in Section zqz.
+These projects utilized the output from this ensemble to filter for parameters that influence the relevant study domains and used the parameter ranges collated in Section \ref{sect:exps}.
 Our project has accelerated parameter exploration work within our collaborator network by providing:
@@
 \item Parameter ranges for nearly 200 parameters
-\item Machine learning smart low resolution unstructured grid (sparsegrid)
+\item Purpose-built unstructured grid (sparsegrid)
 \item Accelerated spinup procedure
@@
 
-We have likewise begun a follow-on activity that perturbs a subset of important parameters with the Latin hypercube sampling strategy to work towards routine model calibration.
+We have likewise begun a follow-on activity that perturbs a subset of important parameters with the Latin hypercube sampling strategy \cite{Mckay_2000} to work towards routine model calibration.
 Our first foray into exploring the potential for global calibration is with leaf area index due to controls that this variable has on biogeophysical and biogeochemical fluxes.
-By investing in the infrastructure that we introduce here, all of our subsequent paramter perturbation experiments have required much less time and effort.
+By investing in the infrastructure that we introduce here, all of our subsequent parameter perturbation experiments have required much less time and effort.
 We expect to continue to extend our efforts in this domain towards model calibration, as well as repeat the foundational one-at-a-time experiments with subsequent model releases.
-
-\section{Conclusion}
-Do we want / need an explicit conclusion?
 

### 2023-08-14 ms/ms.tex
swap fig 8

@@
 \centering
-\includegraphics[width=\textwidth]{../figs/deltmap_panel.pdf}
-\caption{Screenshot of interactive diagnostic for exploring maps of parameter effects.
-In this case, the effect of medlynslope on runoff within the control ensemble.
-Increasing medlynslope tends to reduce runoff, but only in regions with sufficient vegetation activity.}
+\includegraphics[width=\textwidth]{../figs/QRUNOFFabs_x_medlynslope_CTL2010.png}
+\caption{Map of the effect of perturbing medlynslope on runoff within the control ensemble.
+Increasing medlynslope tends to reduce runoff, but only in regions with sufficient vegetation activity.
+This is one example of many plots available online (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}) in a broader diagnostics set.
+}
 \label{fig:panel}

### 2023-08-14 ms/ms.tex
add text

@@
 \label{sect:md}
-The is developed by the CESM Land Model Working Group and maintained at the National Center for Atmospheric Research (NCAR).
 This experiment utilizes the Community Land Model configuration version 5.1 (CLM5.1) of the Community Terrestrial Systems Model.
@@
 The most influential parameters globally include parameters that were important in each of these three biomes.
-Because the number of potential variables, parameters, and geographical ranges of interest to the wider CLM community is larger than we can document here,  we provide a tool that can be used to explore an extended diagnostics set which summarises the $>$2TB of output data and contains 1720 figures (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
 
@@
 
-In many cases we found it easier to utilize interactive plotting utilities to toggle through the various parameters, output variables, and forcing scenarios (Figure \ref{fig:panel}).
-Interactive widgets are relatively easy to generate within a data analysis session, but more difficult to host online, hence the screenshot.
+Because the number of potential variables, parameters, and geographical ranges of interest to the wider CLM community is larger than we can document here,  we provide a tool that can be used to explore an extended diagnostics set which summarises the $>$2TB of output data via approximately 2000 plots (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
+
+The diagnostics website includes ranking plots (as in Figure \ref{fig:lai}) and maps of parameter effects (as in Figure \ref{fig:panel}).
+But these plots are repeated across a combination of model output variables and model parameters.
+Likewise figures are repeated for each of the various forcing scenarios.
 

### 2023-08-30 ms/ms.tex

@@
 \section{Introduction}
-Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within land components of Earth System Models.
-Understanding how these processes respond to CO$_2$ concentrations, and how they themselves impact CO$_2$ concentrations, is a critical facet of climate change research.  
+Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within the terrestrial components of Earth System Models.
+Understanding how these processes respond to and influence CO$_2$ concentrations is a critical facet of climate change research.
 Our certainty in climate model projections varies by domain and generally decreases with extended time horizons \cite{koven2022}.
@@
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
-Effectively synthesizing the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns can help to parameterize increasingly comprehensive land modeling platforms.
+Effectively synthesizing the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns can help to constrain increasingly comprehensive land modeling platforms.
 
-Uncertainty in the model projections of the carbon cycle is manifest in both uncertainty around the appropriate algorithms and model structures used, and also in the parameterizations of these models.
+Uncertainty in the model projections of the carbon cycle is manifest in both uncertainty around the appropriate algorithms and model structures used, and also in the parameterizations of these models (Luo and Schuur 2020).
 Model inter-comparison projects (MIPs) are the primary means by which structural uncertainty is assessed (refs).
 While these have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
-To allow for more tractable experimental design each model is typically allowed only a single parameterization of many plausible parameter sets.
+To allow for more tractable experimental design each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the cost of conducting multiple experiments with different parameterizations (e.g., in the case of the TRENDY land model intercomparison or CMIP).
 Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land state trajectories.
@@
 PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.
+A parameter sensitivity analysis was conducted with ORCHIDEE using a semi-analytic spin-up (SASU) offered by a matrix approach to understand 34 biogeochemistry parameters according to variance explained by each parameter (Huang et al.
+2018).
 A major goal of this project is to enable automated calibration of the Community Land Model (CLM) \cite{lawrence2019}.
@@
 The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.
+(I again can't figure out how to comment, but this maybe involve a discussion on model naming (CTSM vs.
+CLM))
 
 Relative to CLM5.0, version 5.1 includes minor bug fixes, parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}.
-The PPE experiment required additional code modifications to enable systematic variation of the full suite of model parameters (many of which were previously `hard-coded').
+(Birch et al.
+2019) The PPE experiment required additional code modifications to enable systematic variation of the full suite of model parameters (many of which were previously `hard-coded', i.e., written directly as numerical values in the model source code).
 We utilized the biogeochemical cycling version of CLM, which includes active carbon and nitrogen cycles.
-This is as opposed to the satellite phenology mode, which is driven by observed canopy properties, and was explored in \citeA{dagon2020}.
+This is as opposed to the satellite phenology mode, which is driven by observed canopy properties and thus has fewer internal feedbacks, and was explored in \citeA{dagon2020}.
 We used the model in land-only mode driven by observed meteorological forcing from the GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}), with the crop model turned off.
@@
 
-To manage computational cost we leveraged the Matrix-CN spin-up mode recently implemented within CLM \cite{lu2020}.
-This new module utilizes a linearized simplification of CLM's biogeochemistry to significantly reduce spin-up time.
-Our spin-up protocol featured 20 years in accelerated decomposition mode (see \citeA{lawrence2019} for details), followed by 80 years of Matrix-CN, followed by 40 years of `normal' mode, cycling over a ten-year forcing dataset (described below).
-This protocol was designed to achieve sufficiently equilibrated model states, while minimizing computational time.
-This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter depth).
+To manage computational cost we leveraged the Semi-Analytic spin-up mode (i.e., SASU) recently implemented within CLM5 \cite{lu2020} (Liao et al.
+2023).
+This new module utilizes a matrix representation on CLM's biogeochemistry to significantly reduce spin-up time by one order of magnitude (Liao et al.
+2023, Luo et al.
+2022).
+Our spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (i.e.
+AD) (see \citeA{lawrence2019} (Thornton  &  Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (i.e.
+ND), cycling over a ten-year forcing dataset (described below, Liao et al.
+2023).
+This protocol was designed to achieve sufficiently equilibrated carbon and nitrogen states, while minimizing computational time.
+This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter depth) in very few points at high latitude.
 Certain inferences about deep soil carbon from this ensemble are therefore subject to uncertainty due to spin-up concerns.
 
-\subsection{Sparsegrid}
+\subsection{Sparse grid}
 \label{sect:sg}
-Another control on model cost is resolution.
-Most CLM simulations utilize nominal 1$^\circ$ resolution, which equates to about 20,000 land grid cells.
+Another control on model cost is spatial resolution.
+Most global CLM simulations utilize nominal 1$^\circ$ resolution, which equates to about 20,000 land grid cells.
 In order to manage computational cost, parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}.
@@
 To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter.
-In some cases literature values were directly utilized, but in the many cases, expert judgment was used.
+In some cases literature values were directly utilized, but in many cases, expert judgment was used.
 The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.
@@
 We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
-Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the corresponding mean climate change signal (see Table \ref{tab:exps}).
+Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the corresponding mean climate change signal (see Table \ref{tab:exps}; AF2095 and AF1855).
 We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.
-Our future climate scenario utilizes the SSP3-7.0 forcing in order to allow comparison with the existing CESM2 large ensemble.
+Our future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, forcing in order to allow comparison with the existing CESM2 large ensemble.
 Our software infrastructure has greatly reduced the effort and computational cost associated with ensemble generation, such that it would be easy to add additional ensembles, to explore other emission scenarios or climate regimes (e.g.
@@
    AF2095    & 2091-2100 & 367 & - & late century climate (SSP3-7.0) \\
-   NDEP      & 2005-2014 & 367 & 5g/m$^2$ & enhanced nitrogen deposition \\
+   NDEP      & 2005-2014 & 367 & 5g N m$^-2$  y$^-1$ & enhanced nitrogen deposition \\
  \hline
@@
 Given the challenging size of the CLM5-BGC parameter space, and the need to probe and document parametric uncertainties within operational land surface models, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
-Combining the CN-Matrix spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times cheaper than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
+Combining the SASU spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times cheaper than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
 There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
-Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the CN-matrix approach yielded satisfactory spin-up for our experiment within 140 years.
+Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for our experiment within 140 years.
 
@@
 A perturbation of $\pm$20\% likely exceeds the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils.
-Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, which may belie unforeseen tipping points, brittle parameterizations,  and/or bugs.
+Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, including unforeseen tipping points, brittle parameterizations,  and/or bugs.
 
 A small fraction of parameters tend to explain a large amount of the ensemble variance (Figure \ref{fig:variance}) for any given variable or metric.
-For example the same four parameters explain up to 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig \ref{supp:fcev}).
-Most of the parameter perturbations had a relatively small impact on any given output variable.
-As such, the effective parameter space dimensionality for any given CLM process is likely much smaller than would be expected from having 193 independent parameters, reflecting the fact that land surface models like CLM comprise representations of numerous interconnected domains.
+For example, the same four parameters explain up to 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig \ref{supp:fcev}).
+In general,most of the parameter perturbations had a relatively small impact on any given output variable.
+As such, the effective parameter space dimensionality for any given CLM process is likely much smaller than would be expected from having 193 independent parameters, reflecting the fact that land models like CLM comprise representations of numerous interconnected domains.
 While the principle of land models is that these are all interlinked via complex feedback processes, many parameters nonetheless have impacts that are mostly limited to their own domain.
@@
 \caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across six forcing scenarios.
-Circles mark the CLM5.1 default simulation, and the bars span the ensemble ranges.}
+Circles mark the CLM5.1 default simulation, and the bars span the ensemble ranges generated by individual parameter perturbations.}
 \label{fig:ranges}
@@
 \caption{Cumulative fraction of variance explained by the most influential parameters on gross primary production and canopy evaporation.
-For example approximately 96\% of canopy evaporation variance is explained by the first ten parameters regardless of forcing, whereas between 74 and 92\% of GPP variance is explained by the first ten parameters.
+For example, approximately 96\% of canopy evaporation variance is explained by the first ten parameters regardless of forcing, whereas between 74 and 92\% of GPP variance is explained by the first ten parameters.
 Only the top 50 parameters are shown (of 193) to improve the visualization.
@@
 There were several barriers to perturbing the full set of CLM5-BGC parameters.
-Firstly, many parameters had not been officially identified as such.
+First, many parameters had not been officially identified as such.
 In such cases, we identified hard-coded values, established an appropriate parameter name, and extracted that parameter to the CLM parameter file for easier manipulation.
 In this experiment we perturbed 193 parameters across a variety of CLM processes (Figure \ref{fig:params}).
-This is not the full set of CLM parameters, as some processes were not included, such as crops.
+This is not the full set of CLM parameters, as some processes were not included, such as crop phenology and management.
 In carrying out this experiment, we have likewise realized that the epistemology of climate model `parameters' is highly nuanced, and that defining the parameters within a given model structure can be somewhat subjective, such that it would be difficult to collate a comprehensive or definitive set of parameters for a model like CLM.
@@
 Quantifying parameter effects is a necessary prerequisite to automated calibration and uncertainty quantification.
-Because the parameter space of CLM is quite large and the model response potentially non-linear, large ensembles are required to adequately resolve response surfaces.
+Because the parameter space of CLM is quite large and the model response is potentially non-linear, large ensembles are required to adequately resolve response surfaces.
 With standard CLM configurations, such ensembles would far exceed our computing allocation.
 As such, the computational cost of inferring parameter effects is a major constraint.
-We were able to reduce ensemble generation computational cost 500x by strategically reducing the number of model grid cells (Figure \ref{fig:sg}) and by leveraging a linearized spin-up solver (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
+We were able to reduce ensemble generation computational cost 500x by strategically reducing the number of model grid cells (Figure \ref{fig:sg}) and by leveraging a Semi-Analytic Spin-Up (SASU) offered by the matrix approach to land biogeochemistry models (Lu et al.
+2020, Luo 2022, Liao et al.
+2023) and a linearized spin-up solver (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
 As long as computational resources remain constrained, designing faster model configurations and more efficient sampling strategies will be important for effective model calibration and uncertainty quantification.
@@
 We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:ranges}).
-That being said, the majority of parameter perturbations had small effects for any specific model variable or metric, such that a majority of simulations were clustered around the default simulation.
+That said, the majority of parameter perturbations had small effects for any specific model variable or metric, such that a majority of simulations were clustered around the default simulation.
 In the case of canopy evaporation, for example, more than 95\% of the ensemble variance could be explained by the ten most influential parameters (Figure \ref{fig:variance}).
@@
 Similarly, the parameters controlling leaf area index varied significantly depending on biome (Figure \ref{fig:lai}).
+This emphasizes the importance of testing models not only under mean-state conditions, but also under experimentally perturbed conditions under global change analogues \cite(Wieder:2019).
 
@@
 Figure \ref{fig:panel}).
+This could be useful for investigating obvious model deficiencies, such as regional LAI biases, or to understand potential model responses under different forcing scenarios (e.g.
+plant survival under low CO2 conditions).
 
@@
 \acknowledgments
-Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
+Enter acknowledgments here.
+This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
+This material is based upon work supported by the National Center for Atmospheric Research (NCAR), which is a major facility sponsored by the National Science Foundation under Cooperative Agreement No.
+1852977.
+Computing and data storage resources, including the Cheyenne supercomputer (doi:10.5065/D6RX99HX) were provided by the Climate Simulation Laboratory at NCAR’s Computational and Information Systems Laboratory (CISL).
+FL is supported by the National Key Research and Development Program of China (2022YFE0106500).
+CDK acknowledges support by the Director, Office of Science, Office of Biological and Environmental Research of the US Department of Energy under contract DE-AC02-05CH11231 through the Regional and Global Model Analysis Program (RUBISCO SFA).
 

### 2023-08-31 ms/ms.tex
adding refs

@@
 
-Several promising avenues for model calibration are in development (lavendar, musa, CES), many of which require the construction of large parameter perturbation ensembles (PPEs) \cite{qian2018}. 
+Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large parameter perturbation ensembles (PPEs) \cite{qian2018}.
 PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
@@
 Our spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (i.e.
-AD) (see \citeA{lawrence2019} (Thornton  &  Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (i.e.
+AD) (see \citeA{lawrence2019} (Thornton and Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (i.e.
 ND), cycling over a ten-year forcing dataset (described below, Liao et al.

### 2023-08-31 ms/ms.tex
overleaf

@@
 
-Uncertainty in the model projections of the carbon cycle is manifest in both uncertainty around the appropriate algorithms and model structures used, and also in the parameterizations of these models (Luo and Schuur 2020).
-Model inter-comparison projects (MIPs) are the primary means by which structural uncertainty is assessed (refs).
+Uncertainty in the model projections of the carbon cycle is manifest in uncertainty around the appropriate algorithms and equations used, and also in the parameterizations of these models \cite{luo2020}.
+Model inter-comparison projects (MIPs) are the primary means by which structural uncertainty is assessed.
 While these have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
@@
 PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.
-A parameter sensitivity analysis was conducted with ORCHIDEE using a semi-analytic spin-up (SASU) offered by a matrix approach to understand 34 biogeochemistry parameters according to variance explained by each parameter (Huang et al.
+A parameter sensitivity analysis was conducted with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) using a semi-analytic spin-up (SASU) offered by a matrix approach to understand 34 biogeochemistry parameters according to variance explained by each parameter (Huang et al.
 2018).

### 2023-08-31 ms/ms.tex
adding refs

@@
 Uncertainty in the model projections of the carbon cycle is manifest in uncertainty around the appropriate algorithms and equations used, and also in the parameterizations of these models \cite{luo2020}.
-Model inter-comparison projects (MIPs) are the primary means by which structural uncertainty is assessed.
+Model inter-comparison projects (MIPs) are the primary means by which structural uncertainty is assessed \cite{henderson-sellers1995,pitman1999,wood1998,schlosser2000,eyring2016,friedlingstein2022}.
 While these have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
-To allow for more tractable experimental design each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the cost of conducting multiple experiments with different parameterizations (e.g., in the case of the TRENDY land model intercomparison or CMIP).
+To allow for more tractable experimental design each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the cost of conducting multiple experiments with different parameterizations, e.g., in the case of the TRENDY land model intercomparison or the coupled model intercomparisons \cite{friedlingstein2022,eyring2016}.
 Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land state trajectories.
@@
 2018).
-A major goal of this project is to enable automated calibration of the Community Land Model (CLM) \cite{lawrence2019}.
+A major goal of our project is to enable automated calibration of the Community Land Model (CLM) \cite{lawrence2019}.
 This extends the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
@@
 \label{sect:md}
-This experiment utilizes the Community Land Model configuration version 5.1 (CLM5.1) of the Community Terrestrial Systems Model.
+This experiment utilizes the Community Land Model configuration (version 5.1, i.e.
+CLM5.1) of the Community Terrestrial Systems Model.
 The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.
@@
 
-We have likewise begun a follow-on activity that perturbs a subset of important parameters with the Latin hypercube sampling strategy \cite{Mckay_2000} to work towards routine model calibration.
+We have likewise begun a follow-on activity that perturbs a subset of important parameters with the Latin hypercube sampling strategy \cite{mckay2000} to work towards routine model calibration.
 Our first foray into exploring the potential for global calibration is with leaf area index due to controls that this variable has on biogeophysical and biogeochemical fluxes.

### 2023-08-31 ms/ms.tex
overleaf

@@
 2018).
-A major goal of our project is to enable automated calibration of the Community Land Model (CLM) \cite{lawrence2019}.
-This extends the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
+A major goal of our project is to enable automated calibration of the Community Terrestrial Systems Model (CTSM) \cite{lawrence2019}.
+We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
 Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization.
@@
 \citeA{collier2018}.
-Our goal in this project is to systematize the parameter perturbation process within the CLM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.
+Our goal in this project is to systematize the parameter perturbation process within the CTSM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.
 In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
@@
 The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.
-(I again can't figure out how to comment, but this maybe involve a discussion on model naming (CTSM vs.
-CLM))
 

### 2023-08-31 ms/ms.tex
refs

@@
 
-Relative to CLM5.0, version 5.1 includes minor bug fixes, parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}.
-(Birch et al.
-2019) The PPE experiment required additional code modifications to enable systematic variation of the full suite of model parameters (many of which were previously `hard-coded', i.e., written directly as numerical values in the model source code).
+Relative to CLM5.0, version 5.1 includes minor bug fixes, parameter adjustments \cite{birch2021}, and the implementation of biomass heat storage \cite{swenson2019}.
+The PPE experiment required additional code modifications to enable systematic variation of the full suite of model parameters (many of which were previously `hard-coded', i.e., written directly as numerical values in the model source code).
 We utilized the biogeochemical cycling version of CLM, which includes active carbon and nitrogen cycles.

### 2023-08-31 ms/ms.tex
Updates from Overleaf

@@
 
-Relative to CLM5.0, version 5.1 includes minor bug fixes, parameter adjustments \cite{birch2021}, and the implementation of biomass heat storage \cite{swenson2019}.
+Relative to CLM5.0, version 5.1 includes minor bug fixes, parameter adjustments, and the implementation of biomass heat storage \cite{swenson2019}.
 The PPE experiment required additional code modifications to enable systematic variation of the full suite of model parameters (many of which were previously `hard-coded', i.e., written directly as numerical values in the model source code).

### 2023-08-31 ms/ms.tex
resolving comments etc

--- 2023-08-31 ms/ms.tex
@@
 2022).
-Our spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (i.e.
-AD) (see \citeA{lawrence2019} (Thornton and Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (i.e.
+Our spin-up protocol featured 20 years in Accelerated Decomposition (AD) spin-up mode (see \citeA{lawrence2019} (Thornton and Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (i.e.
 ND), cycling over a ten-year forcing dataset (described below, Liao et al.
@@
 Another control on model cost is spatial resolution.
-Most global CLM simulations utilize nominal 1$^\circ$ resolution, which equates to about 20,000 land grid cells.
+Most global CTSM simulations utilize nominal 1$^\circ$ resolution, which equates to about 20,000 land grid cells.
 In order to manage computational cost, parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}.
 Here, we use a clustering algorithm to achieve an alternative low resolution configuration.
+The goal is to more strategically distribute grid cells, opting for more grid cells in areas that are especially distinct or variable, and fewer in areas that can be easily approximated based on output elsewhere (e.g.
+neighboring grid cells).
 
 Multivariate spatio-temporal clustering (MVSC) has been utilized to extract patterns of climatological significance from climate model output \cite{hoffman2005} and applied to design a representativeness-based sampling network \cite{hoffman2013}.
-Instead of lowering resolution by coarsening a rectilinear grid, we here used MVSC to strategically remove effectively redundant grid cells, leaving only 400 grid cells that efficiently sample important model dynamics across the globe.
+Instead of lowering resolution by coarsening a rectilinear grid, we here used MVSC to strategically remove effectively redundant grid cells.
 
-We used k-means clustering to identify groups of grid cells with similar dynamics based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
+We used k-means clustering to identify groups of grid cells with similar dynamics based on a 2$^\circ$ transient simulation (1850-2014) using the CTSM-PPE codebase.
 We selected one representative grid cell from each cluster to stand in for the entire cluster.
@@
 
-Clustering was based on a subset of 18 meaningful CLM variables (Table \ref{tab:sg}).
-The clustering algorithm analyzed 12 observations of each variable per grid cell, namely the mean and interannual variability computed for six 30-year climatology windows (1865-1894, 1895-1924, ...
+Clustering was based on a subset of 18 CTSM variables (Table \ref{tab:sg}).
+The MVSC algorithm delineates clusters based on the mean and variability of each variable at each grid cell computed for six 30-year climatology windows (1865-1894, 1895-1924, ...
 , 1985-2014).
@@
 \begin{table}[h]
-\caption{Clustering inputs categorized into three groups, with the CLM variable name in parentheses}
+\caption{Clustering inputs categorized into three groups, with the CTSM variable name in parentheses}
 \centering
@@
  \hline
- Climate forcing variables & Ecosystem state variables &Ecosystem flux variables \\
+ Climate variables & Ecosystem state variables &Ecosystem flux variables \\
  \hline
@@
 \subsection{Experimental Design}
-Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants previously considered as `internal' are hard-coded and not amenable to systematic perturbation.
-For the purposes of this activity, we identified what we consider a very broad set of 193 CLM-BGC parameters, and enabled their modification via the model input files.
-We opted not to perturb crop parameters within this ensemble, as our initial focus is on unmanaged land.
+Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants were hard-coded and not amenable to systematic perturbation.
+For the purposes of this activity, we identified a broad set of 193 CLM-BGC parameters, and enabled their modification via the model input files.
+Several domains were not perturbed, including crops, CH$_4$, VOC, and dust emissions, urban properties, and river flow parameters.
+The crop model was not utilized as our initial focus is on unmanaged land.
 

### 2023-08-31 ms/ms.tex
overleaf

@@
 
-The process of parameter selection for land  models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryology, biogeochemistry, physiology, land management, micrometeorology, etc.).
+The process of parameter selection for land  models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryosphere, biogeochemistry, physiology, land management, etc.).
 The number of model parameters required to represent all the relevant processes is large, and model speeds are relatively slow, presenting a major barrier to robust objective calibration, model interpretation and development \cite{fisher2020,dagon2020}.
@@
 Plant hydraulics parameters were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
-The most influential parameters globally include parameters that were important in each of these three biomes.
+The most influential parameters for leaf area index globally include parameters that were important in each of these three biomes.
 
@@
 In such cases, we identified hard-coded values, established an appropriate parameter name, and extracted that parameter to the CLM parameter file for easier manipulation.
-In this experiment we perturbed 193 parameters across a variety of CLM processes (Figure \ref{fig:params}).
-This is not the full set of CLM parameters, as some processes were not included, such as crop phenology and management.
+Although in this experiment we perturbed a large number of parameters (193) across a variety of CLM processes (Figure \ref{fig:params}), this still does not cover the full set of CLM parameters, as some processes were not included, such as crop phenology and management.
 In carrying out this experiment, we have likewise realized that the epistemology of climate model `parameters' is highly nuanced, and that defining the parameters within a given model structure can be somewhat subjective, such that it would be difficult to collate a comprehensive or definitive set of parameters for a model like CLM.
@@
 plant survival under low CO2 conditions).
+
+Land only caveat?
+Our PPE is run as land-only simulations, using prescribed atmospheric forcing.
+This is likely to be adequate for estimating many land surface fluxes and pools, in particular of carbon, but could lead to biased estiamtes of water and energy fluxes, as we know that atmospheric feedbacks to land surface characteristics modify the state of the land surface itself (Laguë et al.
+2019).
+Additional work is planned to leverage this one-at-a-time land-only PPE to quatnify how much, where, and for which processes atmospheric feedbacks are likely to change the land surface state substatnially.
 

### 2023-09-01 ms/ms.tex
overleaf

@@
 While these have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
-To allow for more tractable experimental design each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the cost of conducting multiple experiments with different parameterizations, e.g., in the case of the TRENDY land model intercomparison or the coupled model intercomparisons \cite{friedlingstein2022,eyring2016}.
+To allow for more tractable experimental design, in most MIPs, each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the cost of conducting multiple experiments with different parameterizations, e.g., in the case of the TRENDY land model intercomparison or the coupled model intercomparisons \cite{friedlingstein2022,eyring2016}.
 Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land state trajectories.
@@
 Although in this experiment we perturbed a large number of parameters (193) across a variety of CLM processes (Figure \ref{fig:params}), this still does not cover the full set of CLM parameters, as some processes were not included, such as crop phenology and management.
-In carrying out this experiment, we have likewise realized that the epistemology of climate model `parameters' is highly nuanced, and that defining the parameters within a given model structure can be somewhat subjective, such that it would be difficult to collate a comprehensive or definitive set of parameters for a model like CLM.
+In carrying out this process of parameter identification, we  likewise unearthed many nuances in  the epistemology of what  `parameters' are in the context of Earth system models.
+Defining the parameters within a given model structure can be somewhat subjective, such that it would be difficult to collate a comprehensive or definitive set of parameters for a model like CLM.
 
@@
 In some cases, literature values were explicitly used (e.g.
-medlynslope, \citeA{lin2015}), but the most common range was $\pm20\%$.
+for the slope parameter of the stomatsl conductance model, 'medlynslope', \citeA{lin2015}), but the most common range was $\pm20\%$.
 It is exceedingly difficult to set parameter ranges that sample comparable probability density, even in a univariate experiment, like this one.
 The number of parameters is large, with many lacking sufficient empirical backing for robust range evaluation.
+Defining appropriate ranges for the full suite of land model parameters is a major activity of the scientific subdomains that comprise the land model (e.g.
+Kattge et al.
+2020).
 Even with parameters that have an empirical basis, they may behave differently at the coarse climate modeling spatial resolution (zqz), complicating the process of uncertainty quantification.
@@
 Because the parameter space of CLM is quite large and the model response is potentially non-linear, large ensembles are required to adequately resolve response surfaces.
-With standard CLM configurations, such ensembles would far exceed our computing allocation.
+With standard CLM configurations, such ensembles would far exceed our computing resources.
 As such, the computational cost of inferring parameter effects is a major constraint.
@@
 Our PPE is run as land-only simulations, using prescribed atmospheric forcing.
-This is likely to be adequate for estimating many land surface fluxes and pools, in particular of carbon, but could lead to biased estiamtes of water and energy fluxes, as we know that atmospheric feedbacks to land surface characteristics modify the state of the land surface itself (Laguë et al.
+This is likely to be adequate for estimating many land surface fluxes and pools, in particular of carbon, but could lead to biased estimates of water and energy fluxes, as we know that atmospheric feedbacks to land surface characteristics modify the state of the land surface itself (Laguë et al.
 2019).
@@
 
-The model code for this experiment is contained in a development tag (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
+The model code for this experiment is contained in a development tag of the CTSM (\url{https://github.com/ESCOMP/CTSM/tree/branch_tags/PPE.n11_ctsm5.1.dev030}).
 
- (CTSM component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP})
+ (The CTSM component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP})
 
@@
 CDK acknowledges support by the Director, Office of Science, Office of Biological and Environmental Research of the US Department of Energy under contract DE-AC02-05CH11231 through the Regional and Global Model Analysis Program (RUBISCO SFA).
+RF and BS acknowledge funding by the European Union’s Horizon 2020 (H2020) research and innovation program under Grant Agreement No.
+101003536 (ESM2025 – Earth System Models for the Future) and 821003 (4C, Climate-Carbon Interactions in the Coming Century).
 

### 2023-09-01 ms/ms.tex
affiliations

@@
 
-\authors{D. Kennedy\affil{1} and ...}
+\authors{D. Kennedy\affil{1,2}, K. Dagon\affil{1}, D.M. Lawrence\affil{1}, R.A. Fisher\affil{3}, B.M. Sanderson\affil{3}, 
+N. Collier\affil{4}, F.M. Hoffman\affil{4}, C.D. Koven\affil{5}, E. Kluzek\affil{1}, S. Levis\affil{1}, X. Lu\affil{6}, K.W. Oleson\affil{1}, C.M. Zarakas\affil{7}, 
+G. Bonan\affil{1}, Y. Cheng\affil{8}, A.C. Foster\affil{1}, M.D. Fowler\affil{1}, L.R. Hawkins\affil{9}, T. Kavoo\affil{10}, S. Kumar\affil{10}, A.J. Newman\affil{8}, P.J. Lawrence\affil{1}, 
+F. Li\affil{11}, D.L. Lombardozzi\affil{1,12}, Y. Luo\affil{12}, J.K. Shuman\affil{13}, A.L.S. Swann\affil{7,14}, S.C. Swenson\affil{1}, G. Tang\affil{1}, W.R. Wieder\affil{1}, and A.W. Wood\affil{1}}
 
-\affiliation{1}{Climate and Global Dynamics Laboratory, NCAR, Boulder, CO, USA.}
+\affiliation{1}{Climate and Global Dynamics Laboratory, NCAR, Boulder, CO, USA}
+\affiliation{2}{Earth Research Institute, University of California, Santa Barbara, CA, USA}
+\affiliation{3}{CICERO Centre for International Climate and Environmental Research, Oslo, Norway}
+\affiliation{4}{Oak Ridge National Laboratory, Oak Ridge, TN, USA}
+\affiliation{5}{Climate and Ecosystem Sciences Division, Lawrence Berkeley National Lab, Berkeley, CA, USA}
+\affiliation{6}{School of Atmospheric Sciences, Sun Yat-sen University, Guangzhou, Guangdong, China}
+\affiliation{7}{Department of Atmospheric Sciences, University of Washington, Seattle, WA, USA}
+\affiliation{8}{Research Applications Laboratory, National Center for Atmospheric Research, Boulder, CO, USA}
+\affiliation{9}{Department of Earth and Environmental Engineering, Columbia University, New York, NY, USA}
+\affiliation{10}{College of Forestry, Wildlife and Environment, Auburn University, AL, USA}
+\affiliation{11}{International Center for Climate and Environment Sciences, Institute of Atmospheric Physics, Chinese Academy of Sciences, Beijing, China}
+\affiliation{12}{Colorado State University, Fort Collins, CO, USA}
+\affiliation{13}{School of Integrative Plant Science, Cornell University, NY, USA}
+\affiliation{14}{Earth Science Division, National Aeronautics and Space Administration Ames Research Center, Moffett Field, CA, USA}
+\affiliation{15}{Department of Biology, University of Washington, Seattle, WA, USA}
 \correspondingauthor{Daniel Kennedy}{djk2120@ucar.edu}

### 2023-09-01 pyth/utils.py
update template

@@ -1,0 +1,218 @@
+import xarray as xr
+import pandas as pd
+import numpy as np
+import glob
+
+
+def get_exp(exp,tape='h0',dvs=[]):
+    
+    oaats=['AF1855','AF2095','C285','C867','NDEP','CTL2010']
+    if exp in oaats:
+        key='/glade/campaign/asp/djk2120/PPEn11/csvs/surviving.csv'
+        top='/glade/campaign/asp/djk2120/PPEn11/'
+        code='OAAT'
+        yy=()
+        minmax=True
+        
+        
+        
+    files,appends,dims=get_files(exp,tape,code=code,yy=yy,top=top,key=key,minmax=minmax)
+    ds=get_ds(files,dims,dvs=dvs,appends=appends)
+    
+    return ds
+
+def amean(da):
+    #annual mean of monthly data
+    m  = da['time.daysinmonth']
+    cf = 1/365
+    xa = cf*(m*da).groupby('time.year').sum().compute()
+    xa.name=da.name
+    xa.attrs=da.attrs
+    return xa
+
+def gmean(da,la):
+    if 'gridcell' in da.dims:
+        dim='gridcell'
+    else:
+        dim=['lat','lon']
+    x=(da*la).sum(dim=dim)/la.sum()
+    return x.compute()
+
+def get_map(da,sgmap=None,file='sgmap.nc'):
+    if not sgmap:
+        sgmap=xr.open_dataset(file)
+    return da.sel(gridcell=sgmap.cclass).where(sgmap.notnan).compute()
+
+def rank_plot(da,nx,ax=None):
+    x = top_n(da,nx)
+    xdef = da.sel(param='default',minmax='min')
+    
+    if x.dims[0]=='minmax':
+        x=x.T
+    
+    if not ax:
+        fig=plt.figure()
+        ax=fig.add_subplot()
+    
+    ax.plot([xdef,xdef],[0,nx-1],'k:',label='default')
+    ax.scatter(x.sel(minmax='min'),range(nx),marker='o',facecolors='none', edgecolors='r',label='low-val')
+    ax.plot(x.sel(minmax='max'),range(nx),'ro',label='high-val')
+
+    i=-1
+    for xmin,xmax in x:
+        i+=1
+        ax.plot([xmin,xmax],[i,i],'r')
+    ax.set_yticks(range(nx))
+    ax.set_yticklabels([p[:15] for p in x.param.values]) #[:15] shortens param names to 15 characters max
+        
+def top_n(da,nx):
[…]

### 2023-09-01 ms/ms.tex
Updates from Overleaf

@@
 2019).
-Additional work is planned to leverage this one-at-a-time land-only PPE to quatnify how much, where, and for which processes atmospheric feedbacks are likely to change the land surface state substatnially.
+Additional work is planned to leverage this one-at-a-time land-only PPE to quantify how much, where, and for which processes atmospheric feedbacks are likely to change the land surface state substantially.
 

### 2023-09-01 ms/ms.tex
reorganize intro

@@
 
-The process of parameter selection for land  models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryosphere, biogeochemistry, physiology, land management, etc.).
+Parametric sensitivity tests can be used to gauge parametric uncertainty (e.g. need refs).
+A collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with many examples in the coupled (refs) and land-only contexts (huang2018) \cite{dagon2020}.
+PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
+PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.
+Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large parameter perturbation ensembles (PPEs) \cite{qian2018}.
+In this project we create a large PPE of the Community Terrestrial Systems Model (CTSM) \cite{lawrence2019}, in the Community Land Model with biogeochemistry configuration (CLM-BGC).
+We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
+Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization.
+
+One goal of this project is to enable automated calibration of CLM-BGC.
+The process of parameter estimation for land models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryosphere, biogeochemistry, physiology, land management, etc.).
 The number of model parameters required to represent all the relevant processes is large, and model speeds are relatively slow, presenting a major barrier to robust objective calibration, model interpretation and development \cite{fisher2020,dagon2020}.
@@
 It typically requires extensive computational time,  software engineering, and domain-specific expert scientific knowledge.
-
-Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large parameter perturbation ensembles (PPEs) \cite{qian2018}.
-PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.
-A parameter sensitivity analysis was conducted with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) using a semi-analytic spin-up (SASU) offered by a matrix approach to understand 34 biogeochemistry parameters according to variance explained by each parameter (Huang et al.
-2018).
-A major goal of our project is to enable automated calibration of the Community Terrestrial Systems Model (CTSM) \cite{lawrence2019}.
-We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
-Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization.
+While we are not yet at the point of automated calibration, we present herein several innovations that will serve to make an automated calibration system more technically feasible.
 
@@
 Certain inferences about deep soil carbon from this ensemble are therefore subject to uncertainty due to spin-up concerns.
+
+ A parameter sensitivity analysis was conducted with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) using a semi-analytic spin-up (SASU) offered by a matrix approach to understand 34 biogeochemistry parameters according to variance explained by each parameter (Huang et al.
+2018).
 

### 2023-10-11 ms/ms.tex
Updates from Overleaf

@@
 \authors{D. Kennedy\affil{1,2}, K. Dagon\affil{1}, D.M. Lawrence\affil{1}, R.A. Fisher\affil{3}, B.M. Sanderson\affil{3}, 
-N. Collier\affil{4}, F.M. Hoffman\affil{4}, C.D. Koven\affil{5}, E. Kluzek\affil{1}, S. Levis\affil{1}, X. Lu\affil{6}, K.W. Oleson\affil{1}, C.M. Zarakas\affil{7}, 
-G. Bonan\affil{1}, Y. Cheng\affil{8}, A.C. Foster\affil{1}, M.D. Fowler\affil{1}, L.R. Hawkins\affil{9}, T. Kavoo\affil{10}, S. Kumar\affil{10}, A.J. Newman\affil{8}, P.J. Lawrence\affil{1}, 
+N.
+Collier\affil{4}, F.M.
+Hoffman\affil{4}, C.D.
+Koven\affil{5}, E.
+Kluzek\affil{1}, S.
+Levis\affil{1}, X.
+Lu\affil{6}, K.W.
+Oleson\affil{1}, C.M.
+Zarakas\affil{7}, Y.
+Cheng\affil{8}, A.C.
+Foster\affil{1}, M.D.
+Fowler\affil{1}, L.R.
+Hawkins\affil{9}, T.
+Kavoo\affil{10}, S.
+Kumar\affil{10}, A.J.
+Newman\affil{8}, P.J.
+Lawrence\affil{1},
 F. Li\affil{11}, D.L. Lombardozzi\affil{1,12}, Y. Luo\affil{12}, J.K. Shuman\affil{13}, A.L.S. Swann\affil{7,14}, S.C. Swenson\affil{1}, G. Tang\affil{1}, W.R. Wieder\affil{1}, and A.W. Wood\affil{1}}
@@
 While these have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
-To allow for more tractable experimental design, in most MIPs, each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the cost of conducting multiple experiments with different parameterizations, e.g., in the case of the TRENDY land model intercomparison or the coupled model intercomparisons \cite{friedlingstein2022,eyring2016}.
+To allow for more tractable experimental design, in most MIPs, each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the cost of conducting multiple experiments with different parameterizations, e.g.
+the TRENDY land model intercomparison or the coupled model intercomparison project \cite{friedlingstein2022,eyring2016}.
 Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land state trajectories.
 
-Parametric sensitivity tests can be used to gauge parametric uncertainty (e.g. need refs).
+Parameter sensitivity tests can be used to gauge parametric uncertainty (e.g. need refs).
 A collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with many examples in the coupled (refs) and land-only contexts (huang2018) \cite{dagon2020}.
@@
 Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large parameter perturbation ensembles (PPEs) \cite{qian2018}.
-In this project we create a large PPE of the Community Terrestrial Systems Model (CTSM) \cite{lawrence2019}, in the Community Land Model with biogeochemistry configuration (CLM-BGC).
+In this project we create a large PPE with the Community Terrestrial Systems Model (CTSM) \cite{lawrence2019}, in the Community Land Model with biogeochemistry configuration (CLM-BGC).
 We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
@@
 The number of model parameters required to represent all the relevant processes is large, and model speeds are relatively slow, presenting a major barrier to robust objective calibration, model interpretation and development \cite{fisher2020,dagon2020}.
-Thorough assessment of the parameter sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}.
+Thorough assessment of the parametric sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}.
 It typically requires extensive computational time,  software engineering, and domain-specific expert scientific knowledge.
+Our work here tries to alleviate each of these challenges in turn, at least within the context of the CLM-BGC model.
 While we are not yet at the point of automated calibration, we present herein several innovations that will serve to make an automated calibration system more technically feasible.
@@
 \citeA{collier2018}.
-Our goal in this project is to systematize the parameter perturbation process within the CTSM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes.
+Our goal in this project is to systematize the parameter perturbation process within the CTSM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across a wide range of land model processes.
 In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
@@
 In some cases, literature values were explicitly used (e.g.
-for the slope parameter of the stomatsl conductance model, 'medlynslope', \citeA{lin2015}), but the most common range was $\pm20\%$.
+for the slope parameter of the stomatal conductance model, 'medlynslope', \citeA{lin2015}), but the most common range was $\pm20\%$.
 It is exceedingly difficult to set parameter ranges that sample comparable probability density, even in a univariate experiment, like this one.

### 2023-10-12 ms/ms.tex
Updates from Overleaf

@@
 \centering
-\includegraphics[width=\textwidth]{../figs/bar.pdf}
+\includegraphics[width=\textwidth]{../figs/main/bar.png}
 \caption{193 parameters were perturbed across the various domains of the land model.}
@@
 \centering
-\includegraphics[width=30pc]{../figs/sims.pdf}
+\includegraphics[width=30pc]{../figs/main/sims.png}
 \caption{The approximate number of simulations afforded by 1 million core-hours on the Cheyenne supercomputer for a range of CLM configurations.
-Configurations are labeled according to spin-up procedure (CN-matrix or the standard 1500-year spinup) and horizontal resolution (`sg' signifies sparsegrid).
+Configurations are labeled according to spin-up procedure (SASU or the standard 1500-year spinup) and horizontal resolution (`sg' signifies sparsegrid).
 The inset map shows the locations of the 400 sparse grid cells.

### 2023-12-19 postp/batch.sh
postprocessing

@@ -1,0 +1,8 @@
+for i in {0..211}; do
+    echo $i
+
+    job='job'$i'.job'
+    sed 's/num/'$i'/g' template.sh > $job
+    qsub $job
+        
+done

### 2023-12-19 postp/depr/oatdiff.py

@@ -1,0 +1,24 @@
+import xarray as xr
+import numpy as np
+import glob
+
+
+f='/glade/campaign/cgd/tss/projects/PPE/PPEn11_OAAT/helpers/sparsegrid_landarea.nc'
+la=xr.open_dataset(f).landarea
+lasum=la.sum().compute()
+
+def gmean(da):
+    return 1/lasum*(da*la).sum(dim='gridcell')
+
+d='/glade/campaign/cgd/tss/projects/PPE/PPEn11_OAAT/C285/hist/'
+
+f0=d+'PPEn11_C285_OAAT0000.clm2.h0.2005-02-01-00000.nc'
+f1=d+'PPEn11_C285_OAAT0221.clm2.h0.2005-02-01-00000.nc'
+
+
+ds0=xr.open_dataset(f0)
+ds1=xr.open_dataset(f1)
+
+for v in ds0.data_vars:
+    if not ds0[v].equals(ds1[v]):
+        print(v)

### 2023-12-19 postp/depr/qc.py

@@ -1,0 +1,32 @@
+import xarray as xr
+import numpy as np
+import glob
+
+
+f='/glade/campaign/cgd/tss/projects/PPE/PPEn11_OAAT/helpers/sparsegrid_landarea.nc'
+la=xr.open_dataset(f).landarea
+lasum=la.sum().compute()
+
+def gmean(da):
+    return 1/lasum*(da*la).sum(dim='gridcell')
+
+d='/glade/campaign/cgd/tss/projects/PPE/PPEn11_OAAT/C285/hist/'
+
+f0=d+'PPEn11_C285_OAAT0000.clm2.h0.2005-02-01-00000.nc'
+f1=d+'PPEn11_C285_OAAT0221.clm2.h0.2005-02-01-00000.nc'
+
+
+ds0=xr.open_dataset(f0)
+ds1=xr.open_dataset(f1)
+
+files=sorted(glob.glob(d+'*.h0.*'))
+f=d+'PPEn11_C285_OAAT0222.clm2.h0.2005-02-01-00000.nc'
+ds=xr.open_dataset(f)
+
+cf=24*60*60
+
+k='OAAT'+f.split('OAAT')[2].split('.')[0]
+print(k,ds.GPP.equals(ds0.GPP),ds.GPP.equals(ds1.GPP))
+print(k,np.round(cf*gmean(ds.GPP).mean(dim='time').values,3),
+      np.round(cf*gmean(ds0.GPP).mean(dim='time').values,3),
+               np.round(cf*gmean(ds1.GPP).mean(dim='time').values,3))

### 2023-12-19 postp/enscat.sh

@@ -1,0 +1,16 @@
+#!/bin/bash
+#PBS -N enscat
+#PBS -q casper
+#PBS -l walltime=1:00:00
+#PBS -A P93300641
+#PBS -j oe
+#PBS -k eod
+#PBS -l select=1:ncpus=1
+
+
+module load nco
+ncecat -u param /glade/scratch/djk2120/postp/oaat/oaat* postp.nc
+source ~/.bashrc
+conda activate ppe-py
+python label.py
+rm postp.nc

### 2023-12-19 postp/label.py

@@ -1,0 +1,15 @@
+import xarray as xr
+import pandas as pd
+
+ds=xr.open_dataset('postp.nc')
+key='/glade/campaign/cgd/tss/projects/PPE/PPEn11_OAAT/helpers/surviving.csv'
+df=pd.read_csv(key)
+ds['param']=df.param.values
+
+ds.attrs={'contact':'djk2120@ucar.edu',
+          'date':'18-Dec-2023',
+          'key':key,
+          'source':'/glade/campaign/cgd/tss/projects/PPE/PPEn11_OAAT',
+          'script':'/glade/u/home/djk2120/oaat_clm5_ppe/postp/oaat_postp.py'}
+
+ds.to_netcdf('oaat_postp.nc')

### 2023-12-19 postp/oaat_postp.py

@@ -1,0 +1,105 @@
+import xarray as xr
+import numpy as np
+import pandas as pd
+import sys
+
+def fix_time(ds):
+    yr0=str(ds['time.year'][0].values)
+    nt=len(ds.time)
+    ds['time'] = xr.cftime_range(yr0,periods=nt,freq='MS',calendar='noleap') #fix time bug
+    return ds
+
+def amean(da):
+    #annual mean of monthly data
+    m  = da['time.daysinmonth']
+    cf = 1/365
+    xa = cf*(m*da).groupby('time.year').sum().compute()
+    return xa
+
+def gmean(da,la):
+    if 'gridcell' in da.dims:
+        dim='gridcell'
+    else:
+        dim=['lat','lon']
+    x=(da*la).sum(dim=dim)/la.sum()
+
+    return x.compute()
+
+def bmean(da,la,biome):
+    x=1/la.groupby(biome).sum()*(da*la).groupby(biome).sum()
+    return x.compute()
+
+def pmean(da,lapft):
+    x=1/lapft.groupby(da.pft).sum()*(da*lapft).groupby('pft').sum()
+    return x.compute()
+
+def cpds(ds1,ds2,suff):
+    for v in ds1.data_vars:
+        ds2[v+suff]=ds1[v]
+    return ds2
+
+def crunch(ds,op,domain,la=[],biome=[],lapft=[]):
+    xx=[]
+    if op=='mean':
+        x=amean(ds).mean(dim='year')
+    elif op=='std':
+        x=amean(ds).std(dim='year')
+    elif op=='amp':
+        x=(ds.groupby('time.year').max()-ds.groupby('time.year').min()).mean(dim='year')
+          
+    if domain=='global':
+        xx=gmean(x,la)
+    elif domain=='biome':
+        xx=bmean(x,la,biome)
+    elif domain=='pft':
+        xx=pmean(x,lapft)
+        
+    for v in ds.data_vars:
+        xx[v].attrs=ds[v].attrs
+    
+    return xx
+            
+def postp(exp,key):
+    dsout=xr.Dataset()
+    f0='/glade/campaign/cgd/tss/projects/PPE/PPEn11_OAAT/{}/hist/PPEn11_{}_{}.clm2.{}.2005-02-01-00000.nc'
+    f=f0.format(exp,exp,key,'h0')
+    dvs=['GPP','TLAI']
[…]

### 2023-12-19 postp/template.sh

@@ -1,0 +1,18 @@
+#!/bin/bash
+#PBS -N oaat_postp_num
+#PBS -q casper
+#PBS -l walltime=1:00:00
+#PBS -A P93300641
+#PBS -j oe
+#PBS -k eod
+#PBS -l select=1:ncpus=1
+
+source ~/.bashrc
+conda activate ppe-py
+
+python oaat_postp.py num
+
+
+
+
+

### 2023-12-19 ms/ms.tex

@@
 Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants were hard-coded and not amenable to systematic perturbation.
-For the purposes of this activity, we identified a broad set of 193 CLM-BGC parameters, and enabled their modification via the model input files.
+For the purposes of this activity, we identified a broad set of 211 CLM-BGC parameters, and enabled their modification via the model input files.
 Several domains were not perturbed, including crops, CH$_4$, VOC, and dust emissions, urban properties, and river flow parameters.
@@
    AF2095    & 2091-2100 & 367 & - & late century climate (SSP3-7.0) \\
-   NDEP      & 2005-2014 & 367 & 5g N m$^-2$  y$^-1$ & enhanced nitrogen deposition \\
+   NDEP      & 2005-2014 & 367 & 5g N m$^{-2}$  y$^{-1}$ & enhanced nitrogen deposition \\
  \hline
@@
 
-A total of 193 parameters were targeted for sensitivity testing across a wide variety of CLM subcomponents (Figure \ref{fig:params}).
+Given the challenging size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the need to probe and document parametric uncertainties within operational land surface models, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
 
@@
 \includegraphics[width=\textwidth]{../figs/main/bar.png}
-\caption{193 parameters were perturbed across the various domains of the land model.}
+\caption{211 parameters were perturbed across the various domains of the land model.}
 \label{fig:params}
@@
 
-Given the challenging size of the CLM5-BGC parameter space, and the need to probe and document parametric uncertainties within operational land surface models, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
 Combining the SASU spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times cheaper than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
@@
 In general,most of the parameter perturbations had a relatively small impact on any given output variable.
-As such, the effective parameter space dimensionality for any given CLM process is likely much smaller than would be expected from having 193 independent parameters, reflecting the fact that land models like CLM comprise representations of numerous interconnected domains.
+As such, the effective parameter space dimensionality for any given CLM process is likely much smaller than would be expected from having 211 independent parameters, reflecting the fact that land models like CLM comprise representations of numerous interconnected domains.
 While the principle of land models is that these are all interlinked via complex feedback processes, many parameters nonetheless have impacts that are mostly limited to their own domain.
@@
 For example, approximately 96\% of canopy evaporation variance is explained by the first ten parameters regardless of forcing, whereas between 74 and 92\% of GPP variance is explained by the first ten parameters.
-Only the top 50 parameters are shown (of 193) to improve the visualization.
+Only the top 50 parameters are shown (of 211) to improve the visualization.
 Note that parameter rankings differ by variable.}
@@
 
-In this project, we identified and perturbed 193 CLM parameters to create a large one-at-a-time PPE.
+In this project, we identified and perturbed 211 CLM parameters to create a large one-at-a-time PPE.
 This ensemble is useful for understanding parametric controls on CLM processes.
@@
 In such cases, we identified hard-coded values, established an appropriate parameter name, and extracted that parameter to the CLM parameter file for easier manipulation.
-Although in this experiment we perturbed a large number of parameters (193) across a variety of CLM processes (Figure \ref{fig:params}), this still does not cover the full set of CLM parameters, as some processes were not included, such as crop phenology and management.
+Although in this experiment we perturbed a large number of parameters (211) across a variety of CLM processes (Figure \ref{fig:params}), this still does not cover the full set of CLM parameters, as some processes were not included, such as crop phenology and management.
 In carrying out this process of parameter identification, we  likewise unearthed many nuances in  the epistemology of what  `parameters' are in the context of Earth system models.
@@
 In the case of canopy evaporation, for example, more than 95\% of the ensemble variance could be explained by the ten most influential parameters (Figure \ref{fig:variance}).
-This indicates that there may be tractable parameter estimation sub-problems if global calibration of 193 parameters proves unattainable.
+This indicates that there may be tractable parameter estimation sub-problems if global calibration of 211 parameters proves unattainable.
 

### 2024-04-04 ms/ms.tex

@@
 \centering
-\includegraphics[width=\textwidth]{../figs/variance.pdf}
-\caption{Cumulative fraction of variance explained by the most influential parameters on gross primary production and canopy evaporation.
-For example, approximately 96\% of canopy evaporation variance is explained by the first ten parameters regardless of forcing, whereas between 74 and 92\% of GPP variance is explained by the first ten parameters.
-Only the top 50 parameters are shown (of 211) to improve the visualization.
-Note that parameter rankings differ by variable.}
+\includegraphics[width=\textwidth]{../figs/variance.png}
+\caption{Cumulative fraction of variance explained by the most influential parameters on GPP and canopy evaporation.
+Approximately 96\% of canopy evaporation variance can be explained by the ten most influential parameters, whereas 30 parameters are required to explain 96\% of GPP variance.}
 \label{fig:variance}
@@
 These parameters, while not especially influential on present-day ET, would be among the most influential in governing the trend in ET going forward, illustrating the potential limitations of calibration to ambient conditions.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\textwidth]{../figs/ET_response.png}
-\caption{(a) Global annual average evapotranspiration (ET) in the future climate scenario vs.
-our control experiment.
-The shading spans the default ET in the control experiment, $\pm$1 W/m2.
-The response of ET to future climate with default parameters is +8.8\%.
-The orange line shows the expected future ET assuming a linear default ET response.
-(b) The top 8 parameters governing global ET in the control experiment.
-(c) The top 8 parameters governing ET response to future climate, i.e. the ratio of future ET to control ET.
-Parameters governing the change in ET expected in the future tend to differ from the parameters controlling present-day ET.}
-\label{fig:et}
-\end{figure}
 
@@
 Likewise figures are repeated for each of the various forcing scenarios.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/main/NEP_response.png}
+\caption{NEP response}
+\label{fig:nep}
+\end{figure}
 

### 2024-04-05 ms/ms.tex
ms edits

@@
 \usepackage{pdflscape}
+
+\usepackage{xr}
+\externaldocument{suppinfo}
+
 \linenumbers
@@
 Understanding how these processes respond to and influence CO$_2$ concentrations is a critical facet of climate change research.
-Our certainty in climate model projections varies by domain and generally decreases with extended time horizons \cite{koven2022}.
+Our uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}.
 Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}.
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
-Effectively synthesizing the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns can help to constrain increasingly comprehensive land modeling platforms.
+Our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns.
+However, given increasingly comprehensive land modeling systems, and the wide array of observational products, several technical hurdles exist that hinder effective model development.
 
-Uncertainty in the model projections of the carbon cycle is manifest in uncertainty around the appropriate algorithms and equations used, and also in the parameterizations of these models \cite{luo2020}.
-Model inter-comparison projects (MIPs) are the primary means by which structural uncertainty is assessed \cite{henderson-sellers1995,pitman1999,wood1998,schlosser2000,eyring2016,friedlingstein2022}.
+Model inter-comparison projects (MIPs) are a major component of the model development cycle, and have been the primary means by which model projection uncertainty is assessed \cite{henderson-sellers1995,pitman1999,wood1998,schlosser2000,eyring2016,friedlingstein2022}.
 While these have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
-To allow for more tractable experimental design, in most MIPs, each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the cost of conducting multiple experiments with different parameterizations, e.g.
-the TRENDY land model intercomparison or the coupled model intercomparison project \cite{friedlingstein2022,eyring2016}.
+In most MIPs, each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the high cost of each simulation, e.g.
+the TRENDY land model intercomparison or the Coupled Model Intercomparison Project \cite{friedlingstein2022,eyring2016}.
 Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land state trajectories.
 
-Parameter sensitivity tests can be used to gauge parametric uncertainty (e.g. need refs).
-A collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with many examples in the coupled (refs) and land-only contexts (huang2018) \cite{dagon2020}.
+Understanding the range of outcomes arising from the various plausible parameter sets is a critical step in robust uncertainty quantification. 
+Parameter sensitivity tests can be used to gauge parametric uncertainty, and
+a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with many examples in the coupled (refs) and land-only contexts (huang2018) \cite{dagon2020}.
 PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.
-Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large parameter perturbation ensembles (PPEs) \cite{qian2018}.
-In this project we create a large PPE with the Community Terrestrial Systems Model (CTSM) \cite{lawrence2019}, in the Community Land Model with biogeochemistry configuration (CLM-BGC).
-We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
-Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization.
 
-One goal of this project is to enable automated calibration of CLM-BGC.
+Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.
+Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large PPEs \cite{qian2018}.
+In a more general sense, PPEs yield a robust knowledge basis for understanding and working with the given model. 
+This can aid in introducing new users to the model, or to help steer and facilitate model development.
+
 The process of parameter estimation for land models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryosphere, biogeochemistry, physiology, land management, etc.).
@@
 
-Maintaining, improving, and interpreting complex land models benefits from thoughtful investment in software to automate and routinize important components of the development process, e.g.
-\citeA{collier2018}.
+In this project we create a large PPE with the Community Terrestrial Systems Model (CTSM) \cite{lawrence2019}, in the Community Land Model with biogeochemistry configuration (CLM-BGC).
+We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
+Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization.
 Our goal in this project is to systematize the parameter perturbation process within the CTSM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across a wide range of land model processes.
@@
 Last Glacial Maximum).
+
+We constrained the ranges of several parameters based on historical survivability (Figure \ref{abcdefg}).
 
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=42pc]{../figs/supp/ilamb.pdf}
-\caption{ILAMB 2.5 overall scores comparing remapped sparsegrid output to the full grid 2$^{\circ}$ model output.
-Column headings indicate the number of clusters, and whether annual means or annual means and standard deviations were used as input to the clustering algorithm.
-We should try to remove the gray bits (zqz).
-Full, interactive results are available at \url{www.ilamb.org/PPE/CLM/2021-02}.
-See main text Section \ref{sect:sg} for clustering details.}
-\label{supp:ilamb}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\textwidth]{../figs/supp/FCEV_global_mean.png}
-\caption{The top ten parameters controlling global annual canopy evaporation (CLM variable: FCEV) across the six forcing scenarios.
-Note that this plot is one of many from our more comprehensive diagnostics set \url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT/}.}
-\label{supp:fcev}
-\end{figure}
+extra text:
+Maintaining, improving, and interpreting complex land models benefits from thoughtful investment in software to automate and routinize important components of the development process, e.g.
+\citeA{collier2018}.
 

### 2024-04-09 ms/ms.tex

@@
 \usepackage{xr}
-\externaldocument{suppinfo}
+\externaldocument{ms/supp}
 
@@
 
-We constrained the ranges of several parameters based on historical survivability (Figure \ref{abcdefg}).
+We constrained the ranges of several parameters based on historical survivability (Figure \ref{supp:surv}).
 
@@
 Given the challenging size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the need to probe and document parametric uncertainties within operational land surface models, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
+Combining the SASU spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
+There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
+Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for our experiment within 140 years.
 
@@
 \end{figure}
-
-Combining the SASU spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times cheaper than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
-There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
-Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for our experiment within 140 years.
 

### 2024-04-09 ms/supp.tex

\documentclass[11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[parfill]{parskip}
\usepackage[margin=1in]{geometry}
\title{Supplementary information for ``CLM5-PPE"}
\author{D. Kennedy}
\renewcommand{\thefigure}{S\arabic{figure}}
\begin{document}
\maketitle

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figs/supp/survivability_c285.png}
\caption{Survivability.}
\label{supp:surv}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figs/supp/survivability_c285.png}
\caption{Survivability.}
\label{supp:surv}
\end{figure}

\end{document}

### 2024-04-09 ms/supp.tex
Updates from Overleaf

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{pdflscape}
 \title{Supplementary information for ``CLM5-PPE"}
@@
 
+\begin{landscape}
 \begin{figure}[h]
 \centering
-\includegraphics[width=\textwidth]{figs/supp/survivability_c285.png}
-\caption{Survivability.}
-\label{supp:surv}
+\includegraphics[width=60pc]{figs/supp/ilamb_lines.png}
+\caption{The convergence of several scoring metrics with increasing sparsegrid resolution across a variety of CLM variables.
+All scores tend towards 1 as the number of sparsegrid clusters (k) approaches the 5666 gridcells for the native resolution of our 2-degree simulations.
+}
+\label{supp:ilamb}
 \end{figure}
+\end{landscape}
 

### 2024-04-09 ms/ms.tex

@@
 Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity.
-Given more clusters, the sparsegrid provides a better approximation of output from the full grid (Figure \ref{fig:sg}).
-Accurate reconstruction of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Supp Figure zqz).
+Given more clusters, the sparsegrid will generally provide a better approximation of output from the full grid.
+As one example, accurate reconstruction of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Supp Figure zqz).
 We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})

### 2024-05-22 ms/ms.tex

@@
 Understanding how these processes respond to and influence CO$_2$ concentrations is a critical facet of climate change research.
-Our uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}.
+Uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}.
 Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}.
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
-Our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns.
+Still, our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns.
 However, given increasingly comprehensive land modeling systems, and the wide array of observational products, several technical hurdles exist that hinder effective model development.
@@
 Model inter-comparison projects (MIPs) are a major component of the model development cycle, and have been the primary means by which model projection uncertainty is assessed \cite{henderson-sellers1995,pitman1999,wood1998,schlosser2000,eyring2016,friedlingstein2022}.
-While these have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
+While MIPs have had tremendous utility in capturing and assessing wide ranges of model assumptions, it can nonetheless be difficult to interpret the differences between models, or even between subsequent versions of the same model, due to the multiplicity of structural and parametric variations \cite{mcneall2016}.
 In most MIPs, each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the high cost of each simulation, e.g.
@@
 
-The process of parameter estimation for land models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryosphere, biogeochemistry, physiology, land management, etc.).
-The number of model parameters required to represent all the relevant processes is large, and model speeds are relatively slow, presenting a major barrier to robust objective calibration, model interpretation and development \cite{fisher2020,dagon2020}.
+The process of parameter estimation for the complex land models typically embedded within Earth System Model is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryosphere, biogeochemistry, physiology, land management, etc.).
+The number of model parameters required to represent all the relevant processes is large, and model simulations are relatively expensive, presenting a major barrier to robust objective calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
 Thorough assessment of the parametric sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}.
 It typically requires extensive computational time,  software engineering, and domain-specific expert scientific knowledge.
-Our work here tries to alleviate each of these challenges in turn, at least within the context of the CLM-BGC model.
+Our work here tries to alleviate each of these challenges in turn, at least within the context of one land model.
 While we are not yet at the point of automated calibration, we present herein several innovations that will serve to make an automated calibration system more technically feasible.
@@
 Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization.
-Our goal in this project is to systematize the parameter perturbation process within the CTSM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across a wide range of land model processes.
+The goal of this project is to systematize the parameter perturbation process within the CTSM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across a wide range of land model processes.
 In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
@@
 2022).
-Our spin-up protocol featured 20 years in Accelerated Decomposition (AD) spin-up mode (see \citeA{lawrence2019} (Thornton and Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (i.e.
+The spin-up protocol featured 20 years in Accelerated Decomposition (AD) spin-up mode (see \citeA{lawrence2019} (Thornton and Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (i.e.
 ND), cycling over a ten-year forcing dataset (described below, Liao et al.
@@
 We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.
-Our future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, forcing in order to allow comparison with the existing CESM2 large ensemble.
+The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, forcing in order to allow comparison with the existing CESM2 large ensemble.
 Our software infrastructure has greatly reduced the effort and computational cost associated with ensemble generation, such that it would be easy to add additional ensembles, to explore other emission scenarios or climate regimes (e.g.
@@
 
-Given the challenging size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the need to probe and document parametric uncertainties within operational land surface models, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
-Combining the SASU spinup approach with our sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
-There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
-Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for our experiment within 140 years.
+Given the very large size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the desire to probe and document parametric uncertainties within operational land surface models, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
+Combining the SASU spinup approach with the sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
+There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM5 simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
+Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for this experiment within 140 years.
 
@@
 Given more clusters, the sparsegrid will generally provide a better approximation of output from the full grid.
-As one example, accurate reconstruction of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Supp Figure zqz).
-We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
+As one example, accurate reconstruction of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters (Figure \ref{fig:sg}).
+Performance continues to improve with added clusters, but with marginal returns.
+We had hoped to see a clear point of diminishing returns across a wide range of variables as we increase the number of clusters.
+Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, convergence slows and presents some small amplitude oscillations (Supp Figure S2).
+We chose 400 clusters, because it provides satisfactory performance across a range of important metrics, while affording us sufficient simulations to perform the full experiment within our computational budget.
 \begin{figure}[h]
 \centering
-\includegraphics[width=25pc]{../figs/sparsegrid_gpp.pdf}
+\includegraphics[width=25pc]{../figs/main/sparsegrid_gpp.pdf}
 \caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last forty years of a transient CLM5.1 simulation.
@@
 
-Overall, we found substantial impacts of parametric uncertainty on model behaviour, in some cases exceeding the magnitude of climate scenario effects (Figure \ref{fig:ranges}).
+This model configuration allowed for over two thousand simulations across six forcing scenarios.
+Overall, we found substantial impacts of parametric uncertainty on model behavior, in some cases exceeding the magnitude of climate scenario effects (Figure \ref{fig:ranges}).
+In the control scenario, gross primary productivity (GPP) ranged from 103 to 158 PgC, and net ecosystem productivity (NEP) ranged from 1.7 to 3.8 PgC.
+NEP was especially variable in the high CO$_2$ scenario, varying from 1.6 to 5.7 PgC.
 One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death.
-A perturbation of $\pm$20\% likely exceeds the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils.
+A perturbation of $\pm$20\% may exceed the reasonable range for this parameter, but this simulation was instructive for exposing the model's response to hot soils.
 Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, including unforeseen tipping points, brittle parameterizations,  and/or bugs.
@@
 A small fraction of parameters tend to explain a large amount of the ensemble variance (Figure \ref{fig:variance}) for any given variable or metric.
-For example, the same four parameters explain up to 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig \ref{supp:fcev}).
-In general,most of the parameter perturbations had a relatively small impact on any given output variable.
-As such, the effective parameter space dimensionality for any given CLM process is likely much smaller than would be expected from having 211 independent parameters, reflecting the fact that land models like CLM comprise representations of numerous interconnected domains.
+For example, just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S3).
+Conversely, GPP, which is a significantly more complicated process, requires 20 parameters to explain 90\% of the control ensemble variance.
+In general, most of the parameter perturbations had a relatively small impact on any given output variable.
+Though this experiment encompasses 211 parameters, no single variable responds to all 211 parameters.
+In fact, the effective parameter space dimensionality for any given output variable tends to be much smaller than 211.
 While the principle of land models is that these are all interlinked via complex feedback processes, many parameters nonetheless have impacts that are mostly limited to their own domain.
@@
 
-We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps}.
-This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control responses to individual forcing variables (CO2, nitrogen deposition) and climate.
-In the case of evapotranspiration (ET), for example, the parameters that control the increase in ET due to warming tend to be different from the parameters that control present-day ET (Figure \ref{fig:et}).
-ET response can differ significantly from the default case (i.e. stray from the orange line), even when ET is near the default under present-day conditions (i.e. within the shaded region).
-Besides \textit{kmax}, all of the parameters that appear in the top eight parameters for ET response to future climate govern plant acclimation to temperature. 
-These parameters, while not especially influential on present-day ET, would be among the most influential in governing the trend in ET going forward, illustrating the potential limitations of calibration to ambient conditions.
-
-The PPE is likewise useful for determining which parameters influence which model processes and where.
+This PPE is especially useful for determining which parameters influence which model processes and where.
 For example the parameters controlling leaf area index vary significantly by biome (Figure \ref{fig:lai}).
 Plant hydraulics parameters were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
-The most influential parameters for leaf area index globally include parameters that were important in each of these three biomes.
+The most influential parameters for leaf area index globally included parameters that were important in each of these three biomes.
+
+We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps}.
+This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control the response to forcing (CO2, nitrogen deposition) and climate.
+For example, the parameters that control present-day NEP differ  from the parameters that control NEP in the high CO$_2$ scenario (Figure \ref{fig:nep}a,b).
+NEP increases by approximately 50\% at 867 ppm with the default parameters, but can actually decrease with certain parameter settings (Figure \ref{fig:nep}c).
+One key parameter is \textit{tpuse\_sf}, which is a scalar perturbation factor influencing the triose phosphate limitations on photosynthesis.
+Triose phosphate limitation is difficult to quantify and has not been observed to significantly hinder photosynthesis under present-day CO$_2$ concentrations (cite Kumarathunge et al).
+In CLM5.1, this parameter is not particularly influential on NEP at 367 ppm (ranked 39, not shown), but is third most influential with high CO$_2$.
+
+The future climate experiment is 3K(zqz) warmer averaged over our study domain (land-only, Antarctica excluded, Supp Figure S4).
+This led to an increase in evapotranspiration of 8.8\% relative to the control simulation when using the default CLM5.1 parameters (Figure \ref{fig:nep}f).
+While the parameters that most strongly affect global average ET in both the control and future climates are generally the same (Figure \ref{fig:nep}d,e), a distinct set of parameters govern the change in ET due to warming (Figure \ref{fig:nep}f).
+While the control simulation is primarily influenced by hydrology and stomatal conductance parameters, the response of ET to warming is largely controlled by photosynthesis acclimation parameters (i.e.
+\textit{tpuse\_sf}, \textit{kcha}, \textit{vcmaxha}, \textit{vcmaxhd}, \textit{lmrhd}, and \textit{cpha}).
+The scale of ET response may be inflated, given that these are land-only experiments, and an interactive atmosphere could modulate the ET response.
 
@@
 Because the number of potential variables, parameters, and geographical ranges of interest to the wider CLM community is larger than we can document here,  we provide a tool that can be used to explore an extended diagnostics set which summarises the $>$2TB of output data via approximately 2000 plots (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
-
 The diagnostics website includes ranking plots (as in Figure \ref{fig:lai}) and maps of parameter effects (as in Figure \ref{fig:panel}).

### 2024-05-22 ms/ms.tex
Updates from Overleaf

@@
 \includegraphics[width=\textwidth]{../figs/lai_biome.pdf}
-\caption{The eight most influential parameters on leaf area index within the control ensemble, globally and within three biomes.}
+\caption{The eight most influential parameters on leaf area index within the control ensemble, globally and within three biomes.
+The solid lines span the range between the two simulations for each parameter, with an open circle indicating the simulation where a parameter is set to its minimum value and a filled circle for its maximum.
+The lines are color-coded according to scientific domain.
+The dashed lines indicate the outcomes for the simulation with default CLM5.1 parameters.
+Parameter rankings vary by biome, with the global rankings seeming to reflect contributions from each of these three biomes.}
 \label{fig:lai}
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/main/NEP_response.png}
-\caption{NEP response}
+\includegraphics[width=\textwidth]{../figs/main/deltas.png}
+\caption{The eight most influential parameters on NEP in the control simulations (a), high CO$_2$ simulations (b), and the relative response to high CO$_2$ (c), as well as on ET in the control simulations (d), future climate simulations (e) and the relative response to future climate (f).
+The parameters that are most influential on present-day climate (a,d) can differ significantly from the parameters that control the response to forcing perturbations (c,f).}
 \label{fig:nep}

### 2024-05-24 ms/ms.tex

@@
 
-We constrained the ranges of several parameters based on historical survivability (Figure \ref{supp:surv}).
+For a small percentage of parameters, the perturbation ranges were severe enough to eradicate a given PFT across all or part of its geographical range (Supp Figure S1). 
+In the case that one or more PFTs did not survive a parameter perturbation, we iteratively re-ran the simulation with slightly less extreme parameter values until the PFT survived.
+A PFT is classified as surviving in a given gridcell if it achieves an LAI of at least 0.1 m2/m2 at any point during the production portion of the simulation (i.e. post-spinup).
+A PFT is classified as surviving for a given parameter perturbation if at least 50\% of the area that
+was alive with the default parameters is alive in the perturbed simulation.
+For the analyses in this paper, we constrained the ranges of several parameters based on survivability in the low CO$_2$, pre-industrial climate, and control climate ensembles to improve the chances that all PFTs would be able to survive to present-day in a transient 1850 to present-day simulation.
+We did not constrain parameter ranges based on future climate, high CO$_2
+$, or increased nitrogen deposition.
+Original and constrained ranges are available in the supplemental materials spreadsheet (zqz).
 
@@
 
-The future climate experiment is 3K(zqz) warmer averaged over our study domain (land-only, Antarctica excluded, Supp Figure S4).
+The future climate experiment is 4.3K warmer averaged over our study domain (land-only, Antarctica excluded, Supp Figure S4).
 This led to an increase in evapotranspiration of 8.8\% relative to the control simulation when using the default CLM5.1 parameters (Figure \ref{fig:nep}f).
@@
 \textit{tpuse\_sf}, \textit{kcha}, \textit{vcmaxha}, \textit{vcmaxhd}, \textit{lmrhd}, and \textit{cpha}).
-The scale of ET response may be inflated, given that these are land-only experiments, and an interactive atmosphere could modulate the ET response.
+In a coupled ESM, land-atmosphere interactions would modulate these parameter effects, such that the parameter rankings are likely more credible than the absolute values of the parameter effects.
 
@@
 
+In this paper, we focus primarily on global and biome-level parameter rankings.
+However it is also possible to inspect the geographic footprint of parameter perturbations by projecting the sparsegrid output to standard lat/lon coordinates (see Section \ref{sect:sg} for details).
+For example, perturbing the \textit{medlynslope} parameter has a large effect on global runoff, but primarily via its effects in vegetated areas (Figure \ref{fig:map}).
 Because the number of potential variables, parameters, and geographical ranges of interest to the wider CLM community is larger than we can document here,  we provide a tool that can be used to explore an extended diagnostics set which summarises the $>$2TB of output data via approximately 2000 plots (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
-The diagnostics website includes ranking plots (as in Figure \ref{fig:lai}) and maps of parameter effects (as in Figure \ref{fig:panel}).
-But these plots are repeated across a combination of model output variables and model parameters.
+The diagnostics website includes ranking plots (as in Figure \ref{fig:lai}) and maps of parameter effects (as in Figure \ref{fig:map}).
+These plots are repeated across a combination of model output variables and model parameters.
 Likewise figures are repeated for each of the various forcing scenarios.
@@
 }
-\label{fig:panel}
+\label{fig:map}
 \end{figure}

### 2024-06-10 ms/supp.tex

@@
 \includegraphics[width=\textwidth]{figs/supp/survivability_c285.png}
-\caption{Survivability.}
+\caption{Survivability in the low CO$_2$ ensemble.
+The sixteen natural vegetation PFTs are on the vertical axis, and the perturbation number sets the horizontal axis.
+A PFT is classified as dead in a given gridcell if it does not achieve an LAI of at least 0.1 m2/m2 at any point during the simulation.
+A PFT is classified as dead for a given perturbation if more than 30\% of the area that was alive with the default parameters is dead in the perturbed simulation.}
 \label{supp:surv}
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{figs/supp/FCEV_global_mean.png}
+\caption{Global mean canopy evaporation (FCEV) parameter rankings across the six forcing scenarios.}
+\label{supp:fcev}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{figs/supp/anomalies.png}
+\caption{Maps of average forcing anomalies in the pre-industrial (left column) and future climate forcing (right column) simulations.
+Global averages over land, Antarctica excluded, are written in text on each subplot.}
+\label{supp:anomalies}
+\end{figure}
+
 \end{document}

### 2024-06-10 ms/supp.tex
Updates from Overleaf

@@
 
+\begin{landscape}
 \begin{figure}[h]
 \centering
-\includegraphics[width=\textwidth]{figs/supp/survivability_c285.png}
-\caption{Survivability in the low CO$_2$ ensemble.
+\includegraphics[width=50pc]{figs/supp/survivability.png}
+\caption{PFT survivability across the six ensembles.
 The sixteen natural vegetation PFTs are on the vertical axis, and the perturbation number sets the horizontal axis.
 A PFT is classified as dead in a given gridcell if it does not achieve an LAI of at least 0.1 m2/m2 at any point during the simulation.
-A PFT is classified as dead for a given perturbation if more than 30\% of the area that was alive with the default parameters is dead in the perturbed simulation.}
+A PFT is classified as dead for a given perturbation if more than 50\% of the area that was alive with the default parameters is dead in the perturbed simulation.}
 \label{supp:surv}
@@
 
-\begin{landscape}
 \begin{figure}[h]

### 2024-06-10 ms/ms.tex

@@
 The crop model was not utilized as our initial focus is on unmanaged land.
+Of the 211 parameters, 145 are scalar-valued, with a single parameter value acting globally.
+The remaining 66 parameters include a plant functional type (PFT) dimension, with the potential for different parameter values for each unique plant functional type.
+Not all parameters are currently utilizing this functionality, such as \textit{medlynintercept} where the default value is 100 $\mu$mol H$_2$O for all sixteen of CLM's natural vegetation PFTs.
 
@@
 In the case that one or more PFTs did not survive a parameter perturbation, we iteratively re-ran the simulation with slightly less extreme parameter values until the PFT survived.
-A PFT is classified as surviving in a given gridcell if it achieves an LAI of at least 0.1 m2/m2 at any point during the production portion of the simulation (i.e. post-spinup).
-A PFT is classified as surviving for a given parameter perturbation if at least 50\% of the area that
+A PFT was classified as surviving in a given gridcell if it achieves an LAI of at least 0.1 m2/m2 at any point during the production portion of the simulation (i.e. post-spinup).
+A PFT was classified as surviving for a given parameter perturbation if at least 50\% of the area that
 was alive with the default parameters is alive in the perturbed simulation.
@@
 $, or increased nitrogen deposition.
-Original and constrained ranges are available in the supplemental materials spreadsheet (zqz).
+In total 16 of the 211 parameters had the parameter range constrained to some extent. Original and constrained ranges are available in the supplemental materials spreadsheet (zqz).
 

### 2024-07-15 ms/ms.tex
overleaf

@@
 \begin{abstract}
-[ enter your Abstract here ]
+\begin{itemize}
+\item 211 parameters perturbed
+\item featherweight clm configuration, 500x less costly
+\item parameter effects can exceed scenario effects
+\item large dataset available online providing parameter ranges and effect sizes useful for future studies
+\item software infrastructure developed to facilitate routine investigation of parameter sensitivity and uncertainty, as well as automated calibration
+\item small number of parameters explains a large fraction of variance 
+\item most important parameters can vary regionally and also based on climate forcing
+
+\end{itemize}
 \end{abstract}
@@
 
-Given the very large size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the desire to probe and document parametric uncertainties within operational land surface models, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations given the available computational resources.
+Given the very large size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the relatively expensive computational cost of standard model configurations, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations.
 Combining the SASU spinup approach with the sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
@@
 As one example, accurate reconstruction of global photosynthesis could be achieved with a relatively small number of clusters, with R$^2>$0.95 achieved with only 200 clusters (Figure \ref{fig:sg}).
-Performance continues to improve with added clusters, but with marginal returns.
+Performance continues to improve with added clusters, but with marginal returns (Supp Figure S2).
 We had hoped to see a clear point of diminishing returns across a wide range of variables as we increase the number of clusters.
-Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, convergence slows and presents some small amplitude oscillations (Supp Figure S2).
+Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, convergence slows and presents some small amplitude oscillations.
 We chose 400 clusters, because it provides satisfactory performance across a range of important metrics, while affording us sufficient simulations to perform the full experiment within our computational budget.
@@
 
-This model configuration allowed for over two thousand simulations across six forcing scenarios.
+This model configuration allowed for over 2000 simulations across six forcing scenarios.
 Overall, we found substantial impacts of parametric uncertainty on model behavior, in some cases exceeding the magnitude of climate scenario effects (Figure \ref{fig:ranges}).
@@
 \textit{tpuse\_sf}, \textit{kcha}, \textit{vcmaxha}, \textit{vcmaxhd}, \textit{lmrhd}, and \textit{cpha}).
-In a coupled ESM, land-atmosphere interactions would modulate these parameter effects, such that the parameter rankings are likely more credible than the absolute values of the parameter effects.
+In a coupled ESM, land-atmosphere interactions would modulate these parameter effects, such that the parameter rankings are likely more robust than the absolute values of the parameter effects.
 

### 2024-08-06 ms/ms.tex

@@
 
-Understanding the range of outcomes arising from the various plausible parameter sets is a critical step in robust uncertainty quantification. 
+Understanding the range of outcomes arising from the many plausible parameter combinations is a critical step in robust uncertainty quantification. 
 Parameter sensitivity tests can be used to gauge parametric uncertainty, and
-a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with many examples in the coupled (refs) and land-only contexts (huang2018) \cite{dagon2020}.
-PPEs have played an important role in assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-
-Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian uncertainty quantification \cite{cleary2021}.
+a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled (refs) and land-only contexts (huang2018) \cite{dagon2020}.
+One important application of PPEs has been assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
+Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian methods \cite{cleary2021}.
 Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large PPEs \cite{qian2018}.
-In a more general sense, PPEs yield a robust knowledge basis for understanding and working with the given model. 
+In a more general sense, PPEs yield a robust knowledge basis for understanding and working with a given climate model. 
 This can aid in introducing new users to the model, or to help steer and facilitate model development.
 
-The process of parameter estimation for the complex land models typically embedded within Earth System Model is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, cryosphere, biogeochemistry, physiology, land management, etc.).
+The process of parameter estimation for the complex land models typically embedded within Earth System Models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, snow and ice sheets, biogeochemistry, physiology, land management, etc.).
 The number of model parameters required to represent all the relevant processes is large, and model simulations are relatively expensive, presenting a major barrier to robust objective calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
@@
 In this paper we present that dataset, describe how it was produced, and survey some potential applications.
-The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CLM.
+The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CTSM/CLM.
 
@@
 
-To manage computational cost we leveraged the Semi-Analytic spin-up mode (i.e., SASU) recently implemented within CLM5 \cite{lu2020} (Liao et al.
+To manage computational cost we leveraged the Semi-Analytic spin-up mode (SASU) recently implemented within CLM5 \cite{lu2020} (Liao et al.
 2023).
@@
 2022).
-The spin-up protocol featured 20 years in Accelerated Decomposition (AD) spin-up mode (see \citeA{lawrence2019} (Thornton and Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (i.e.
-ND), cycling over a ten-year forcing dataset (described below, Liao et al.
+The spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (see \citeA{lawrence2019} (Thornton and Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode, cycling over a ten-year forcing dataset (described below, Liao et al.
 2023).
@@
 The lines are color-coded according to scientific domain.
-The dashed lines indicate the outcomes for the simulation with default CLM5.1 parameters.
+The dashed lines indicate the LAI for the simulation with default CLM5.1 parameters.
 Parameter rankings vary by biome, with the global rankings seeming to reflect contributions from each of these three biomes.}

### 2024-08-06 ms/ms.tex
Updates from Overleaf

@@
 Still, our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns.
-However, given increasingly comprehensive land modeling systems, and the wide array of observational products, several technical hurdles exist that hinder effective model development.
+The capability to efficiently ingest these data and improve simulation performance is especially valuable for actionable science \cite{cheng2023}.
+However, given increasingly comprehensive land modeling systems, and the wide array of observational products, several technical hurdles exist that hinder effective model development and calibration.
 
@@
 In most MIPs, each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the high cost of each simulation, e.g.
-the TRENDY land model intercomparison or the Coupled Model Intercomparison Project \cite{friedlingstein2022,eyring2016}.
+the TRENDY land model intercomparison (Sitch2024) or the Coupled Model Intercomparison Project \cite{eyring2016}.
 Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land state trajectories.
@@
 Parameter sensitivity tests can be used to gauge parametric uncertainty, and
-a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled (refs) and land-only contexts (huang2018) \cite{dagon2020}.
-One important application of PPEs has been assessing projection uncertainty in climate models \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
+a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \cite{dagon2020}.
+One important application of PPEs has been assessing uncertainty in climate model projections \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
 Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian methods \cite{cleary2021}.

### 2024-08-06 ms/supp.tex
overleaf
supp info extended

@@
 \end{figure}
+\end{landscape}
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{figs/supp/biome_tp.png}
+\caption{Each of the 400 sparse gridcells plotted according to their Whittaker biomes}
+\label{supp:whit1}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{figs/supp/biome_latlon.png}
+\caption{Cartesian plot of the 400 sparse gridcells labeled by their Whittaker biomes}
+\label{supp:whit2}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=23pc]{figs/supp/biome_areas.png}
+\caption{Total represented land area associated with each of the ten Whittaker biomes utilizing the CLM5.1 sparse grid}
+\label{supp:whit3}
+\end{figure}
+
+\begin{landscape}
 \begin{figure}[h]
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{figs/supp/anomaly_errors.png}
+\caption{buggy climate forcing}
+\label{supp:abug}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=30pc]{figs/supp/AF2095_rankings.png}
+\caption{buggy climate forcing}
+\label{supp:abug}
+\end{figure}
+
 \end{document}

### 2024-08-12 ms/ms.tex
adding refs

@@
 Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}.
+Something about water \cite{zarakas2024}.
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
@@
 In most MIPs, each model is typically allowed only a single parameterization (despite the existence of many plausible parameter sets) due to the high cost of each simulation, e.g.
-the TRENDY land model intercomparison (Sitch2024) or the Coupled Model Intercomparison Project \cite{eyring2016}.
+the TRENDY land model intercomparison \cite{sitch2024} or the Coupled Model Intercomparison Project \cite{eyring2016}.
 Thus, MIPs typically conflate parametric and structural uncertainty, and de-emphasize the consequences of uncertain model calibration in estimates of future land state trajectories.
@@
 
- A parameter sensitivity analysis was conducted with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) using a semi-analytic spin-up (SASU) offered by a matrix approach to understand 34 biogeochemistry parameters according to variance explained by each parameter (Huang et al.
-2018).
+ A parameter sensitivity analysis was conducted with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) using a semi-analytic spin-up (SASU) offered by a matrix approach to understand 34 biogeochemistry parameters according to variance explained by each parameter \cite{huang2018}.
 
@@
 
-Biome analysis
+Biome analysis (Supp Figures S2-4)
 

### 2024-08-12 ms/ms.tex
overleaf

@@
 Uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}.
-Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}.
-Something about water \cite{zarakas2024}.
+Land processes substantially influence climate directly through, for example, evapotranspiration \cite{zarakas2024}, and indirectly through carbon-climate feedbacks.
+Making centennial-scale projections of the cumulative terrestrial carbon sink gas been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}, and limited efficacy of emergent constraints.
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
@@
 In this paper we present that dataset, describe how it was produced, and survey some potential applications.
-The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CTSM/CLM.
+The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b,zarakas2024}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CTSM/CLM.
 

### 2024-08-12 ms/ms.tex
overleaf

@@
 
-A small fraction of parameters tend to explain a large amount of the ensemble variance (Figure \ref{fig:variance}) for any given variable or metric.
+A small fraction of parameters tend to explain a large amount of the ensemble variance for any given variable or metric (Figure \ref{fig:variance}).
 For example, just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S3).
@@
 2023) and a linearized spin-up solver (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
-As long as computational resources remain constrained, designing faster model configurations and more efficient sampling strategies will be important for effective model calibration and uncertainty quantification.
+As long as computational resources remain constrained, designing faster model configurations and efficient sampling strategies will be important for effective model calibration and uncertainty quantification.
 
@@
 
-Land only caveat?
 Our PPE is run as land-only simulations, using prescribed atmospheric forcing.
@@
 2019).
-Additional work is planned to leverage this one-at-a-time land-only PPE to quantify how much, where, and for which processes atmospheric feedbacks are likely to change the land surface state substantially.
+Additional work is ongoing to extend this work in a coupled model framework to quantify how much, where, and for which processes atmospheric feedbacks are likely to modulate the land parameter effects substantially.
 
@@
 We have likewise begun a follow-on activity that perturbs a subset of important parameters with the Latin hypercube sampling strategy \cite{mckay2000} to work towards routine model calibration.
-Our first foray into exploring the potential for global calibration is with leaf area index due to controls that this variable has on biogeophysical and biogeochemical fluxes.
 By investing in the infrastructure that we introduce here, all of our subsequent parameter perturbation experiments have required much less time and effort.
-We expect to continue to extend our efforts in this domain towards model calibration, as well as repeat the foundational one-at-a-time experiments with subsequent model releases.
+We expect to continue to extend our efforts in this domain towards model calibration and uncertainty quantification, as well as repeating the foundational one-at-a-time experiments with subsequent model releases.
 

### 2024-08-12 ms/ms.tex
more refs

@@
 
-To manage computational cost we leveraged the Semi-Analytic spin-up mode (SASU) recently implemented within CLM5 \cite{lu2020} (Liao et al.
-2023).
-This new module utilizes a matrix representation on CLM's biogeochemistry to significantly reduce spin-up time by one order of magnitude (Liao et al.
-2023, Luo et al.
-2022).
-The spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (see \citeA{lawrence2019} (Thornton and Rosenbloom, 2005) for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode, cycling over a ten-year forcing dataset (described below, Liao et al.
-2023).
+To manage computational cost we leveraged the Semi-Analytic spin-up mode (SASU) recently implemented within CLM5 \cite{lu2020,liao2023}.
+This new module utilizes a matrix representation on CLM's biogeochemistry to significantly reduce spin-up time by one order of magnitude \cite{luo2022,liao2023}, and had been previously leveraged for a PPE with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) model \cite{huang2018}.
+Our spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (see \citeA{lawrence2019} and \citeA{thornton2005} for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode, each cycling over a ten-year forcing dataset (described below).
 This protocol was designed to achieve sufficiently equilibrated carbon and nitrogen states, while minimizing computational time.
 This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter depth) in very few points at high latitude.
-Certain inferences about deep soil carbon from this ensemble are therefore subject to uncertainty due to spin-up concerns.
-
- A parameter sensitivity analysis was conducted with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) using a semi-analytic spin-up (SASU) offered by a matrix approach to understand 34 biogeochemistry parameters according to variance explained by each parameter \cite{huang2018}.
+Certain inferences about deep soil carbon from this ensemble may therefore subject to uncertainty due to spin-up concerns.
 
@@
 One key parameter is \textit{tpuse\_sf}, which is a scalar perturbation factor influencing the triose phosphate limitations on photosynthesis.
-Triose phosphate limitation is difficult to quantify and has not been observed to significantly hinder photosynthesis under present-day CO$_2$ concentrations (cite Kumarathunge et al).
+Triose phosphate limitation is difficult to quantify and has not been observed to significantly hinder photosynthesis under present-day CO$_2$ concentrations \cite{kumarathunge2019}.
 In CLM5.1, this parameter is not particularly influential on NEP at 367 ppm (ranked 39, not shown), but is third most influential with high CO$_2$.
@@
 In a coupled ESM, land-atmosphere interactions would modulate these parameter effects, such that the parameter rankings are likely more robust than the absolute values of the parameter effects.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/main/deltas.png}
+\caption{The eight most influential parameters on NEP in the control simulations (a), high CO$_2$ simulations (b), and the relative response to high CO$_2$ (c), as well as on ET in the control simulations (d), future climate simulations (e) and the relative response to future climate (f).
+The parameters that are most influential on present-day climate (a,d) can differ significantly from the parameters that control the response to forcing perturbations (c,f).}
+\label{fig:nep}
+\end{figure}
 
@@
 Likewise figures are repeated for each of the various forcing scenarios.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=\textwidth]{../figs/main/deltas.png}
-\caption{The eight most influential parameters on NEP in the control simulations (a), high CO$_2$ simulations (b), and the relative response to high CO$_2$ (c), as well as on ET in the control simulations (d), future climate simulations (e) and the relative response to future climate (f).
-The parameters that are most influential on present-day climate (a,d) can differ significantly from the parameters that control the response to forcing perturbations (c,f).}
-\label{fig:nep}
-\end{figure}
 
@@
 In some cases, literature values were explicitly used (e.g.
-for the slope parameter of the stomatal conductance model, 'medlynslope', \citeA{lin2015}), but the most common range was $\pm20\%$.
+for the slope parameter of the stomatal conductance model, `medlynslope', \citeA{lin2015}), but the most common range was $\pm20\%$.
 It is exceedingly difficult to set parameter ranges that sample comparable probability density, even in a univariate experiment, like this one.
@@
 We opted for six forcing scenarios to identify parameter effects not just in present-day but in response to pre-industrial or future climate conditions.
-The parameters that were most influential under present-day conditions did not necessarily match the set of parameters controlling the response to future forcing (Figure \ref{fig:et}).
+The parameters that were most influential under present-day conditions did not necessarily match the set of parameters controlling the response to future forcing (Figure \ref{fig:nep}).
 We found that many acclimation parameters, which were not as important for determining present-day evapotranspiration (ET), were among the most influential on the response of ET to future climate.
 Similarly, the parameters controlling leaf area index varied significantly depending on biome (Figure \ref{fig:lai}).
-This emphasizes the importance of testing models not only under mean-state conditions, but also under experimentally perturbed conditions under global change analogues \cite(Wieder:2019).
+This emphasizes the importance of testing models not only under mean-state conditions, but also under experimentally perturbed conditions under global change analogues \cite{wieder2019}.
 
@@
 Now, in addition to seeing the definition of a given parameter, and the relevant equations, a model user can easily investigate the magnitude and spatial patterns of its effects (e.g.
-Figure \ref{fig:panel}).
+Figure \ref{fig:map}).
 This could be useful for investigating obvious model deficiencies, such as regional LAI biases, or to understand potential model responses under different forcing scenarios (e.g.
@@
 
-Our PPE is run as land-only simulations, using prescribed atmospheric forcing.
-This is likely to be adequate for estimating many land surface fluxes and pools, in particular of carbon, but could lead to biased estimates of water and energy fluxes, as we know that atmospheric feedbacks to land surface characteristics modify the state of the land surface itself (Laguë et al.
-2019).
-Additional work is ongoing to extend this work in a coupled model framework to quantify how much, where, and for which processes atmospheric feedbacks are likely to modulate the land parameter effects substantially.
+Our PPE utilizes land-only simulations, with prescribed atmospheric forcing.
+This is likely to be adequate for estimating many land surface fluxes and pools, in particular of carbon, but could lead to biased estimates of water and energy fluxes, as we know that atmospheric feedbacks to land surface characteristics modify the state of the land surface itself \cite{lague2019}.
+Additional work is ongoing to extend this work in a coupled model framework to quantify how much, where, and for which processes atmospheric feedbacks are likely to modulate land parameter effects substantially.
 
@@
 \acknowledgments
-Enter acknowledgments here.
-This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
 This material is based upon work supported by the National Center for Atmospheric Research (NCAR), which is a major facility sponsored by the National Science Foundation under Cooperative Agreement No.
@@
 Computing and data storage resources, including the Cheyenne supercomputer (doi:10.5065/D6RX99HX) were provided by the Climate Simulation Laboratory at NCAR’s Computational and Information Systems Laboratory (CISL).
+DK is supported via the Drought Task Force IV organized by the NOAA Climate Program Office’s Modeling, Analysis, Predictions, and Projections.
 FL is supported by the National Key Research and Development Program of China (2022YFE0106500).

### 2024-08-12 ms/ms.tex
overleaf

@@
 The PPE experiment required additional code modifications to enable systematic variation of the full suite of model parameters (many of which were previously `hard-coded', i.e., written directly as numerical values in the model source code).
+These modifications were incorporated into the main branch of the CTSM codebase for future use (see Open Research Section).
 We utilized the biogeochemical cycling version of CLM, which includes active carbon and nitrogen cycles.
@@
 This new module utilizes a matrix representation on CLM's biogeochemistry to significantly reduce spin-up time by one order of magnitude \cite{luo2022,liao2023}, and had been previously leveraged for a PPE with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) model \cite{huang2018}.
-Our spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (see \citeA{lawrence2019} and \citeA{thornton2005} for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode, each cycling over a ten-year forcing dataset (described below).
+Our spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (see \citeA{lawrence2019} and \citeA{thornton2005} for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode, each cycling over a ten-year forcing dataset (described in Section below).
 This protocol was designed to achieve sufficiently equilibrated carbon and nitrogen states, while minimizing computational time.
@@
 We tested 15 values of $k$, ranging from 10 to 800.
-Utilizing the ILAMB2.5 benchmarking software \cite{collier2018}, we calculated skill scores to quantify how well each candidate sparsegrid mirrored the full grid output (Supp Figure \ref{supp:ilamb}).
+Utilizing the ILAMB2.5 benchmarking software \cite{collier2018}, we calculated skill scores to quantify how well each candidate sparsegrid mirrored the full grid output (Supp Figure Stk).
 We opted for a 400-cluster sparsegrid, to balance computational cost against model fidelity.

### 2024-08-12 ms/ms.tex
refs

@@
 The number of parameters is large, with many lacking sufficient empirical backing for robust range evaluation.
-Defining appropriate ranges for the full suite of land model parameters is a major activity of the scientific subdomains that comprise the land model (e.g.
-Kattge et al.
-2020).
+Defining appropriate ranges for the full suite of land model parameters is a major activity of the scientific subdomains that comprise the land model \cite{kattge2020}.
 Even with parameters that have an empirical basis, they may behave differently at the coarse climate modeling spatial resolution (zqz), complicating the process of uncertainty quantification.

### 2024-08-12 ms/ms.tex
overleaf

@@
 This is as opposed to the satellite phenology mode, which is driven by observed canopy properties and thus has fewer internal feedbacks, and was explored in \citeA{dagon2020}.
-We used the model in land-only mode driven by observed meteorological forcing from the GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}), with the crop model turned off.
+We used the model in land-only mode driven by reanalysis atmosphere with the crop model turned off.
 
@@
 This new module utilizes a matrix representation on CLM's biogeochemistry to significantly reduce spin-up time by one order of magnitude \cite{luo2022,liao2023}, and had been previously leveraged for a PPE with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) model \cite{huang2018}.
-Our spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (see \citeA{lawrence2019} and \citeA{thornton2005} for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode, each cycling over a ten-year forcing dataset (described in Section below).
+Our spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (see \citeA{lawrence2019} and \citeA{thornton2005} for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode, each cycling over a ten-year forcing dataset (described in Section \ref{sect:exp}).
 This protocol was designed to achieve sufficiently equilibrated carbon and nitrogen states, while minimizing computational time.
@@
 
-\subsection{Experimental Design}
+\subsection{Parameter identification and ranges}
+
 Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants were hard-coded and not amenable to systematic perturbation.
@@
 
-Each simulation ran for 150 years, with the first 140 for spin-up (as described above in Section \ref{sect:mcn}), followed by a 10-year period for analysis.
-We opted for six different forcing scenarios to understand the intersection of parameter effects with different forcing agents associated with climate change (Table \ref{tab:exps}).
-The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and is the default forcing data for CLM5 \cite{lawrence2019}.
-We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
-Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the corresponding mean climate change signal (see Table \ref{tab:exps}; AF2095 and AF1855).
-We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for the six atmospheric forcing variables.
-The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, forcing in order to allow comparison with the existing CESM2 large ensemble.
-Our software infrastructure has greatly reduced the effort and computational cost associated with ensemble generation, such that it would be easy to add additional ensembles, to explore other emission scenarios or climate regimes (e.g.
-Last Glacial Maximum).
-
 For a small percentage of parameters, the perturbation ranges were severe enough to eradicate a given PFT across all or part of its geographical range (Supp Figure S1). 
@@
 $, or increased nitrogen deposition.
-In total 16 of the 211 parameters had the parameter range constrained to some extent. Original and constrained ranges are available in the supplemental materials spreadsheet (zqz).
+In total 16 of the 211 parameters had the parameter range constrained to some extent. Original and constrained ranges are available in a supplemental materials spreadsheet.
+
+\subsection{Experimental Design}
+\label{sect:exp}
+For each of the 211 parameters, we ran 12 simulations, one for each the minimum and maximum parameter values, across six different forcing scenarios.
+Each simulation ran for 150 years, with the first 140 for spin-up (as described above in Section \ref{sect:mcn}), followed by a 10-year period for analysis.
+It is important to note that these are equilibrium-style simulations, where the land has been allowed to reach steady state conditions relative to the climate forcing.
+This is achieved by repeating the same 10-year forcing dataset 14 times during the spin-up procedure.
+As such these simulations are not directly comparable to transient simulations or satellite observations, where present-day vegetation has not yet reached equilibrium with present-day CO$_2$.
+
+We opted for six different forcing scenarios to understand the intersection of parameter effects with different forcing agents associated with climate change (Table \ref{tab:exps}).
+The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and was the default forcing data for CLM5 \cite{lawrence2019}.
+We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
+Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the corresponding mean climate change signal (see Table \ref{tab:exps}; AF2095 and AF1855).
+We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for six atmospheric forcing variables (temperature, humidity, precipitation, wind, longwave and shortwave radiation; see Supp Figure Stk for climatologies).
+The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, forcing in line with the existing CESM2 large ensemble \cite{rodgers2021}.
+Since completing the experiment we uncovered a software bug affecting the AF1855 and AF2095 simulations.
+Namely that anomalies were not applied to certain coastal gridcells (Supp Fig Stk).
+In all this affected less than 3\% of land area, and did not significantly alter parameter rankings (Supp Fig Stk).
 
@@
 As such, the computational cost of inferring parameter effects is a major constraint.
-We were able to reduce ensemble generation computational cost 500x by strategically reducing the number of model grid cells (Figure \ref{fig:sg}) and by leveraging a Semi-Analytic Spin-Up (SASU) offered by the matrix approach to land biogeochemistry models (Lu et al.
-2020, Luo 2022, Liao et al.
-2023) and a linearized spin-up solver (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
+We were able to reduce ensemble generation computational cost 500x by strategically reducing the number of model grid cells (Figure \ref{fig:sg}) and by leveraging a Semi-Analytic Spin-Up (SASU) offered by the matrix approach to land biogeochemistry models \cite{lu2020,luo2022,liao2023} (see Sections \ref{sect:mcn} and \ref{sect:sg} for details).
 As long as computational resources remain constrained, designing faster model configurations and efficient sampling strategies will be important for effective model calibration and uncertainty quantification.
@@
 In the case of canopy evaporation, for example, more than 95\% of the ensemble variance could be explained by the ten most influential parameters (Figure \ref{fig:variance}).
-This indicates that there may be tractable parameter estimation sub-problems if global calibration of 211 parameters proves unattainable.
+This indicates that there may be tractable parameter estimation sub-problems if global calibration of all 211 parameters proves infeasible.
 
@@
 We found that many acclimation parameters, which were not as important for determining present-day evapotranspiration (ET), were among the most influential on the response of ET to future climate.
+This emphasizes the importance of testing models not only under mean-state conditions, but also under experimentally perturbed conditions under global change analogues \cite{wieder2019}.
 Similarly, the parameters controlling leaf area index varied significantly depending on biome (Figure \ref{fig:lai}).
-This emphasizes the importance of testing models not only under mean-state conditions, but also under experimentally perturbed conditions under global change analogues \cite{wieder2019}.
 

### 2024-08-13 ms/supp.tex
reorg supp

--- 2024-06-10 ms/supp.tex
@@
 \centering
-\includegraphics[width=50pc]{figs/supp/survivability.png}
-\caption{PFT survivability across the six ensembles.
-The sixteen natural vegetation PFTs are on the vertical axis, and the perturbation number sets the horizontal axis.
-A PFT is classified as dead in a given gridcell if it does not achieve an LAI of at least 0.1 m2/m2 at any point during the simulation.
-A PFT is classified as dead for a given perturbation if more than 50\% of the area that was alive with the default parameters is dead in the perturbed simulation.}
-\label{supp:surv}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=60pc]{figs/supp/ilamb_lines.png}
+\includegraphics[width=60pc]{../figs/supp/ilamb_lines.png}
 \caption{The convergence of several scoring metrics with increasing sparsegrid resolution across a variety of CLM variables.
@@
 
+\begin{landscape}
 \begin{figure}[h]
 \centering
-\includegraphics[width=\textwidth]{figs/supp/FCEV_global_mean.png}
-\caption{Global mean canopy evaporation (FCEV) parameter rankings across the six forcing scenarios.}
-\label{supp:fcev}
+\includegraphics[width=50pc]{../figs/supp/survivability.png}
+\caption{PFT survivability across the six ensembles.
+The sixteen natural vegetation PFTs are on the vertical axis, and the perturbation number sets the horizontal axis.
+A PFT is classified as dead in a given gridcell if it does not achieve an LAI of at least 0.1 m2/m2 at any point during the simulation.
+A PFT is classified as dead for a given perturbation if more than 50\% of the area that was alive with the default parameters is dead in the perturbed simulation.}
+\label{supp:surv}
 \end{figure}
+\end{landscape}
 
@@
 \centering
-\includegraphics[width=\textwidth]{figs/supp/anomalies.png}
+\includegraphics[width=\textwidth]{../figs/supp/anomalies.png}
 \caption{Maps of average forcing anomalies in the pre-industrial (left column) and future climate forcing (right column) simulations.
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/supp/anomaly_errors.png}
+\caption{Due to a software error, certain gridcells (yellow) were not appropriately processed to receive climate anomalies for the AF1855 and AF2095 simulations.
+Instead those gridcells experience the control climate.
+In total these errant gridcells total less than 3\% of global land area.
+}
+\label{supp:abug}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=30pc]{../figs/supp/AF2095_rankings.png}
+\caption{Comparing parameter rankings with and without the errant gridcells from Supp Figure S4 across a selection of output variables.}
+\label{supp:abug}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/supp/biome_tp.png}
+\caption{Each of the 400 sparse gridcells plotted according to their Whittaker biomes.}
+\label{supp:whit1}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/supp/biome_latlon.png}
+\caption{Cartesian plot of the 400 sparse gridcells labeled by their Whittaker biomes.}
+\label{supp:whit2}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=23pc]{../figs/supp/biome_areas.png}
+\caption{Total represented land area associated with each of the ten Whittaker biomes utilizing the CLM5.1 sparsegrid.}
+\label{supp:whit3}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{../figs/supp/FCEV_global_mean.png}
+\caption{Global mean canopy evaporation (FCEV) parameter rankings across the six forcing scenarios.}
+\label{supp:fcev}
+\end{figure}
+
 \end{document}

### 2024-08-13 ms/ms.tex
Updates from Overleaf

@@
 
-For a small percentage of parameters, the perturbation ranges were severe enough to eradicate a given PFT across all or part of its geographical range (Supp Figure S1). 
+For a small percentage of parameters, the perturbation ranges were severe enough to eradicate a given PFT across all or part of its geographical range (Supp Figure S2). 
 In the case that one or more PFTs did not survive a parameter perturbation, we iteratively re-ran the simulation with slightly less extreme parameter values until the PFT survived.
@@
 Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the corresponding mean climate change signal (see Table \ref{tab:exps}; AF2095 and AF1855).
-We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for six atmospheric forcing variables (temperature, humidity, precipitation, wind, longwave and shortwave radiation; see Supp Figure Stk for climatologies).
+We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for six atmospheric forcing variables (temperature, humidity, precipitation, wind, longwave and shortwave radiation; see Supp Figure S3 for climatologies).
 The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, forcing in line with the existing CESM2 large ensemble \cite{rodgers2021}.
 Since completing the experiment we uncovered a software bug affecting the AF1855 and AF2095 simulations.
-Namely that anomalies were not applied to certain coastal gridcells (Supp Fig Stk).
-In all this affected less than 3\% of land area, and did not significantly alter parameter rankings (Supp Fig Stk).
+Namely that anomalies were not applied to certain coastal gridcells (Supp Fig S4).
+In all this affected less than 3\% of land area, and did not significantly alter parameter rankings (Supp Fig S5).
 

### 2024-08-13 ms/ms.tex
overleaf

@@
 To manage computational cost we leveraged the Semi-Analytic spin-up mode (SASU) recently implemented within CLM5 \cite{lu2020,liao2023}.
-This new module utilizes a matrix representation on CLM's biogeochemistry to significantly reduce spin-up time by one order of magnitude \cite{luo2022,liao2023}, and had been previously leveraged for a PPE with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) model \cite{huang2018}.
-Our spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (see \citeA{lawrence2019} and \citeA{thornton2005} for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode, each cycling over a ten-year forcing dataset (described in Section \ref{sect:exp}).
+This new module utilizes a matrix representation on CLM's biogeochemistry to reduce spin-up time by one order of magnitude \cite{luo2022,liao2023}, and had been previously leveraged for a PPE with the Organizing Carbon and Hydrology in Dynamic Ecosystems (ORCHIDEE) model \cite{huang2018}.
+Our spin-up protocol featured 20 years in Accelerated Decomposition spin-up mode (see \citeA{lawrence2019} and \citeA{thornton2005} for details), followed by 80 years of SASU, followed by 40 years of Native Dynamic spin-up mode (details on forcing data in Section \ref{sect:exp}).
 This protocol was designed to achieve sufficiently equilibrated carbon and nitrogen states, while minimizing computational time.
@@
 We tested 15 values of $k$, ranging from 10 to 800.
-Utilizing the ILAMB2.5 benchmarking software \cite{collier2018}, we calculated skill scores to quantify how well each candidate sparsegrid mirrored the full grid output (Supp Figure Stk).
+Utilizing the ILAMB2.5 benchmarking software \cite{collier2018}, we calculated skill scores to quantify how well each candidate sparsegrid mirrored the full grid output (Supp Figure S1).
 We opted for a 400-cluster sparsegrid, to balance computational cost against model fidelity.
@@
 
-Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants were hard-coded and not amenable to systematic perturbation.
+Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants are hard-coded and not amenable to systematic perturbation.
 For the purposes of this activity, we identified a broad set of 211 CLM-BGC parameters, and enabled their modification via the model input files.

### 2024-08-13 ms/ms.tex
refs

@@
 
-Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants are hard-coded and not amenable to systematic perturbation.
+Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants are hard-coded and not amenable to systematic perturbation \cite{mendoza2015,cuntz2016}.
 For the purposes of this activity, we identified a broad set of 211 CLM-BGC parameters, and enabled their modification via the model input files.

### 2024-08-13 ms/ms.tex
overleaf

@@
 We likewise opted to perturb soil hydraulic parameters directly (e.g.
-saturated hydraulic conductivity) and also via perturbations to the soil texture (percent sand, clay, and silt).
+saturated hydraulic conductivity) and also separately via perturbations to the soil texture (percent sand, clay, and silt).
 
@@
 It is important to note that these are equilibrium-style simulations, where the land has been allowed to reach steady state conditions relative to the climate forcing.
-This is achieved by repeating the same 10-year forcing dataset 14 times during the spin-up procedure.
+This is achieved by repeating the same 10-year forcing dataset 15 times during the spin-up procedure and production phase of our simulations.
 As such these simulations are not directly comparable to transient simulations or satellite observations, where present-day vegetation has not yet reached equilibrium with present-day CO$_2$.
@@
 
-\subsection{Analyses}
-
-Biome analysis (Supp Figures S2-4)
+\subsection{Biome definitions}
+In addition to analyzing parameter effects on various output variables globally, we were also interested in understanding parameter effects on different biomes.
+As such, we categorized each of the 400 sparse grid cells according to their Whittaker biome, which delineates biomes based on average temperature and precipitation \cite{whittaker1970}. 
+Biomes were delineated based on the atmospheric forcing temperature and precipitation data averaged across 2005 to 2014 (Supp Figure S6).
+Overall there are nine vegetated biomes and an additional ice sheet biome, where no vegetation is present (Supp Figures S7,8).
 

### 2024-08-19 ms/ms.tex
overleaf

@@
 Depending on the evaluation criteria and model configuration, CLM5 requires between 800 and 2000 years (or more) to reach steady-state conditions \cite{lawrence2019}.
-In the absence of equilibrium, the drift towards steady state can obscure important model dynamics or features.
+In the absence of equilibrium, the drift towards steady state can obscure important model dynamics or features \cite{seferian2016}.
 Because each member of the PPE can have a unique steady state, we performed an independent spin-up for each member.
@@
 \subsection{Biome definitions}
+\label{sect:whit}
 In addition to analyzing parameter effects on various output variables globally, we were also interested in understanding parameter effects on different biomes.
@@
 
-A small fraction of parameters tend to explain a large amount of the ensemble variance for any given variable or metric (Figure \ref{fig:variance}).
+A small fraction of parameters tends to explain a large amount of the ensemble variance for any given variable or metric (Figure \ref{fig:variance}).
 For example, just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S3).
@@
 
-This PPE is especially useful for determining which parameters influence which model processes and where.
-For example the parameters controlling leaf area index vary significantly by biome (Figure \ref{fig:lai}).
-Plant hydraulics parameters were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
-The most influential parameters for leaf area index globally included parameters that were important in each of these three biomes.
-
 We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps}.
@@
 For example, the parameters that control present-day NEP differ  from the parameters that control NEP in the high CO$_2$ scenario (Figure \ref{fig:nep}a,b).
-NEP increases by approximately 50\% at 867 ppm with the default parameters, but can actually decrease with certain parameter settings (Figure \ref{fig:nep}c).
+NEP increases by approximately 50\% with the increase of CO$_2$ from 367 to 867 ppm when using the CLM5.1 default parameters, but can actually decrease with certain parameter settings (Figure \ref{fig:nep}c).
 One key parameter is \textit{tpuse\_sf}, which is a scalar perturbation factor influencing the triose phosphate limitations on photosynthesis.
@@
 \textit{tpuse\_sf}, \textit{kcha}, \textit{vcmaxha}, \textit{vcmaxhd}, \textit{lmrhd}, and \textit{cpha}).
-In a coupled ESM, land-atmosphere interactions would modulate these parameter effects, such that the parameter rankings are likely more robust than the absolute values of the parameter effects.
+In a coupled ESM, land-atmosphere interactions would modulate these parameter effects, which could dampen parameter effects.
+While we feel relatively confident identifying influential parameters using land-only simulations, projecting their exact impacts in a coupled framework will require coupled model sensitivity tests.
+
+We repeated parameter ranking analyses in each of nine Whittaker biomes (see Section \ref{sect:whit}).
+We found that the parameters controlling leaf area index, for example, vary significantly by biome (Figure \ref{fig:lai}).
+Plant hydraulics parameters were the most important in the tropical rain forest, photosynthetic capacity in the boreal forest, and runoff and soil evaporation in the temperate grassland/desert biome.
+The most influential parameters for leaf area index globally included parameters that were important in each of these three biomes.
 

### 2024-08-19 ms/ms.tex
refs

@@
 Parameter sensitivity tests can be used to gauge parametric uncertainty, and
-a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \cite{dagon2020}.
+a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \cite{dagon2020,macneall2024}.
 One important application of PPEs has been assessing uncertainty in climate model projections \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian methods \cite{cleary2021}.
+Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,macneall2024} or Bayesian methods \cite{cleary2021}.
 Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large PPEs \cite{qian2018}.

### 2024-08-19 ms/ms.tex
fix refs

--- 2024-08-19 ms/ms.tex
@@
 Parameter sensitivity tests can be used to gauge parametric uncertainty, and
-a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \cite{dagon2020}.
+a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \cite{dagon2020,mcneall2024}.
 One important application of PPEs has been assessing uncertainty in climate model projections \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021} or Bayesian methods \cite{cleary2021}.
+Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024} or Bayesian methods \cite{cleary2021}.
 Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large PPEs \cite{qian2018}.

### 2024-08-28 ms/ms.tex
more eidts

@@
 Lawrence\affil{1},
-F. Li\affil{11}, D.L. Lombardozzi\affil{1,12}, Y. Luo\affil{12}, J.K. Shuman\affil{13}, A.L.S. Swann\affil{7,14}, S.C. Swenson\affil{1}, G. Tang\affil{1}, W.R. Wieder\affil{1}, and A.W. Wood\affil{1}}
+F. Li\affil{11}, D.L. Lombardozzi\affil{1,12}, Y. Luo\affil{13}, J.K. Shuman\affil{14}, A.L.S. Swann\affil{7,15}, S.C. Swenson\affil{1}, G. Tang\affil{1}, W.R. Wieder\affil{1}, and A.W. Wood\affil{1}}
 
@@
 \affiliation{11}{International Center for Climate and Environment Sciences, Institute of Atmospheric Physics, Chinese Academy of Sciences, Beijing, China}
-\affiliation{12}{Colorado State University, Fort Collins, CO, USA}
+\affiliation{12}{Department of Ecosystem Science and Sustainability, Colorado State University, Fort Collins, CO, USA}
 \affiliation{13}{School of Integrative Plant Science, Cornell University, NY, USA}
@@
 \begin{itemize}
-\item 211 parameters perturbed
-\item featherweight clm configuration, 500x less costly
+\item 211 parameters perturbed across six forcing scenarios
+\item expansive dataset available which can be used to identify the most influential parameters on a wide range of output variables globally, by biome, or by plant functional type
 \item parameter effects can exceed scenario effects
-\item large dataset available online providing parameter ranges and effect sizes useful for future studies
-\item software infrastructure developed to facilitate routine investigation of parameter sensitivity and uncertainty, as well as automated calibration
 \item small number of parameters explains a large fraction of variance 
 \item most important parameters can vary regionally and also based on climate forcing
-
+\item software infrastructure developed to facilitate routine investigation of parameter sensitivity and uncertainty, as well as automated calibration
 \end{itemize}
@@
 Uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}.
-Land processes substantially influence climate directly through, for example, evapotranspiration \cite{zarakas2024}, and indirectly through carbon-climate feedbacks.
+Land processes substantially influence climate directly through, for example, evapotranspiration \cite{zarakas2024}, and indirectly through carbon-climate feedbacks \cite{cox2000}.
 Making centennial-scale projections of the cumulative terrestrial carbon sink gas been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}, and limited efficacy of emergent constraints.
@@
 The remaining 66 parameters include a plant functional type (PFT) dimension, with the potential for different parameter values for each unique plant functional type.
-Not all parameters are currently utilizing this functionality, such as \textit{medlynintercept} where the default value is 100 $\mu$mol H$_2$O for all sixteen of CLM's natural vegetation PFTs.
+Not all parameters are currently utilizing this functionality, such as \textit{medlynintercept} where the default value is 100 $\mu$mol H$_2$O m$^{-2}$ s$^{-1}$ for all sixteen of CLM's natural vegetation PFTs.
 
@@
 To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter.
-In some cases literature values were directly utilized, but in many cases, expert judgment was used.
-The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.
+In some cases literature values were directly utilized, but in many cases, expert judgment was used or else a default perturbation of $\pm$20\%.
+The parameter range spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.
 In some cases parameters could not or should not be perturbed independently, because they feature inherent co-dependence.
@@
 We likewise opted to perturb soil hydraulic parameters directly (e.g.
-saturated hydraulic conductivity) and also separately via perturbations to the soil texture (percent sand, clay, and silt).
+saturated hydraulic conductivity) and also indirectly, in a parallel set of simulations via perturbations to the soil texture (percent sand, clay, and silt).
 
@@
 In the case that one or more PFTs did not survive a parameter perturbation, we iteratively re-ran the simulation with slightly less extreme parameter values until the PFT survived.
-A PFT was classified as surviving in a given gridcell if it achieves an LAI of at least 0.1 m2/m2 at any point during the production portion of the simulation (i.e. post-spinup).
-A PFT was classified as surviving for a given parameter perturbation if at least 50\% of the area that
+A PFT was classified as surviving in a given gridcell if it achieves an LAI of at least 0.1 m2/m2 at any point after the spinup period.
+A PFT was classified as surviving globally if at least 50\% of the area that
 was alive with the default parameters is alive in the perturbed simulation.
-For the analyses in this paper, we constrained the ranges of several parameters based on survivability in the low CO$_2$, pre-industrial climate, and control climate ensembles to improve the chances that all PFTs would be able to survive to present-day in a transient 1850 to present-day simulation.
+For the analyses in this paper, we constrained the ranges of several parameters based on survivability in the low CO$_2$, pre-industrial climate, and control climate ensembles to improve the chances that all PFTs would be able to survive in a transient 1850 to present-day simulation.
 We did not constrain parameter ranges based on future climate, high CO$_2
@@
 \label{sect:exp}
-For each of the 211 parameters, we ran 12 simulations, one for each the minimum and maximum parameter values, across six different forcing scenarios.
+For each of the 211 parameters, we ran 12 simulations, one for each of the minimum and maximum parameter values, across six different forcing scenarios.
 Each simulation ran for 150 years, with the first 140 for spin-up (as described above in Section \ref{sect:mcn}), followed by a 10-year period for analysis.
 It is important to note that these are equilibrium-style simulations, where the land has been allowed to reach steady state conditions relative to the climate forcing.
-This is achieved by repeating the same 10-year forcing dataset 15 times during the spin-up procedure and production phase of our simulations.
+This is achieved by repeating a single 10-year forcing dataset 15 times during the spin-up procedure and production phase of our simulations.
 As such these simulations are not directly comparable to transient simulations or satellite observations, where present-day vegetation has not yet reached equilibrium with present-day CO$_2$.
@@
  We opted for six different forcing scenarios to understand the intersection of parameter effects with different forcing agents associated with climate change (Table \ref{tab:exps}).
-The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, and was the default forcing data for CLM5 \cite{lawrence2019}.
+The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, as it is the default forcing data for CLM5 \cite{lawrence2019}.
 We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
@@
 We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for six atmospheric forcing variables (temperature, humidity, precipitation, wind, longwave and shortwave radiation; see Supp Figure S3 for climatologies).
-The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, forcing in line with the existing CESM2 large ensemble \cite{rodgers2021}.
+The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, which was chosen to align with the existing CESM2 large ensemble \cite{rodgers2021}.
 Since completing the experiment we uncovered a software bug affecting the AF1855 and AF2095 simulations.
@@
 In this project, we identified and perturbed 211 CLM parameters to create a large one-at-a-time PPE.
-This ensemble is useful for understanding parametric controls on CLM processes.
-The software and analysis tools generated to enable and utilize this experiment will greatly reduce the burden of generating future PPEs and parameter sensitivity experiments.
-
 There were several barriers to perturbing the full set of CLM5-BGC parameters.
-First, many parameters had not been officially identified as such.
+First, many parameters had not been officially identified as parameters.
 In such cases, we identified hard-coded values, established an appropriate parameter name, and extracted that parameter to the CLM parameter file for easier manipulation.
-Although in this experiment we perturbed a large number of parameters (211) across a variety of CLM processes (Figure \ref{fig:params}), this still does not cover the full set of CLM parameters, as some processes were not included, such as crop phenology and management.
+Although in this experiment we perturbed a large number of parameters across a variety of CLM processes (Figure \ref{fig:params}), this still does not cover the full set of CLM parameters, as some processes were not included, such as crop phenology and management.
 In carrying out this process of parameter identification, we likewise unearthed many nuances in the epistemology of what `parameters' are in the context of Earth system models.
+For instance, is soil texture a forcing variable or a parameter?
 Defining the parameters within a given model structure can be somewhat subjective, such that it would be difficult to collate a comprehensive or definitive set of parameters for a model like CLM.
+For example, for several empirical regressions, we scaled the slope and intercept terms in tandem, because we lacked sufficient information to define two independent perturbation ranges.
 
@@
 The number of parameters is large, with many lacking sufficient empirical backing for robust range evaluation.
-Defining appropriate ranges for the full suite of land model parameters is a major activity of the scientific subdomains that comprise the land model \cite{kattge2020}.
-Even with parameters that have an empirical basis, they may behave differently at the coarse climate modeling spatial resolution (zqz), complicating the process of uncertainty quantification.
+Defining appropriate ranges for the full suite of land model parameters is a major activity of the scientific subdomains that comprise land models \cite{kattge2020}.
+Even with parameters that have an empirical basis from field studies, they may behave differently at the coarse climate modeling spatial resolution (zqz), complicating the process of uncertainty quantification.
 As such, we cannot claim that the parameters are equivalently sampled, whereby parameter effects and parameter rankings could be subject to sampling asymmetries.
@@
 This emphasizes the importance of testing models not only under mean-state conditions, but also under experimentally perturbed conditions under global change analogues \cite{wieder2019}.
-Similarly, the parameters controlling leaf area index varied significantly depending on biome (Figure \ref{fig:lai}).
+Furthermore, because parameter values in CLM are time-invariant, plants must be able to survive pre-industrial conditions in order to be present for the later stages of a transient simulation, such as those performed for many model intercomparison experiments.
+Several perturbations did indeed cause survivability issues (Supp Figure S2), in which case we opted to constrain the parameter ranges until all PFTs were able to survive in the low CO$_2$, control, and pre-industrial climate simulations.
+In future experiments we may instead opt for fully transient simulations from 1850-2100, as the computational cost is comparable given the overhead of independent spinups for each forcing scenario.
+
+Just as the most influential parameters varied by forcing scenario, parameter rankings varied significantly by biome (Figure \ref{fig:lai}).
+Different PFTs present varying trait strategies, and different climates inflict unique challenges to vegetation activity, such as light availability, extreme temperatures, or water stress.
+As such, perturbations that serve to ease or exacerbate the most substantial stressor for the set of PFTs in a given biome will have greater effect. 
+Understanding the regional signatures of parameter perturbations is important for model calibration.
+Previous efforts have experienced difficulty resolving regional biases due to a limited amount of parameter flexibility.
+For example, \cite{dagon2020} could not simultaneously optimize plant productivity in the Amazon and Sahel regions.
+In some cases, perturbing PFT-specific parameters independently could aid in providing regional levers for bias reduction.
+In our experiments, PFT-specific parameters were perturbed in unison, but their effects can be analyzed by PFT for many variables that are modeled at the PFT-level, such as photosynthesis and transpiration.
+To first order, we expect that PFT-specific parameter effects will be independent, but CLM does feature some implicit competition among natural vegetation PFTs, such as through a shared soil moisture reservoir, which could impose some cross PFT interaction.
 
@@
 
-Our PPE utilizes land-only simulations, with prescribed atmospheric forcing.
-This is likely to be adequate for estimating many land surface fluxes and pools, in particular of carbon, but could lead to biased estimates of water and energy fluxes, as we know that atmospheric feedbacks to land surface characteristics modify the state of the land surface itself \cite{lague2019}.
+Another caveat is that our PPE utilizes land-only simulations, with prescribed atmospheric forcing.
+This is likely to be adequate for estimating many land surface fluxes and pools, in particular of carbon, but precludes the influence of land-atmosphere feedbacks.
+This could lead to biased estimates of water and energy fluxes, as atmospheric feedbacks can modify the state of the land surface itself \cite{lague2019}.
 Additional work is ongoing to extend this work in a coupled model framework to quantify how much, where, and for which processes atmospheric feedbacks are likely to modulate land parameter effects substantially.
+Our expectation is that relative parameter rankings are likely robust, but that the absolute value of parameter effects may be significantly biased relative to their effects in the coupled model.
 
-Several spin-off projects have leveraged the work presented here, including some that have already reached publication status \cite{cheng2023,yan2023a,yan2023b}.
-These projects utilized the output from this ensemble to filter for parameters that influence the relevant study domains and used the parameter ranges collated in Section \ref{sect:exps}.
-Our project has accelerated parameter exploration work within our collaborator network by providing:
+The primary deliverable of this project is the PPE dataset that captures the effects of 212 parameters on a wide range of CLM variables across six forcing scenarios.
+Several spin-off projects have leveraged this dataset, including some that have already reached publication \cite{cheng2023,yan2023a,yan2023b}.
+These projects utilized our ensemble to filter for parameters that influence the relevant study domains and variables, and performed follow-on experiments using the parameter ranges collated in Section \ref{sect:exps}.
+
+The infrastructure underpinning this experiment is another major deliverable. Our project has accelerated parameter exploration work within our collaborator network by providing:
 
 \begin{itemize}
-\item Parameter ranges for nearly 200 parameters
+\item Parameter ranges for more than 200 parameters
 \item Purpose-built unstructured grid (sparsegrid)
@@
 \item Ensemble generation scripting toolchain
-\item Parameter sensitivity diagnostics for nearly 200 parameters across 6 forcing scenarios
+\item Global, PFT, and biome-level parameter sensitivity diagnostics across six forcing scenarios
 \end{itemize}
 
-We have likewise begun a follow-on activity that perturbs a subset of important parameters with the Latin hypercube sampling strategy \cite{mckay2000} to work towards routine model calibration.
+The software, datasets, and analysis tools generated to enable and utilize this experiment will greatly reduce the burden of generating future PPEs and parameter sensitivity experiments.
+All of the code to generate this ensemble is available via github.
+We have also made available the collated parameter ranges and the output from this ensemble (see the Open Research Section for details).
+
+We have already begun a follow-on activity that perturbs a subset of important parameters with a more exhaustive sampling approach to work towards calibration of leaf area index.
 By investing in the infrastructure that we introduce here, all of our subsequent parameter perturbation experiments have required much less time and effort.
-We expect to continue to extend our efforts in this domain towards model calibration and uncertainty quantification, as well as repeating the foundational one-at-a-time experiments with subsequent model releases.
+We continue to extend our efforts in this domain towards model calibration and uncertainty quantification, and we anticipate repeating the foundational one-at-a-time experiment with subsequent model releases.
+Reducing both the human and computer time required for these experiments is allowing us to repeat them more frequently through the model development cycle.
 
@@
 
- (The CTSM component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP})
+The CTSM component set longname is: \\ \texttt{2000\_DATM\%GSWP3v1\_CLM51\%BGC\_SICE\_SOCN\_SROF\_SGLC\_SWAV\_SIAC\_SESP}
 
-This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
-Data cannot be listed as ''Available from authors'' or stored solely in supporting information.
-Citations to archived data should be included in your reference list.
-Wiley will publish it as a separate section on the paper’s page.
-Examples and complete information are here:
-https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
+A relatively small ($\sim$300MB), post-processed dataset is freely available online.
+All of the figures on our diagnostics website can be made from this one file.
+The raw dataset will be available on the NSF-NCAR computing system, and can be made available online as well, but we are gauging interest before committing to finding a host for the 2TB of data.
+
+Ensemble generation scripts are available via: \\
+\url{https://github.com/djk2120/ppe_tools}
+
+All of the code to make the figures in this manuscript is available via: \\
+\url{https://github.com/djk2120/oaat_clm5_ppe}
 

### 2024-08-28 ms/ms.tex
bib

@@
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
-Still, our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns.
+Still, our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns \cite{worden2021,baldocchi2024}.
 The capability to efficiently ingest these data and improve simulation performance is especially valuable for actionable science \cite{cheng2023}.

### 2024-08-28 ms/ms.tex
more edits

@@
 \begin{keypoints}
-\item enter point 1 here
+\item We constructed a large parameter pertubration ensemble of the Community Land Model, version 5.1, perturbing 211 parameters independently across six forcing scenarios.
+\item Parameter effects can exceed scenario effects and parameter rankings vary by biome and based on the forcing scenario.
+\item The software infrastructure developed for this experiment has greatly reduced the human and computer time needed for constructing this type of ensemble, which can facilitate more routine PPEs.
 \end{keypoints}
@@
 \begin{abstract}
-\begin{itemize}
-\item 211 parameters perturbed across six forcing scenarios
-\item expansive dataset available which can be used to identify the most influential parameters on a wide range of output variables globally, by biome, or by plant functional type
-\item parameter effects can exceed scenario effects
-\item small number of parameters explains a large fraction of variance 
-\item most important parameters can vary regionally and also based on climate forcing
-\item software infrastructure developed to facilitate routine investigation of parameter sensitivity and uncertainty, as well as automated calibration
-\end{itemize}
+Comprehensive land models are subject to significant parametric uncertainty, which can be hard to quantify due to the large number of parameters and high model computational costs.
+We constructed a large parameter pertubration ensemble (PPE) for the Community Terrestrial Systems Model in the Community Land Model with biogeochemistry configuration (CLM-BGC).
+We performed more than 2000 simulations perturbing 211 parameters across six forcing scenarios.
+This provides an expansive dataset, which can be used to identify the most influential parameters on a wide range of output variables globally, by biome, or by plant functional type.
+We found that parameter effects can exceed scenario effects and that a small number of parameters explains a large fraction of variance across our ensemble. 
+The most important parameters can vary regionally and also based on climate forcing.
+The software infrastructure developed for this experiment has greatly reduced the human and computer time needed for CLM PPEs, which can facilitate routine investigation of parameter sensitivity and uncertainty, as well as automated calibration.
 \end{abstract}
@@
 Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many processes represented within the terrestrial components of Earth System Models.
-Understanding how these processes respond to and influence CO$_2$ concentrations is a critical facet of climate change research.
+Understanding how these processes respond to rising CO$_2$ and even influence CO$_2$ concentration trajectories is a critical facet of climate change research.
+Land processes substantially influence climate directly through, for example, evapotranspiration \cite{zarakas2024}, and indirectly through carbon-climate feedbacks \cite{cox2000}.
 Uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}. 
-Land processes substantially influence climate directly through, for example, evapotranspiration \cite{zarakas2024}, and indirectly through carbon-climate feedbacks \cite{cox2000}.
-Making centennial-scale projections of the cumulative terrestrial carbon sink gas been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}, and limited efficacy of emergent constraints.
+Making centennial-scale projections of the cumulative terrestrial carbon sink gas been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}, and limited efficacy of emergent constraints \cite{liu2024}.
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
@@
 
-The process of parameter estimation for the complex land models typically embedded within Earth System Models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that impact the terrestrial biosphere (e.g, hydrology, snow and ice sheets, biogeochemistry, physiology, land management, etc.).
+The process of parameter estimation for the complex land models typically embedded within Earth System Models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that comprise the terrestrial biosphere (e.g, hydrology, snow and ice sheets, biogeochemistry, physiology, land management, etc.).
 The number of model parameters required to represent all the relevant processes is large, and model simulations are relatively expensive, presenting a major barrier to robust objective calibration, model interpretation, and development \cite{fisher2020,dagon2020}.

### 2024-09-04 ms/ms.tex
edits

@@
+†isl
 \documentclass[draft]{agujournal2019}
@@
 Lawrence\affil{1},
-F. Li\affil{11}, D.L. Lombardozzi\affil{1,12}, Y. Luo\affil{13}, J.K. Shuman\affil{14}, A.L.S. Swann\affil{7,15}, S.C. Swenson\affil{1}, G. Tang\affil{1}, W.R. Wieder\affil{1}, and A.W. Wood\affil{1}}
+F.
+Li\affil{11}, D.L.
+Lombardozzi\affil{1,12}, Y.
+Luo\affil{13}, J.K.
+Shuman\affil{14}, A.L.S.
+Swann\affil{7,15}, S.C.
+Swenson\affil{1}, G.
+Tang\affil{1}, W.R.
+Wieder\affil{1,16}, and A.W.
+Wood\affil{1}}
 
-\affiliation{1}{Climate and Global Dynamics Laboratory, NCAR, Boulder, CO, USA}
+\affiliation{1}{Climate and Global Dynamics Laboratory, NSF NCAR, Boulder, CO, USA}
 \affiliation{2}{Earth Research Institute, University of California, Santa Barbara, CA, USA}
@@
 \affiliation{15}{Department of Biology, University of Washington, Seattle, WA, USA}
+\affiliation{16}{Institute of Arctic and Alpine Research, University of Colorado, Boulder, CO, USA}
 \correspondingauthor{Daniel Kennedy}{djk2120@ucar.edu}
@@
 \begin{keypoints}
-\item We constructed a large parameter pertubration ensemble of the Community Land Model, version 5.1, perturbing 211 parameters independently across six forcing scenarios.
+\item We constructed a large parameter perturbation ensemble of the Community Land Model, version 5.1, perturbing 211 parameters independently across six forcing scenarios.
 \item Parameter effects can exceed scenario effects and parameter rankings vary by biome and based on the forcing scenario.
@@
 Comprehensive land models are subject to significant parametric uncertainty, which can be hard to quantify due to the large number of parameters and high model computational costs.
-We constructed a large parameter pertubration ensemble (PPE) for the Community Terrestrial Systems Model in the Community Land Model with biogeochemistry configuration (CLM-BGC).
+We constructed a large parameter perturbation ensemble (PPE) for the Community Land Model with biogeochemistry configuration (CLM-BGC).
 We performed more than 2000 simulations perturbing 211 parameters across six forcing scenarios.
@@
 We found that parameter effects can exceed scenario effects and that a small number of parameters explains a large fraction of variance across our ensemble.
-The most important parameters can vary regionally and also based on climate forcing.
+The most important parameters can vary regionally and also based on forcing scenario.
 The software infrastructure developed for this experiment has greatly reduced the human and computer time needed for CLM PPEs, which can facilitate routine investigation of parameter sensitivity and uncertainty, as well as automated calibration.
@@
 Uncertainty in climate model projections varies by domain and generally increases with extended time horizons \cite{koven2022}. 
-Making centennial-scale projections of the cumulative terrestrial carbon sink gas been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}, and limited efficacy of emergent constraints \cite{liu2024}.
+Making centennial-scale projections of the cumulative terrestrial carbon sink has been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2006,friedlingstein2014,arora2020}, and limited efficacy of emergent constraints \cite{liu2024}.
 A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
@@
 Understanding the range of outcomes arising from the many plausible parameter combinations is a critical step in robust uncertainty quantification.
-Parameter sensitivity tests can be used to gauge parametric uncertainty, and
-a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \cite{dagon2020,mcneall2024}.
+Parameter sensitivity tests can be used to gauge parametric uncertainty, and a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \cite{dagon2020,mcneall2024}.
 One important application of PPEs has been assessing uncertainty in climate model projections \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through, for example, history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024} or Bayesian methods \cite{cleary2021}.
+Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through various approaches like history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024} or Bayesian methods \cite{cleary2021}.
 Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large PPEs \cite{qian2018}.
@@
 
-The process of parameter estimation for the complex land models typically embedded within Earth System Models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that comprise the terrestrial biosphere (e.g, hydrology, snow and ice sheets, biogeochemistry, physiology, land management, etc.).
-The number of model parameters required to represent all the relevant processes is large, and model simulations are relatively expensive, presenting a major barrier to robust objective calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
+The process of parameter estimation for land models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity of plant and microbial life, and the multiplicity of domains that comprise the terrestrial biosphere (e.g, hydrology, snow, biogeochemistry, physiology, land management, etc.).
+The number of model parameters required to represent all the relevant processes is large and model simulations are relatively expensive; presenting a major barrier to robust objective calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
 Thorough assessment of the parametric sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}.
@@
 
-In this project we create a large PPE with the Community Terrestrial Systems Model (CTSM) \cite{lawrence2019}, in the Community Land Model with biogeochemistry configuration (CLM-BGC).
+In this project we create a large PPE with the Community Land Model with biogeochemistry configuration (CLM-BGC)\cite{lawrence2019}.
 We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
 Here we include a broader suite of parameters and continue to enhance the software tools for parameter perturbation and optimization.
-The goal of this project is to systematize the parameter perturbation process within the CTSM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across a wide range of land model processes.
+The goal of this project is to systematize the parameter perturbation process within the CLM modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across a wide range of land model processes.
 In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
 In this paper we present that dataset, describe how it was produced, and survey some potential applications.
-The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b,zarakas2024}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CTSM/CLM.
+The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b,zarakas2024}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CLM.
 
@@
 This experiment utilizes the Community Land Model configuration (version 5.1, i.e.
-CLM5.1) of the Community Terrestrial Systems Model.
+CLM5.1).
 The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.
@@
 The PPE experiment required additional code modifications to enable systematic variation of the full suite of model parameters (many of which were previously `hard-coded', i.e., written directly as numerical values in the model source code).
-These modifications were incorporated into the main branch of the CTSM codebase for future use (see Open Research Section).
-We utilized the biogeochemical cycling version of CLM, which includes active carbon and nitrogen cycles.
+These modifications were incorporated into the main branch of the CLM codebase for future use (see Open Research Section).
+We utilized the biogeochemistry version of CLM, which includes active carbon and nitrogen cycles.
 This is as opposed to the satellite phenology mode, which is driven by observed canopy properties and thus has fewer internal feedbacks, and was explored in \citeA{dagon2020}.
-We used the model in land-only mode driven by reanalysis atmosphere with the crop model turned off.
+We used the model in land-only mode driven by a data atmosphere from the Global Soil Wetness Project (GSWP3; REF), with the crop model turned off.
 
@@
 In the absence of equilibrium, the drift towards steady state can obscure important model dynamics or features \cite{seferian2016}.
-Because each member of the PPE can have a unique steady state, we performed an independent spin-up for each member.
+Because each member of the PPE can have a unique steady state, we performed an independent spin-up for each ensemble member.
 
@@
 Another control on model cost is spatial resolution.
-Most global CTSM simulations utilize nominal 1$^\circ$ resolution, which equates to about 20,000 land grid cells.
-In order to manage computational cost, parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}.
-Here, we use a clustering algorithm to achieve an alternative low resolution configuration.
+Most global CLM simulations utilize nominal 1$^\circ$ resolution, which equates to about 20,000 land grid cells.
+To reduce computational costs parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}.
+As opposed to this coarse grid resolution, here we apply an alternative sparse grid approach by useing a clustering algorithm to achieve an alternative low resolution configuration.
 The goal is to more strategically distribute grid cells, opting for more grid cells in areas that are especially distinct or variable, and fewer in areas that can be easily approximated based on output elsewhere (e.g.
@@
 Multivariate spatio-temporal clustering (MVSC) has been utilized to extract patterns of climatological significance from climate model output \cite{hoffman2005} and applied to design a representativeness-based sampling network \cite{hoffman2013}.
-Instead of lowering resolution by coarsening a rectilinear grid, we here used MVSC to strategically remove effectively redundant grid cells.
-
-We used k-means clustering to identify groups of grid cells with similar dynamics based on a 2$^\circ$ transient simulation (1850-2014) using the CTSM-PPE codebase.
+Instead of lowering resolution by coarsening a rectilinear grid, we here used MVSC to strategically remove effectively redundant grid cells.We used k-means clustering to identify groups of grid cells with similar dynamics based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
 We selected one representative grid cell from each cluster to stand in for the entire cluster.
@@
 
-Clustering was based on a subset of 18 CTSM variables (Table \ref{tab:sg}).
+Clustering was based on a subset of 18 CLM variables (Table \ref{tab:sg}).
 The MVSC algorithm delineates clusters based on the mean and variability of each variable at each grid cell computed for six 30-year climatology windows (1865-1894, 1895-1924, ...
@@
 Identification of the complete parameter set of a land surface model is in itself a non-trivial exercise, as in practice, many empirical constants are hard-coded and not amenable to systematic perturbation \cite{mendoza2015,cuntz2016}.
-For the purposes of this activity, we identified a broad set of 211 CLM-BGC parameters, and enabled their modification via the model input files.
-Several domains were not perturbed, including crops, CH$_4$, VOC, and dust emissions, urban properties, and river flow parameters.
+For the purposes of this activity, we identified a broad set of 211 CLM-BGC parameters, and enabled their modification via the model parameter file.
+Several domains were not perturbed, including crops, CH$_4$, emissions of biogenic volitile organic compounds and dust , urban properties, and river flow parameters.
 The crop model was not utilized as our initial focus is on unmanaged land.
@@
 We likewise opted to perturb soil hydraulic parameters directly (e.g.
-saturated hydraulic conductivity) and also indirectly, in a parallel set of simulations via perturbations to the soil texture (percent sand, clay, and silt).
+saturated hydraulic conductivity) and also indirectly, in a parallel set of simulations via perturbations to the soil texture (percent sand, silt, and clay).
 
@@
 $, or increased nitrogen deposition.
-In total 16 of the 211 parameters had the parameter range constrained to some extent. Original and constrained ranges are available in a supplemental materials spreadsheet.
+In total, 16 of the 211 parameters had the parameter range constrained to some extent.
+Original and constrained ranges are available in a supplemental materials spreadsheet.
 
@@
 Each simulation ran for 150 years, with the first 140 for spin-up (as described above in Section \ref{sect:mcn}), followed by a 10-year period for analysis.
-It is important to note that these are equilibrium-style simulations, where the land has been allowed to reach steady state conditions relative to the climate forcing.
+It is important to note that these are equilibrium-style simulations, where the land has been allowed to reach steady state conditions, relative to the climate forcing.
 This is achieved by repeating a single 10-year forcing dataset 15 times during the spin-up procedure and production phase of our simulations.
-As such these simulations are not directly comparable to transient simulations or satellite observations, where present-day vegetation has not yet reached equilibrium with present-day CO$_2$.
+As such, these simulations are not directly comparable to transient simulations or satellite observations, where present-day vegetation has not yet reached equilibrium with present-day climate and CO$_2$ concentrations.
 
@@
 In addition to analyzing parameter effects on various output variables globally, we were also interested in understanding parameter effects on different biomes.
-As such, we categorized each of the 400 sparse grid cells according to their Whittaker biome, which delineates biomes based on average temperature and precipitation \cite{whittaker1970}. 
+As such, we categorized each of the 400 sparse grid cells according to their Whittaker biome \cite{whittaker1970}, which delineates biomes based on average temperature and precipitation.
 Biomes were delineated based on the atmospheric forcing temperature and precipitation data averaged across 2005 to 2014 (Supp Figure S6).
@@
 Combining the SASU spinup approach with the sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
-There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM5 simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
+There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM5.1 simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
 Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for this experiment within 140 years.
@@
 Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, convergence slows and presents some small amplitude oscillations.
-We chose 400 clusters, because it provides satisfactory performance across a range of important metrics, while affording us sufficient simulations to perform the full experiment within our computational budget.
+We chose 400 clusters, because it provides satisfactory performance across a range of important metrics, while affording a sufficient number of simulations to perform the full experiment within our computational budget.
 \begin{figure}[h]
@@
 
-This model configuration allowed for over 2000 simulations across six forcing scenarios.
-Overall, we found substantial impacts of parametric uncertainty on model behavior, in some cases exceeding the magnitude of climate scenario effects (Figure \ref{fig:ranges}).
-In the control scenario, gross primary productivity (GPP) ranged from 103 to 158 PgC, and net ecosystem productivity (NEP) ranged from 1.7 to 3.8 PgC.
-NEP was especially variable in the high CO$_2$ scenario, varying from 1.6 to 5.7 PgC.
+This model configuration allowed for over 2000 simulations across each of the six forcing scenarios.
+Overall, we found substantial impacts of parametric uncertainty on model behavior, in some cases exceeding the magnitude of forcing scenario effects (Figure \ref{fig:ranges}).
+In the control scenario, the global sum of gross primary production (GPP) ranged from 103 to 158 PgC y-1 across all ensemble members, and net ecosystem production (NEP) ranged from 1.7 to 3.8 PgC y-1.
+NEP was especially variable in the high CO$_2$ scenario, varying from 1.6 to 5.7 PgC y-1.
 One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death.
@@
 For example, just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S3).
-Conversely, GPP, which is a significantly more complicated process, requires 20 parameters to explain 90\% of the control ensemble variance.
+Conversely, GPP, which is a significantly more complicated process, requires 20 parameters to explain 90\% of the control ensemble variance (Sorry, what figure is this sentence referring too?).
 In general, most of the parameter perturbations had a relatively small impact on any given output variable.
@@
 \includegraphics[width=\textwidth]{../figs/variance.png}
-\caption{Cumulative fraction of variance explained by the most influential parameters on GPP and canopy evaporation.
+\caption{Cumulative fraction of variance explained by the most influential parameters on (a) canopy evaporation and (b) GPP .
 Approximately 96\% of canopy evaporation variance can be explained by the ten most influential parameters, whereas 30 parameters are required to explain 96\% of GPP variance.}
@@
 We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps}.
-This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control the response to forcing (CO2, nitrogen deposition) and climate.
+This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control the response to forcing changes (CO$_2$, nitrogen deposition and climate).
 For example, the parameters that control present-day NEP differ  from the parameters that control NEP in the high CO$_2$ scenario (Figure \ref{fig:nep}a,b).
@@
 The future climate experiment is 4.3K warmer averaged over our study domain (land-only, Antarctica excluded, Supp Figure S4).
-This led to an increase in evapotranspiration of 8.8\% relative to the control simulation when using the default CLM5.1 parameters (Figure \ref{fig:nep}f).
+This led to an increase in evapotranspiration (ET) of 8.8\% relative to the control simulation when using the default CLM5.1 parameters (Figure \ref{fig:nep}f).
 While the parameters that most strongly affect global average ET in both the control and future climates are generally the same (Figure \ref{fig:nep}d,e), a distinct set of parameters govern the change in ET due to warming (Figure \ref{fig:nep}f).
@@
 \includegraphics[width=\textwidth]{../figs/main/deltas.png}
-\caption{The eight most influential parameters on NEP in the control simulations (a), high CO$_2$ simulations (b), and the relative response to high CO$_2$ (c), as well as on ET in the control simulations (d), future climate simulations (e) and the relative response to future climate (f).
+\caption{The eight most influential parameters on NEP (top row) in the control simulations (a), high CO$_2$ simulations (b), and the relative response to high CO$_2$ (c), as well as on evapotranspiration (ET, bottom row) in the control simulations (d), future climate simulations (e) and the relative response to future climate (f).
 The parameters that are most influential on present-day climate (a,d) can differ significantly from the parameters that control the response to forcing perturbations (c,f).}
@@
 \includegraphics[width=\textwidth]{../figs/lai_biome.pdf}
-\caption{The eight most influential parameters on leaf area index within the control ensemble, globally and within three biomes.
+\caption{The eight most influential parameters on leaf area index (LAI) within the control ensemble, globally and within three biomes.
 The solid lines span the range between the two simulations for each parameter, with an open circle indicating the simulation where a parameter is set to its minimum value and a filled circle for its maximum.
-The lines are color-coded according to scientific domain.
+The lines are color-coded according to scientific domain (Fig X).
 The dashed lines indicate the LAI for the simulation with default CLM5.1 parameters.
@@
 In this project, we identified and perturbed 211 CLM parameters to create a large one-at-a-time PPE.
-There were several barriers to perturbing the full set of CLM5-BGC parameters.
+There were several barriers to perturbing the full set of CLM-BGC parameters.
 First, many parameters had not been officially identified as parameters.
@@
 Defining appropriate ranges for the full suite of land model parameters is a major activity of the scientific subdomains that comprise land models \cite{kattge2020}.
-Even with parameters that have an empirical basis from field studies, they may behave differently at the coarse climate modeling spatial resolution (zqz), complicating the process of uncertainty quantification.
+Even with parameters that have an empirical basis from field studies, they may behave differently at the coarse spatial resolution used by climate models (zqz), complicating the process of uncertainty quantification.
 As such, we cannot claim that the parameters are equivalently sampled, whereby parameter effects and parameter rankings could be subject to sampling asymmetries.
@@
 For this experiment we opted for a one-at-a-time perturbation strategy, testing a minimum and maximum value for each parameter.
-We found that parameter effects could be quite large, in some cases exceeding the effects of our various forcing scenarios (Figure \ref{fig:ranges}).
+We found that parameter effects could be quite large, in some cases exceeding the effects of our forcing scenarios (Figure \ref{fig:ranges}).
 That said, the majority of parameter perturbations had small effects for any specific model variable or metric, such that a majority of simulations were clustered around the default simulation.
@@
 Just as the most influential parameters varied by forcing scenario, parameter rankings varied significantly by biome (Figure \ref{fig:lai}).
-Different PFTs present varying trait strategies, and different climates inflict unique challenges to vegetation activity, such as light availability, extreme temperatures, or water stress.
-As such, perturbations that serve to ease or exacerbate the most substantial stressor for the set of PFTs in a given biome will have greater effect. 
+Different PFTs present varying trait strategies, and different climates impart unique challenges to vegetation activity, such as light availability, extreme temperatures, or water stress.
+As such, perturbations that serve to ease or exacerbate the most substantial environmental stress for the set of PFTs in a given biome will have greater effect.
 Understanding the regional signatures of parameter perturbations is important for model calibration.
@@
 This could be useful for investigating obvious model deficiencies, such as regional LAI biases, or to understand potential model responses under different forcing scenarios (e.g.
-plant survival under low CO2 conditions).
+plant survival under low CO$_2$ conditions).
 
@@
 DK is supported via the Drought Task Force IV organized by the NOAA Climate Program Office’s Modeling, Analysis, Predictions, and Projections.
+KD is supported by the U.S.
+Department of Energy, Office of Science, Office of Biological \& Environmental Research (BER), Regional and Global Model Analysis (RGMA) component of the Earth and Environmental System Modeling Program under Award Number DE-SC0022070 and National Science Foundation (NSF) IA 1947282.
 FL is supported by the National Key Research and Development Program of China (2022YFE0106500).

### 2024-09-11 ms/ms.tex
fix
refs
overleaf

@@
-†isl
 \documentclass[draft]{agujournal2019}
@@
 \begin{document}
-\title{The Community Land Model, version 5.1 \\ One-at-a-time Parameter Perturbation Ensemble}
+\title{One-at-a-time Parameter Perturbation Ensemble of the Community Land Model, version 5.1}
 
@@
 \affiliation{6}{School of Atmospheric Sciences, Sun Yat-sen University, Guangzhou, Guangdong, China}
-\affiliation{7}{Department of Atmospheric Sciences, University of Washington, Seattle, WA, USA}
+\affiliation{7}{Department of Atmospheric and Climate Science, University of Washington, Seattle, WA, USA}
 \affiliation{8}{Research Applications Laboratory, National Center for Atmospheric Research, Boulder, CO, USA}
@@
 \item We constructed a large parameter perturbation ensemble of the Community Land Model, version 5.1, perturbing 211 parameters independently across six forcing scenarios.
-\item Parameter effects can exceed scenario effects and parameter rankings vary by biome and based on the forcing scenario.
+\item Parameter effects can exceed scenario effects and parameter effect rankings differ by biome and based on the forcing scenario.
 \item The software infrastructure developed for this experiment has greatly reduced the human and computer time needed for constructing this type of ensemble, which can facilitate more routine PPEs.
@@
 Comprehensive land models are subject to significant parametric uncertainty, which can be hard to quantify due to the large number of parameters and high model computational costs.
-We constructed a large parameter perturbation ensemble (PPE) for the Community Land Model with biogeochemistry configuration (CLM-BGC).
+We constructed a large parameter perturbation ensemble (PPE) for the Community Land Model vesion 5.1 with biogeochemistry configuration (CLM5.1-BGC).
 We performed more than 2000 simulations perturbing 211 parameters across six forcing scenarios.
@@
 We found that parameter effects can exceed scenario effects and that a small number of parameters explains a large fraction of variance across our ensemble.
-The most important parameters can vary regionally and also based on forcing scenario.
+The most important parameters can differ regionally and also based on the forcing scenario.
 The software infrastructure developed for this experiment has greatly reduced the human and computer time needed for CLM PPEs, which can facilitate routine investigation of parameter sensitivity and uncertainty, as well as automated calibration.
@@
 
-The process of parameter estimation for land models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity of plant and microbial life, and the multiplicity of domains that comprise the terrestrial biosphere (e.g, hydrology, snow, biogeochemistry, physiology, land management, etc.).
-The number of model parameters required to represent all the relevant processes is large and model simulations are relatively expensive; presenting a major barrier to robust objective calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
+The process of parameter estimation for the complex land models typically embedded within Earth System Models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that comprise the terrestrial biosphere (e.g, hydrology, snow, biogeochemistry, physiology, land management, etc.).
+The number of model parameters required to represent all the relevant processes is large and model simulations are relatively expensive, presenting a major barrier to robust objective calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
 Thorough assessment of the parametric sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}.
@@
 
-In this project we create a large PPE with the Community Land Model with biogeochemistry configuration (CLM-BGC)\cite{lawrence2019}.
+In this project we create a large PPE with the Community Land Model 5.1 with biogeochemistry configuration (CLM5.1-BGC, hereafter CLM)\cite{lawrence2019}.
 We extend the work of \citeA{dagon2020} which utilized a PPE to calibrate a subset of CLM parameters.
@@
 This experiment utilizes the Community Land Model configuration (version 5.1, i.e.
-CLM5.1).
+CLM5.1) of the Community Terrestrial Systems Model.
 The model source code and documentation are available online (\url{https://github.com/ESCOMP/CTSM}), as is a full model description \cite{lawrence2019}.
@@
 This is as opposed to the satellite phenology mode, which is driven by observed canopy properties and thus has fewer internal feedbacks, and was explored in \citeA{dagon2020}.
-We used the model in land-only mode driven by a data atmosphere from the Global Soil Wetness Project (GSWP3; REF), with the crop model turned off.
+We used the model in land-only mode driven by a data atmosphere (see Section \ref{sect:exp} for details), with the crop model turned off.
 
@@
 This protocol was designed to achieve sufficiently equilibrated carbon and nitrogen states, while minimizing computational time.
-This spin-up methodology did not always reach full equilibration of deep soil carbon (beyond 1 meter depth) in very few points at high latitude.
-Certain inferences about deep soil carbon from this ensemble may therefore subject to uncertainty due to spin-up concerns.
+This spin-up methodology did not always reach full equilibrium of soil carbon (Supp Figure S1).
+However we believe that this spinup methodology is sufficient to robustly infer parameter rankings for carbon cycle processes, if not the exact absolute values of parameter effects.
 
@@
 To reduce computational costs parameter perturbation experiments often use lower resolution, such as 4$^\circ$x5$^\circ$ \cite{dagon2020}.
-As opposed to this coarse grid resolution, here we apply an alternative sparse grid approach by useing a clustering algorithm to achieve an alternative low resolution configuration.
+As opposed to selecting a coarse grid resolution, here we apply an alternative sparse grid approach by using a clustering algorithm to achieve an alternative low resolution configuration.
 The goal is to more strategically distribute grid cells, opting for more grid cells in areas that are especially distinct or variable, and fewer in areas that can be easily approximated based on output elsewhere (e.g.
 neighboring grid cells).
-
 Multivariate spatio-temporal clustering (MVSC) has been utilized to extract patterns of climatological significance from climate model output \cite{hoffman2005} and applied to design a representativeness-based sampling network \cite{hoffman2013}.
-Instead of lowering resolution by coarsening a rectilinear grid, we here used MVSC to strategically remove effectively redundant grid cells.We used k-means clustering to identify groups of grid cells with similar dynamics based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
+Instead of lowering resolution by coarsening a rectilinear grid, we used MVSC to strategically remove effectively redundant grid cells.
+We used k-means clustering to identify groups of grid cells with similar dynamics based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase.
 We selected one representative grid cell from each cluster to stand in for the entire cluster.
 The representative grid cell is whichever is located nearest the cluster centroid in climate space.
-The set of representative grid cells comprise a `sparsegrid', which are used in lieu of a `coarse' grid.
+The set of representative grid cells comprise a `sparse grid', which are used in lieu of a `coarse' grid.
 To recompose mapped output and compute global means, the output from the representative grid cell is substituted for all members of the cluster cohort.
@@
 We tested 15 values of $k$, ranging from 10 to 800.
-Utilizing the ILAMB2.5 benchmarking software \cite{collier2018}, we calculated skill scores to quantify how well each candidate sparsegrid mirrored the full grid output (Supp Figure S1).
+Utilizing the ILAMB2.5 benchmarking software \cite{collier2018}, we calculated skill scores to quantify how well each candidate sparsegrid mirrored the full grid output (Supp Figure S2).
 We opted for a 400-cluster sparsegrid, to balance computational cost against model fidelity.
@@
 For the purposes of this activity, we identified a broad set of 211 CLM-BGC parameters, and enabled their modification via the model parameter file.
-Several domains were not perturbed, including crops, CH$_4$, emissions of biogenic volitile organic compounds and dust , urban properties, and river flow parameters.
+Several domains were not perturbed, including crops, methane emissions, emissions of biogenic volatile organic compounds and dust, urban properties, and river flow parameters.
 The crop model was not utilized as our initial focus is on unmanaged land.
@@
 In some cases literature values were directly utilized, but in many cases, expert judgment was used or else a default perturbation of $\pm$20\%.
-The parameter range spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.
+The parameter range spreadsheet is provided within the Supplementary Information.
 In some cases parameters could not or should not be perturbed independently, because they feature inherent co-dependence.
 In the case of nitrogen fixation costs, we opted to perturb the parameters independently, but also as a group (`KCN'), to enforce a change in overall nitrogen limitation in lieu of switching from one uptake pathway to another.
-We likewise opted to perturb soil hydraulic parameters directly (e.g.
-saturated hydraulic conductivity) and also indirectly, in a parallel set of simulations via perturbations to the soil texture (percent sand, silt, and clay).
+We likewise opted to perturb soil hydraulics directly in two distinct ways, first by directly perturbing the hydraulic properties (e.g.
+saturated hydraulic conductivity) and also then also in a parallel set of simulations perturbing the soil texture (percent sand, silt, and clay).
 
-For a small percentage of parameters, the perturbation ranges were severe enough to eradicate a given PFT across all or part of its geographical range (Supp Figure S2).
+For a small percentage of parameters, the perturbation ranges were severe enough to eradicate a given PFT across all or part of its geographical range (Supp Figure S3).
 In the case that one or more PFTs did not survive a parameter perturbation, we iteratively re-ran the simulation with slightly less extreme parameter values until the PFT survived.
@@
 In total, 16 of the 211 parameters had the parameter range constrained to some extent.
-Original and constrained ranges are available in a supplemental materials spreadsheet.
+Original and constrained ranges are available in a supplemental information spreadsheet.
+
+In one case, a significant CLM bug was identified after the completion of our primary experiment (\url{https://github.com/ESCOMP/CTSM/issues/2120}.
+Namely that the values for two nitrogen uptake parameters had been transposed in the
 
@@
 The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, as it is the default forcing data for CLM5 \cite{lawrence2019}.
+Our six forcing scenarios were: a 2010-era control climate, high and low CO$_2$, future and pre-industrial climate anomaly forcing, and enhanace nitrogen deposition (see Table \ref{tab:exps}).
 We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
-Future and pre-industrial climate forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to the corresponding mean climate change signal (see Table \ref{tab:exps}; AF2095 and AF1855).
-We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest and present day for six atmospheric forcing variables (temperature, humidity, precipitation, wind, longwave and shortwave radiation; see Supp Figure S3 for climatologies).
+Future and pre-industrial climate anomaly forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to a mean climate change signal.
+We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest (either 1850s or 2090s) and present day for six atmospheric forcing variables (temperature, humidity, precipitation, wind, longwave and shortwave radiation; see Supp Figure S4 for climatologies).
 The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, which was chosen to align with the existing CESM2 large ensemble \cite{rodgers2021}.
 Since completing the experiment we uncovered a software bug affecting the AF1855 and AF2095 simulations.
-Namely that anomalies were not applied to certain coastal gridcells (Supp Fig S4).
-In all this affected less than 3\% of land area, and did not significantly alter parameter rankings (Supp Fig S5).
+Namely that climate anomalies were not applied to certain coastal gridcells (Supp Fig S5).
+In all this affected less than 3\% of land area, and did not significantly alter parameter rankings (Supp Fig S6).
 
@@
 Just as the most influential parameters varied by forcing scenario, parameter rankings varied significantly by biome (Figure \ref{fig:lai}).
-Different PFTs present varying trait strategies, and different climates impart unique challenges to vegetation activity, such as light availability, extreme temperatures, or water stress.
+Different PFTs present varying trait strategies, and different climates impose unique challenges to vegetation activity, such as light availability, extreme temperatures, or water stress.
 As such, perturbations that serve to ease or exacerbate the most substantial environmental stress for the set of PFTs in a given biome will have greater effect.
@@
 
-The primary deliverable of this project is the PPE dataset that captures the effects of 212 parameters on a wide range of CLM variables across six forcing scenarios.
+The primary deliverable of this project is the PPE dataset that captures the effects of 211 parameters on a wide range of CLM variables across six forcing scenarios.
 Several spin-off projects have leveraged this dataset, including some that have already reached publication \cite{cheng2023,yan2023a,yan2023b}.
@@
 All of the figures on our diagnostics website can be made from this one file.
-The raw dataset will be available on the NSF-NCAR computing system, and can be made available online as well, but we are gauging interest before committing to finding a host for the 2TB of data.
+The raw dataset will be available on the NSF-NCAR computing system, and can be made available online as well, but we are gauging interest before committing to host the full 2TB dataset.
 
@@
 \acknowledgments
-This material is based upon work supported by the National Center for Atmospheric Research (NCAR), which is a major facility sponsored by the National Science Foundation under Cooperative Agreement No.
+This material is based upon work supported by the NSF National Center for Atmospheric Research, which is a major facility sponsored by the National Science Foundation under Cooperative Agreement No.
 1852977.
-Computing and data storage resources, including the Cheyenne supercomputer (doi:10.5065/D6RX99HX) were provided by the Climate Simulation Laboratory at NCAR’s Computational and Information Systems Laboratory (CISL).
-DK is supported via the Drought Task Force IV organized by the NOAA Climate Program Office’s Modeling, Analysis, Predictions, and Projections.
+Computing and data storage resources, including the Cheyenne supercomputer (doi:10.5065/D6RX99HX) were provided by the Climate Simulation Laboratory at NSF-NCAR’s Computational and Information Systems Laboratory (CISL).
+DK and DML acknowledge funding from the National Oceanic and Atmospheric Administration (NOAA) Modelling, Analysis, Predictions and Projections (MAPP) program (award number NA20OAR4310413) and funding from the National Center for Atmospheric Research, which is a major facility sponsored by the NSF under Cooperative Agreement No.
+1852977.
 KD is supported by the U.S.
 Department of Energy, Office of Science, Office of Biological \& Environmental Research (BER), Regional and Global Model Analysis (RGMA) component of the Earth and Environmental System Modeling Program under Award Number DE-SC0022070 and National Science Foundation (NSF) IA 1947282.
+ALSS recognizes support from the National Science Foundation (NSF) (AGS-1553715, AGS-2330096).
 FL is supported by the National Key Research and Development Program of China (2022YFE0106500).

### 2024-09-11 ms/supp.tex
overleaf

@@
 \maketitle
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=40pc]{../figs/supp/spinup.png}
+\caption{Independent spinup was performed for each simulation of the ensemble.
+Whereas gross primary production (GPP) and total water storage (TWS) equilibrated quickly, total carbon soil organic matter in the top meter of the soil column (TOTSOMC\_1m) required longer spinup and did not always reach equilibrium.}
+\label{supp:spinup}
+\end{figure}
 

### 2024-09-11 ms/ms.tex
refs
overleaf

@@
 
-In one case, a significant CLM bug was identified after the completion of our primary experiment (\url{https://github.com/ESCOMP/CTSM/issues/2120}.
-Namely that the values for two nitrogen uptake parameters had been transposed in the
+In one case, a significant CLM bug was identified after the completion of our primary experiment (\url{https://github.com/ESCOMP/CTSM/issues/2120}).
+Namely that the values for two nitrogen uptake parameters had been transposed in \citeA{brzostek2014}.
+Resolving the issue resulted in an approximately 5\% decrease in gross primary production of (Supp Figure S4) in the control simulations, and also significantly altered the size of the various parameter effects.
+Any PPE should be seen as intrinsically connected to its underlying codebase (bugs and all), and, reflecting this, we did not rerun the rest of the PPE.
+Therefore the majority of simulations contain the erroneous parameter values, and may slightly overestimate GPP.
+The bug has since been resolved in later model versions.
 
@@
 Future and pre-industrial climate anomaly forcing datasets were prepared by adding GSWP3v1 anomalies from 2005-2014 to a mean climate change signal.
-We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest (either 1850s or 2090s) and present day for six atmospheric forcing variables (temperature, humidity, precipitation, wind, longwave and shortwave radiation; see Supp Figure S4 for climatologies).
+We inferred the mean climate change signal using the CESM2 large ensemble experiment \cite{rodgers2021}, computed as the average of the difference between the period of interest (either 1850s or 2090s) and present day for six atmospheric forcing variables (temperature, humidity, precipitation, wind, longwave and shortwave radiation; see Supp Figure S5 for climatologies).
 The future climate scenario utilizes the end-of-century climate from the SSP3-7.0, a high-emissions scenario, which was chosen to align with the existing CESM2 large ensemble \cite{rodgers2021}.
 Since completing the experiment we uncovered a software bug affecting the AF1855 and AF2095 simulations.
-Namely that climate anomalies were not applied to certain coastal gridcells (Supp Fig S5).
-In all this affected less than 3\% of land area, and did not significantly alter parameter rankings (Supp Fig S6).
+Namely that climate anomalies were not applied to certain coastal gridcells (Supp Fig S6).
+In all this affected less than 3\% of land area, and did not significantly alter parameter rankings (Supp Fig S7).
 
@@
 As such, we categorized each of the 400 sparse grid cells according to their Whittaker biome \cite{whittaker1970}, which delineates biomes based on average temperature and precipitation.
-Biomes were delineated based on the atmospheric forcing temperature and precipitation data averaged across 2005 to 2014 (Supp Figure S6).
-Overall there are nine vegetated biomes and an additional ice sheet biome, where no vegetation is present (Supp Figures S7,8).
+Biomes were delineated based on the atmospheric forcing temperature and precipitation data averaged across 2005 to 2014 (Supp Figure S8).
+Overall there are nine vegetated biomes and an additional ice sheet biome, where no vegetation is present (Supp Figures S9,10).
 
 \section{Results}
-
-Given the very large size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the relatively expensive computational cost of standard model configurations, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations.
-Combining the SASU spinup approach with the sparsegrid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
-There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM5.1 simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
-Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for this experiment within 140 years.
-
 \begin{figure}[h]
@@
 \end{figure}
+Given the very large size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the relatively expensive computational cost of standard model configurations, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations.
+Combining the SASU spinup approach with the sparse grid formulation (see Section \ref{sect:sg} for details) yielded a configuration approximately 500 times less costly than the 1-degree configuration most often used for CLM simulations (Figure \ref{fig:sims}).
+There are 22648, 5666, and 1764 land grid cells in standard 1$^{\circ}$, 2$^{\circ}$, and 4$^{\circ}$x5$^{\circ}$ CLM5.1 simulations, respectively, as compared to just 400 grid cells in the sparsegrid.
+Likewise, whereas previous spin-up methodologies required 1500 years or longer to satisfy equilibrium criteria, the SASU approach enabled by matrix representation yielded satisfactory spin-up efficiency for this experiment within 140 years (Supp Figure S1).
 
@@
 We had hoped to see a clear point of diminishing returns across a wide range of variables as we increase the number of clusters.
-Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, convergence slows and presents some small amplitude oscillations.
+Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, convergence slows and presents some small amplitude oscillations (Supp Figure S2).
 We chose 400 clusters, because it provides satisfactory performance across a range of important metrics, while affording a sufficient number of simulations to perform the full experiment within our computational budget.
@@
 
-This model configuration allowed for over 2000 simulations across each of the six forcing scenarios.
-Overall, we found substantial impacts of parametric uncertainty on model behavior, in some cases exceeding the magnitude of forcing scenario effects (Figure \ref{fig:ranges}).
-In the control scenario, the global sum of gross primary production (GPP) ranged from 103 to 158 PgC y-1 across all ensemble members, and net ecosystem production (NEP) ranged from 1.7 to 3.8 PgC y-1.
-NEP was especially variable in the high CO$_2$ scenario, varying from 1.6 to 5.7 PgC y-1.
+This model configuration allowed for over 2000 simulations across our six forcing scenarios.
+Overall, we found substantial impacts of parametric uncertainty on model behavior, in some cases exceeding the magnitude of scenario effects (Figure \ref{fig:ranges}).
+In the control scenario, the global sum of gross primary production (GPP) ranged from 103 to 158 PgC per year across all ensemble members, and net ecosystem production (NEP) ranged from 1.7 to 3.8 PgC per year.
+NEP was especially variable in the high CO$_2$ scenario, varying from 1.6 to 5.7 PgC per year.
 One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death.
@@
 A small fraction of parameters tends to explain a large amount of the ensemble variance for any given variable or metric (Figure \ref{fig:variance}).
-For example, just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S3).
-Conversely, GPP, which is a significantly more complicated process, requires 20 parameters to explain 90\% of the control ensemble variance (Sorry, what figure is this sentence referring too?).
+GPP, which is a complicated process, requires 20 parameters to explain 90\% of the control ensemble variance.
+Conversely, canopy evaporation is governed by fewer parameters.
+Just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S3).
 In general, most of the parameter perturbations had a relatively small impact on any given output variable.
@@
 \includegraphics[width=\textwidth]{../figs/variance.png}
-\caption{Cumulative fraction of variance explained by the most influential parameters on (a) canopy evaporation and (b) GPP .
+\caption{Cumulative fraction of variance explained by the most influential parameters on gross primary production and canopy evaporation.
 Approximately 96\% of canopy evaporation variance can be explained by the ten most influential parameters, whereas 30 parameters are required to explain 96\% of GPP variance.}
@@
 
-We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps}.
+We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps} (Figure \ref{fig:ranges}).
 This ensures that we can identify parameters that are important not just under present-day conditions, but also parameters that control the response to forcing changes (CO$_2$, nitrogen deposition and climate).

### 2024-09-11 ms/ms.tex
overleaf

@@
 One key parameter is \textit{tpuse\_sf}, which is a scalar perturbation factor influencing the triose phosphate limitations on photosynthesis.
-Triose phosphate limitation is difficult to quantify and has not been observed to significantly hinder photosynthesis under present-day CO$_2$ concentrations \cite{kumarathunge2019}.
-In CLM5.1, this parameter is not particularly influential on NEP at 367 ppm (ranked 39, not shown), but is third most influential with high CO$_2$.
+Triose phosphate limitation is difficult to quantify and has not been observed to significantly hinder photosynthesis under present-day CO$_2$ concentrations, but may play a larger role in the future \cite{lombardozzi2018,kumarathunge2019}.
+In CLM5.1, this parameter is not particularly influential on NEP at 367 ppm (ranked 39, not shown), but is third most influential when we increased CO$_2$ to 867 ppm.
 
-The future climate experiment is 4.3K warmer averaged over our study domain (land-only, Antarctica excluded, Supp Figure S4).
+The future climate experiment is 4.3K warmer averaged over our study domain (land-only, Antarctica excluded, Supp Figure S5).
 This led to an increase in evapotranspiration (ET) of 8.8\% relative to the control simulation when using the default CLM5.1 parameters (Figure \ref{fig:nep}f).

### 2024-09-13 ms/ms.tex
overleaf

@@
 Triose phosphate limitation is difficult to quantify and has not been observed to significantly hinder photosynthesis under present-day CO$_2$ concentrations, but may play a larger role in the future \cite{lombardozzi2018,kumarathunge2019}.
-In CLM5.1, this parameter is not particularly influential on NEP at 367 ppm (ranked 39, not shown), but is third most influential when we increased CO$_2$ to 867 ppm.
+This parameter is not particularly influential on NEP at 367 ppm (ranked 39, not shown), but is third most influential when we increased CO$_2$ to 867 ppm.
 
 The future climate experiment is 4.3K warmer averaged over our study domain (land-only, Antarctica excluded, Supp Figure S5).
-This led to an increase in evapotranspiration (ET) of 8.8\% relative to the control simulation when using the default CLM5.1 parameters (Figure \ref{fig:nep}f).
+This contributed to an increase in evapotranspiration (ET) of 8.8\% relative to the control simulation when using the default CLM5.1 parameters (Figure \ref{fig:nep}f).
 While the parameters that most strongly affect global average ET in both the control and future climates are generally the same (Figure \ref{fig:nep}d,e), a distinct set of parameters govern the change in ET due to warming (Figure \ref{fig:nep}f).
@@
 The solid lines span the range between the two simulations for each parameter, with an open circle indicating the simulation where a parameter is set to its minimum value and a filled circle for its maximum.
-The lines are color-coded according to scientific domain (Fig X).
+The lines are color-coded according to scientific domain.
 The dashed lines indicate the LAI for the simulation with default CLM5.1 parameters.
@@
 However it is also possible to inspect the geographic footprint of parameter perturbations by projecting the sparsegrid output to standard lat/lon coordinates (see Section \ref{sect:sg} for details).
-For example, perturbing the \textit{medlynslope} parameter has a large effect on global runoff, but primarily via its effects in vegetated areas (Figure \ref{fig:map}).
+For example, perturbing the \textit{medlynslope} parameter (which controls stomatal conductance) has a large effect on global runoff, but primarily via its effects in vegetated areas (Figure \ref{fig:map}).
 Because the number of potential variables, parameters, and geographical ranges of interest to the wider CLM community is larger than we can document here,  we provide a tool that can be used to explore an extended diagnostics set which summarises the $>$2TB of output data via approximately 2000 plots (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}).
@@
 \includegraphics[width=\textwidth]{../figs/QRUNOFFabs_x_medlynslope_CTL2010.png}
-\caption{Map of the effect of perturbing medlynslope on runoff within the control ensemble.
+\caption{Map of the effect of perturbing the medlynslope stomatal conductance parameter on runoff within the control ensemble.
 Increasing medlynslope tends to reduce runoff, but only in regions with sufficient vegetation activity.
@@
 In such cases, we identified hard-coded values, established an appropriate parameter name, and extracted that parameter to the CLM parameter file for easier manipulation.
-Although in this experiment we perturbed a large number of parameters across a variety of CLM processes (Figure \ref{fig:params}), this still does not cover the full set of CLM parameters, as some processes were not included, such as crop phenology and management.
+Though we perturbed a large number of parameters across a variety of CLM processes (Figure \ref{fig:params}), this still does not cover the full set of CLM parameters, as some processes were not included, such as crop phenology and management.
 In carrying out this process of parameter identification, we likewise unearthed many nuances in the epistemology of what `parameters' are in the context of Earth system models.
@@
 Defining appropriate ranges for the full suite of land model parameters is a major activity of the scientific subdomains that comprise land models \cite{kattge2020}.
-Even with parameters that have an empirical basis from field studies, they may behave differently at the coarse spatial resolution used by climate models (zqz), complicating the process of uncertainty quantification.
+Even with parameters that have an empirical basis from field studies, uncertainty quantification is difficult to the scale mismatch between site-level observations and land model grid cells \cite{liu2021}.
 As such, we cannot claim that the parameters are equivalently sampled, whereby parameter effects and parameter rankings could be subject to sampling asymmetries.
@@
 As such, perturbations that serve to ease or exacerbate the most substantial environmental stress for the set of PFTs in a given biome will have greater effect.
+For example, modifying photosynthetic capacity (via jmaxb0, which sets baseline proportion of nitrogen allocated for electron transport) has a large effect in boreal forests, as light is relatively limiting at high latitudes.
 Understanding the regional signatures of parameter perturbations is important for model calibration.
@@
 Additional work is ongoing to extend this work in a coupled model framework to quantify how much, where, and for which processes atmospheric feedbacks are likely to modulate land parameter effects substantially.
+Furthermore our experiments are conducted at nominal 2-degree resolution, certain processes subject to land heterogeneity may respond differently at higher resolution, such as permafrost depth in the Arctic \cite{schickhoff2024}.
 Our expectation is that relative parameter rankings are likely robust, but that the absolute value of parameter effects may be significantly biased relative to their effects in the coupled model.
@@
 These projects utilized our ensemble to filter for parameters that influence the relevant study domains and variables, and performed follow-on experiments using the parameter ranges collated in Section \ref{sect:exps}.
-
-The infrastructure underpinning this experiment is another major deliverable. Our project has accelerated parameter exploration work within our collaborator network by providing:
+Maintaining, improving, and interpreting complex land models benefits from thoughtful investment in software to automate and routinize important components of the development process, such as the International Land Model Benchmarking (ILAMB) system \cite{collier2018}.
+Therefore we also see the infrastructure underpinning our experiment as a major deliverable. Our project has accelerated parameter exploration work within our collaborator network by providing:
 
@@
 
-We have already begun a follow-on activity that perturbs a subset of important parameters with a more exhaustive sampling approach to work towards calibration of leaf area index.
+We have already started a follow-on activity that perturbs a subset of important parameters with a more exhaustive sampling approach to work towards calibration of leaf area index.
 By investing in the infrastructure that we introduce here, all of our subsequent parameter perturbation experiments have required much less time and effort.
-We continue to extend our efforts in this domain towards model calibration and uncertainty quantification, and we anticipate repeating the foundational one-at-a-time experiment with subsequent model releases.
+We continue to extend our efforts in this domain towards model calibration and uncertainty quantification, and we anticipate repeating this foundational one-at-a-time experiment with subsequent model releases.
 Reducing both the human and computer time required for these experiments is allowing us to repeat them more frequently through the model development cycle.
@@
 
-A relatively small ($\sim$300MB), post-processed dataset is freely available online.
-All of the figures on our diagnostics website can be made from this one file.
+A relatively small ($\sim$300MB), post-processed dataset is freely available online (n.b.
+we will create a DOI and link accordingly once our figures are finalized later in the review process).
+All of the figures on our diagnostics website (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT}) can be made from this one file.
 The raw dataset will be available on the NSF-NCAR computing system, and can be made available online as well, but we are gauging interest before committing to host the full 2TB dataset.
@@
 Computing and data storage resources, including the Cheyenne supercomputer (doi:10.5065/D6RX99HX) were provided by the Climate Simulation Laboratory at NSF-NCAR’s Computational and Information Systems Laboratory (CISL).
-DK and DML acknowledge funding from the National Oceanic and Atmospheric Administration (NOAA) Modelling, Analysis, Predictions and Projections (MAPP) program (award number NA20OAR4310413) and funding from the National Center for Atmospheric Research, which is a major facility sponsored by the NSF under Cooperative Agreement No.
+DK and DML acknowledge funding from the National Oceanic and Atmospheric Administration (NOAA) Modelling, Analysis, Predictions and Projections (MAPP) program (award number NA20OAR4310413) and funding from the NSF National Center for Atmospheric Research, which is a major facility sponsored by the NSF under Cooperative Agreement No.
 1852977.
@@
 \section{Supplementary Figures}
-
-extra text:
-Maintaining, improving, and interpreting complex land models benefits from thoughtful investment in software to automate and routinize important components of the development process, e.g.
-\citeA{collier2018}.
 

### 2024-09-13 ms/ms.tex
overleaf

@@
 \begin{keypoints}
-\item We constructed a large parameter perturbation ensemble of the Community Land Model, version 5.1, perturbing 211 parameters independently across six forcing scenarios.
+\item We constructed a parameter perturbation ensemble of the Community Land Model, v5.1, perturbing 211 parameters across six forcing scenarios.
 \item Parameter effects can exceed scenario effects and parameter effect rankings differ by biome and based on the forcing scenario.
-\item The software infrastructure developed for this experiment has greatly reduced the human and computer time needed for constructing this type of ensemble, which can facilitate more routine PPEs.
+\item The software infrastructure developed in our experiment has greatly reduced the human and computer time needed for constructing future PPEs.
 \end{keypoints}
@@
 Not all parameters are currently utilizing this functionality, such as \textit{medlynintercept} where the default value is 100 $\mu$mol H$_2$O m$^{-2}$ s$^{-1}$ for all sixteen of CLM's natural vegetation PFTs.
+
+\begin{landscape}
+ \begin{table}[h]
+ \caption{In total, 211 parameters were perturbed in this experiment.
+These are the parameters mentioned within the main text.
+A complete set of parameters is included in a spreadsheet in the supplementary information.}
+ \centering
+ \begin{tabular}{l l c}
+ \hline
+  Parameter  & Description & Model Domain \\
+ \hline
+cv & Turbulent transfer coefficient & Sensible, latent heat and momentum fluxes \\
+dleaf & Leaf characteristic length & Sensible, latent heat and momentum fluxes \\
+d\_max & Dry surface layer (DSL) parameter & Sensible, latent heat and momentum fluxes \\
+frac\_sat\_soil\_dsl\_init & Fraction of saturated soil at which DSL initiates & Sensible, latent heat and momentum fluxes \\
+fff & Decay factor for fractional saturated area & Hydrology \\
+liq\_canopy\_storage\_scalar & Canopy-storage-of-liquid-water parameter & Hydrology \\
+maximum\_leaf\_wetted\_fraction & Maximum leaf wetted fraction & Hydrology \\
+sand\_pf & Perturbation factor for sand content of soil column & Hydrology \\
+sucsat\_sf & Scale factor for minimum soil water potential & Hydrology \\
+medlynintercept & Medlyn intercept of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
+medlynslope & Medlyn slope of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
+tpu25ratio & Ratio of tpu25top to vcmax25top & Stomatal resistance and photosynthesis \\
+jmaxb0 & Baseline proportion of nitrogen allocated for electron transport & Photosynthetic capacity\\
+jmaxb1 & Response of electron transport rate to light & Photosynthetic capacity\\
+slatop & Specific leaf area at top of canopy & Photosynthetic capacity\\
+theta\_cj & empirical curvature parameter for photosynthesis colimitation& Photosynthetic capacity \\
+wc2wjb0 & Baseline ratio of wc:wj & Photosynthetic capacity \\
+FUN\_fracfixers & fraction of carbon available for N fixation & Fixation and uptake of Nitrogen \\
+KCN & Nitrogen acquisition parameter group & Fixation and uptake of Nitrogen \\
+kmax & Plant segment max conductance & Plant hydraulics \\
+krmax & Root segment max conductance & Plant hydraulics \\
+psi50 & Water potential at 50\% loss of conductance & Plant hydraulics \\
+nstem & Stem number & Biomass heat storage \\
+lmr\_intercept\_atkin & Intercept in the calculation of leaf maintenance respiration& Plant respiration \\
+froot\_leaf & Allocation parameter: new fine root C per new leaf C & Carbon and nitrogen allocation \\
+leafcn & Leaf C:N & Carbon and nitrogen allocation \\
+leaf\_long & Leaf longevity & Vegetation phenology and turnover \\
+cpha & Activation energy for cp & Acclimation parameters \\
+jmaxhd & Deactivation energy for jmax & Acclimation parameters \\
+kcha & Activation energy for kc & Acclimation parameters \\
+lmrha & Activation energy for lmr & Acclimation parameters \\
+lmrhd & Deactivation energy for lmr & Acclimation parameters \\
+tpuha & Activation energy for tpu & Acclimation parameters \\
+tpuse\_sf & Scale factor for tpu entropy term & Acclimation parameters \\
+vcmaxha & Activation energy for vcmax & Acclimation parameters \\
+vcmaxhd & Deactivation energy for vcmax & Acclimation parameters \\
+ \hline
+ \end{tabular}
+ \end{table}
+\end{landscape}
 
@@
 
-\appendix
-\section{Supplementary Figures}
-
-Will be deleted:
-\subsection{Parameters} 
-\begin{landscape}
- \begin{table}[h]
- \caption{Some key parameters}
- \centering
- \begin{tabular}{l l c}
- \hline
-  Parameter  & Description & Model Domain \\
- \hline
-d\_max & Dry surface layer (DSL) parameter & Sensible, latent heat and momentum fluxes \\
-frac\_sat\_soil\_dsl\_init & Fraction of saturated soil at which DSL initiates & Sensible, latent heat and momentum fluxes \\
-fff & Decay factor for fractional saturated area & Hydrology \\
-liq\_canopy\_storage\_scalar & Canopy-storage-of-liquid-water parameter & Hydrology \\
-maximum\_leaf\_wetted\_fraction & Maximum leaf wetted fraction & Hydrology \\
-medlynintercept & Medlyn intercept of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
-medlynslope & Medlyn slope of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
-tpu25ratio & Ratio of tpu25top to vcmax25top & Stomatal resistance and photosynthesis \\
-jmaxb0 & Baseline proportion of nitrogen allocated for electron transport & Photosynthetic capacity (LUNA) \\
-jmaxb1 & Response of electron transport rate to light & Photosynthetic capacity (LUNA) \\
-slatop & Specific leaf area at top of canopy & Photosynthetic capacity (LUNA) \\
-wc2wjb0 & Baseline ratio of wc:wj & Photosynthetic capacity (LUNA) \\
-kmax & Plant segment max conductance & Plant hydraulics \\
-krmax & Root segment max conductance & Plant hydraulics \\
-psi50 & Water potential at 50\% loss of conductance & Plant hydraulics \\
-nstem & Stem number & Biomass heat storage \\
-lmr\_intercept\_atkin & Intercept in the calculation of leaf maintenance respiration& Plant respiration \\
-froot\_leaf & Allocation parameter: new fine root C per new leaf C & Carbon and nitrogen allocation \\
-leafcn & Leaf C:N & Carbon and nitrogen allocation \\
-leaf\_long & Leaf longevity & Vegetation phenology and turnover \\
-cpha & Activation energy for cp & Acclimation parameters \\
-jmaxhd & Deactivation energy for jmax & Acclimation parameters \\
-kcha & Activation energy for kc & Acclimation parameters \\
-lmrha & Activation energy for lmr & Acclimation parameters \\
-lmrhd & Deactivation energy for lmr & Acclimation parameters \\
-tpuha & Activation energy for tpu & Acclimation parameters \\
-tpuse\_sf & Scale factor for tpu entropy term & Acclimation parameters \\
-vcmaxha & Activation energy for vcmax & Acclimation parameters \\
-vcmaxhd & Deactivation energy for vcmax & Acclimation parameters \\
- \hline
- \end{tabular}
- \end{table}
-\end{landscape}
-
 \end{document}

### 2024-10-08 ms/supp.tex
overleaf

@@
 \centering
+\includegraphics[width=\textwidth]{../figs/supp/funbug.png}
+\caption{Parameter rankings for nitrogen uptake parameters subject to a recently uncovered CLM bug. The values for \textit{kn\_nonmyc} and \textit{kc\_nonmyc} had been accidentally transposed.}
+\label{supp:funbug}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 \includegraphics[width=\textwidth]{../figs/supp/anomalies.png}
@@
 \includegraphics[width=30pc]{../figs/supp/AF2095_rankings.png}
-\caption{Comparing parameter rankings with and without the errant gridcells from Supp Figure S4 across a selection of output variables.}
+\caption{Comparing parameter rankings with and without the errant gridcells from Supp Figure S6 across a selection of output variables.}
 \label{supp:abug}

### 2025-03-31 ms/ms.tex
submission version

@@
 These are the parameters mentioned within the main text.
-A complete set of parameters is included in a spreadsheet in the supplementary information.}
+A complete set of parameters, with ranges, is included in a spreadsheet in the supplementary information.}
  \centering
@@
 \includegraphics[width=\textwidth]{../figs/main/deltas.png}
-\caption{The eight most influential parameters on NEP (top row) in the control simulations (a), high CO$_2$ simulations (b), and the relative response to high CO$_2$ (c), as well as on evapotranspiration (ET, bottom row) in the control simulations (d), future climate simulations (e) and the relative response to future climate (f).
+\caption{The eight most influential parameters \add{(ranked by absolute parameter effect)} on NEP (top row) in the control simulations (a), high CO$_2$ simulations (b), and the relative response to high CO$_2$ (c), as well as on evapotranspiration (ET, bottom row) in the control simulations (d), future climate simulations (e) and the relative response to future climate (f).
 The parameters that are most influential on present-day climate (a,d) can differ significantly from the parameters that control the response to forcing perturbations (c,f).}
@@
 \includegraphics[width=\textwidth]{../figs/lai_biome.pdf}
-\caption{The eight most influential parameters on leaf area index (LAI) within the control ensemble, globally and within three biomes.
+\caption{The eight most influential parameters \add{(ranked by absolute parameter effect)} on leaf area index (LAI) within the control ensemble, globally and within three biomes.
 The solid lines span the range between the two simulations for each parameter, with an open circle indicating the simulation where a parameter is set to its minimum value and a filled circle for its maximum.

### 2025-04-02 ms/ms.tex
Updates from Overleaf

@@
 Understanding the range of outcomes arising from the many plausible parameter combinations is a critical step in robust uncertainty quantification.
-Parameter sensitivity tests can be used to gauge parametric uncertainty, and a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \cite{dagon2020,mcneall2024}.
+Parameter sensitivity tests can be used to gauge parametric uncertainty, and a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \add{zqzbaker2022} \cite{dagon2020,mcneall2024}.
 One important application of PPEs has been assessing uncertainty in climate model projections \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through various approaches like history matching \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024} or Bayesian methods \cite{cleary2021}.
+Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through various approaches like history matching, \add{which does not require assumptions regarding parameter distributions} \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024}, or Bayesian methods \add{zqz zeihn2012,fer2018} \cite{cleary2021}, \add{or using genetic algorithms zqz bastrikov2018}.
 Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large PPEs \cite{qian2018}.
@@
 The process of parameter estimation for the complex land models typically embedded within Earth System Models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that comprise the terrestrial biosphere (e.g, hydrology, snow, biogeochemistry, physiology, land management, etc.).
-The number of model parameters required to represent all the relevant processes is large and model simulations are relatively expensive, presenting a major barrier to robust objective calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
+The number of model parameters required to represent all the relevant processes is large and model simulations are relatively expensive, presenting a major barrier to robust \remove{objective} calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
 Thorough assessment of the parametric sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}.
@@
 In this paper we present that dataset, describe how it was produced, and survey some potential applications.
-The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b,zarakas2024}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and uncertainties within CLM.
+The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b,zarakas2024}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and \remove{uncertainties} \add{sensitivities} within CLM.
 
@@
 The remaining 66 parameters include a plant functional type (PFT) dimension, with the potential for different parameter values for each unique plant functional type.
-Not all parameters are currently utilizing this functionality, such as \textit{medlynintercept} where the default value is 100 $\mu$mol H$_2$O m$^{-2}$ s$^{-1}$ for all sixteen of CLM's natural vegetation PFTs.
+\add{For any given PFT parameter, all PFTs were perturbed in tandem (i.e.
+all 16 PFTs to their maximum values and then to their minimum values), but with potentially different ranges for each PFT.
+Explicit parameter ranges for each parameter (and each PFT) are provided in a supplemental spreadsheet.} \remove{Not all parameters are currently utilizing this functionality, such as \textit{medlynintercept} where the default value is 100 $\mu$mol H$_2$O m$^{-2}$ s$^{-1}$ for all sixteen of CLM's natural vegetation PFTs.}
 
@@
  \centering
- \begin{tabular}{l l c}
+ \begin{tabular}{l l c c}
  \hline
-  Parameter  & Description & Model Domain \\
+  Parameter  & Description & Model Domain & Varies by PFT? \\
  \hline
-cv & Turbulent transfer coefficient & Sensible, latent heat and momentum fluxes \\
-dleaf & Leaf characteristic length & Sensible, latent heat and momentum fluxes \\
-d\_max & Dry surface layer (DSL) parameter & Sensible, latent heat and momentum fluxes \\
-frac\_sat\_soil\_dsl\_init & Fraction of saturated soil at which DSL initiates & Sensible, latent heat and momentum fluxes \\
-fff & Decay factor for fractional saturated area & Hydrology \\
-liq\_canopy\_storage\_scalar & Canopy-storage-of-liquid-water parameter & Hydrology \\
-maximum\_leaf\_wetted\_fraction & Maximum leaf wetted fraction & Hydrology \\
-sand\_pf & Perturbation factor for sand content of soil column & Hydrology \\
-sucsat\_sf & Scale factor for minimum soil water potential & Hydrology \\
-medlynintercept & Medlyn intercept of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
-medlynslope & Medlyn slope of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
-tpu25ratio & Ratio of tpu25top to vcmax25top & Stomatal resistance and photosynthesis \\
-jmaxb0 & Baseline proportion of nitrogen allocated for electron transport & Photosynthetic capacity\\
-jmaxb1 & Response of electron transport rate to light & Photosynthetic capacity\\
-slatop & Specific leaf area at top of canopy & Photosynthetic capacity\\
-theta\_cj & empirical curvature parameter for photosynthesis colimitation& Photosynthetic capacity \\
-wc2wjb0 & Baseline ratio of wc:wj & Photosynthetic capacity \\
-FUN\_fracfixers & fraction of carbon available for N fixation & Fixation and uptake of Nitrogen \\
-KCN & Nitrogen acquisition parameter group & Fixation and uptake of Nitrogen \\
-kmax & Plant segment max conductance & Plant hydraulics \\
-krmax & Root segment max conductance & Plant hydraulics \\
-psi50 & Water potential at 50\% loss of conductance & Plant hydraulics \\
-nstem & Stem number & Biomass heat storage \\
-lmr\_intercept\_atkin & Intercept in the calculation of leaf maintenance respiration& Plant respiration \\
-froot\_leaf & Allocation parameter: new fine root C per new leaf C & Carbon and nitrogen allocation \\
-leafcn & Leaf C:N & Carbon and nitrogen allocation \\
-leaf\_long & Leaf longevity & Vegetation phenology and turnover \\
-cpha & Activation energy for cp & Acclimation parameters \\
-jmaxhd & Deactivation energy for jmax & Acclimation parameters \\
-kcha & Activation energy for kc & Acclimation parameters \\
-lmrha & Activation energy for lmr & Acclimation parameters \\
-lmrhd & Deactivation energy for lmr & Acclimation parameters \\
-tpuha & Activation energy for tpu & Acclimation parameters \\
-tpuse\_sf & Scale factor for tpu entropy term & Acclimation parameters \\
-vcmaxha & Activation energy for vcmax & Acclimation parameters \\
-vcmaxhd & Deactivation energy for vcmax & Acclimation parameters \\
+cv & Turbulent transfer coefficient & Sensible, latent heat and momentum fluxes & \\
+dleaf & Leaf characteristic length & Sensible, latent heat and momentum fluxes & yes \\
+d\_max & Dry surface layer (DSL) parameter & Sensible, latent heat and momentum fluxes &\\
+frac\_sat\_soil\_dsl\_init & Fraction of saturated soil at which DSL initiates & Sensible, latent heat and momentum fluxes &\\
+fff & Decay factor for fractional saturated area & Hydrology &\\
+liq\_canopy\_storage\_scalar & Canopy-storage-of-liquid-water parameter & Hydrology &\\
+maximum\_leaf\_wetted\_fraction & Maximum leaf wetted fraction & Hydrology &\\
+sand\_pf & Perturbation factor for sand content of soil column & Hydrology &\\
+sucsat\_sf & Scale factor for minimum soil water potential & Hydrology &\\
+medlynintercept & Medlyn intercept of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis &yes\\
+medlynslope & Medlyn slope of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis &yes\\
+tpu25ratio & Ratio of tpu25top to vcmax25top & Stomatal resistance and photosynthesis &\\
+jmaxb0 & Baseline proportion of nitrogen allocated for electron transport & Photosynthetic capacity&\\
+jmaxb1 & Response of electron transport rate to light & Photosynthetic capacity&\\
+slatop & Specific leaf area at top of canopy & Photosynthetic capacity&yes\\
+theta\_cj & Empirical curvature parameter for photosynthesis colimitation& Photosynthetic capacity &yes\\
+wc2wjb0 & Baseline ratio of wc:wj & Photosynthetic capacity &\\
+FUN\_fracfixers & fraction of carbon available for N fixation & Fixation and uptake of Nitrogen &yes\\
+KCN & Nitrogen acquisition parameter group & Fixation and uptake of Nitrogen &yes\\
+kmax & Plant segment max conductance & Plant hydraulics &yes\\
+krmax & Root segment max conductance & Plant hydraulics &yes\\
+psi50 & Water potential at 50\% loss of conductance & Plant hydraulics &yes\\
+nstem & Stem number & Biomass heat storage &yes\\
+lmr\_intercept\_atkin & Intercept in the calculation of leaf maintenance respiration& Plant respiration &yes\\
+froot\_leaf & Allocation parameter: new fine root C per new leaf C & Carbon and nitrogen allocation &yes\\
+leafcn & Leaf C:N & Carbon and nitrogen allocation &yes\\
+leaf\_long & Leaf longevity & Vegetation phenology and turnover &yes\\
+cpha & Activation energy for cp & Acclimation parameters &\\
+jmaxhd & Deactivation energy for jmax & Acclimation parameters &\\
+kcha & Activation energy for kc & Acclimation parameters &\\
+lmrha & Activation energy for lmr & Acclimation parameters &\\
+lmrhd & Deactivation energy for lmr & Acclimation parameters &\\
+tpuha & Activation energy for tpu & Acclimation parameters &\\
+tpuse\_sf & Scale factor for tpu entropy term & Acclimation parameters &\\
+vcmaxha & Activation energy for vcmax & Acclimation parameters &\\
+vcmaxhd & Deactivation energy for vcmax & Acclimation parameters &\\
  \hline
@@
 Therefore the majority of simulations contain the erroneous parameter values, and may slightly overestimate GPP.
-The bug has since been resolved in later model versions.
+\add{The interacting effect of this bug fix with all of the other parameter perturbations was not explored.} The bug has since been resolved in later model versions \add{and will be tested in future PPEs}.
 
@@
 
-This model configuration allowed for over 2000 simulations across our six forcing scenarios.
-Overall, we found substantial impacts of parametric uncertainty on model behavior, in some cases exceeding the magnitude of scenario effects (Figure \ref{fig:ranges}).
+This model configuration allowed for over \remove{2000} \add{2500} simulations across our six forcing scenarios.
+Overall, we found substantial impacts of parametric \remove{uncertainty} \add{sensitivity} on model behavior, in some cases exceeding the magnitude of scenario effects (Figure \ref{fig:ranges}).
 In the control scenario, the global sum of gross primary production (GPP) ranged from 103 to 158 PgC per year across all ensemble members, and net ecosystem production (NEP) ranged from 1.7 to 3.8 PgC per year.
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/ranges.png}
-\caption{Global annual gross primary production (GPP) and net ecosystem production (NEP) across six forcing scenarios.
-Circles mark the CLM5.1 default simulation, and the bars span the ensemble ranges generated by individual parameter perturbations.}
+\includegraphics[width=\textwidth]{../figs/main/ranges.png}
+\caption{\add{Global annual mean values for the default simulations (circles) and ensemble ranges (bars) for} (a) gross primary production (GPP), (b) net ecosystem production (NEP), \add{(c) leaf area index (LAI), and (d) evapotranspiration (ET)} across six forcing scenarios.
+\add{(e) Global LAI across the full set of parameter perturbation simulations under the future climate scenario.
+Most simulations fall within 5\% of the default LAI, with a small number of parameters generating larger effects.
+One perturbation, reducing sand heat capacity by 20\% caused a very low LAI.} \remove{Circles mark the CLM5.1 default simulation, and the bars span the ensemble ranges generated by individual parameter perturbations.}}
 \label{fig:ranges}
@@
 Defining appropriate ranges for the full suite of land model parameters is a major activity of the scientific subdomains that comprise land models \cite{kattge2020}.
-Even with parameters that have an empirical basis from field studies, uncertainty quantification is difficult to the scale mismatch between site-level observations and land model grid cells \cite{liu2021}.
+Even with parameters that have an empirical basis from field studies, uncertainty quantification is difficult due to the scale mismatch between site-level observations and land model grid cells zqz \cite{liu2021}.
 As such, we cannot claim that the parameters are equivalently sampled, whereby parameter effects and parameter rankings could be subject to sampling asymmetries.
@@
 We found that parameter effects could be quite large, in some cases exceeding the effects of our forcing scenarios (Figure \ref{fig:ranges}).
-That said, the majority of parameter perturbations had small effects for any specific model variable or metric, such that a majority of simulations were clustered around the default simulation.
+That said, the majority of parameter perturbations had small effects for any specific model variable or metric, such that a majority of simulations were clustered around the default simulation (Figure \ref{fig:ranges}e).
 In the case of canopy evaporation, for example, more than 95\% of the ensemble variance could be explained by the ten most influential parameters (Figure \ref{fig:variance}).
@@
 As such, it is easy to diagnose which parameters are most influential on a given process.
-We have published a large set of ensemble diagnostic plots online, which serves as a valuable enhancement to our model technical documentation.
+We have published a large set of ensemble diagnostic plots online \add{(url:}\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT/}\add{)}, which serves as a valuable enhancement to our model technical documentation.
 Now, in addition to seeing the definition of a given parameter, and the relevant equations, a model user can easily investigate the magnitude and spatial patterns of its effects (e.g.

### 2025-04-09 ms/ms.tex
overleaf

@@
 We likewise opted to perturb soil hydraulics directly in two distinct ways, first by directly perturbing the hydraulic properties (e.g.
-saturated hydraulic conductivity) and also then also in a parallel set of simulations perturbing the soil texture (percent sand, silt, and clay).
+saturated hydraulic conductivity) and also \remove{then also} in a parallel set of simulations perturbing the soil texture (percent sand, silt, and clay) \add{and allowing the soil hydraulic properties to respond through the pedotransfer functions}.
 
@@
 The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, as it is the default forcing data for CLM5 \cite{lawrence2019}.
-Our six forcing scenarios were: a 2010-era control climate, high and low CO$_2$, future and pre-industrial climate anomaly forcing, and enhanace nitrogen deposition (see Table \ref{tab:exps}).
+Our six forcing scenarios were: a 2010-era control climate, high and low CO$_2$, future and pre-industrial climate anomaly forcing, and \add{enhanced} nitrogen deposition (see Table \ref{tab:exps}).
 We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
@@
 Overall there are nine vegetated biomes and an additional ice sheet biome, where no vegetation is present (Supp Figures S9,10).
+\add{These biomes are somewhat arbitrary, and the biomes would vary slightly based on the climatology period, but they provide some coarse delineation that we feel has meaningful ecological significance and have found useful for analyzing this ensemble.}
 
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/variance.png}
+\includegraphics[width=\textwidth]{../figs/main/variance.png}
 \caption{Cumulative fraction of variance explained by the most influential parameters on gross primary production and canopy evaporation.
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/lai_biome.pdf}
+\includegraphics[width=\textwidth]{../figs/main/lai_biome.pdf}
 \caption{The eight most influential parameters \add{(ranked by absolute parameter effect)} on leaf area index (LAI) within the control ensemble, globally and within three biomes.

### 2025-04-09 ms/ms.tex
new references

@@
 Understanding the range of outcomes arising from the many plausible parameter combinations is a critical step in robust uncertainty quantification.
-Parameter sensitivity tests can be used to gauge parametric uncertainty, and a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \add{zqzbaker2022} \cite{dagon2020,mcneall2024}.
+Parameter sensitivity tests can be used to gauge parametric uncertainty, and a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \add{new citation} \cite{dagon2020,baker2022,mcneall2024}.
 One important application of PPEs has been assessing uncertainty in climate model projections \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through various approaches like history matching, \add{which does not require assumptions regarding parameter distributions} \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024}, or Bayesian methods \add{zqz zeihn2012,fer2018} \cite{cleary2021}, \add{or using genetic algorithms zqz bastrikov2018}.
+Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through various approaches like history matching, \add{which does not require assumptions regarding parameter distributions} \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024}, or Bayesian methods \add{new citations} \cite{ziehn2012,fer2018,cleary2021}, \add{or using genetic algorithms} \cite{bastrikov2018}.
 Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large PPEs \cite{qian2018}.
@@
 Defining appropriate ranges for the full suite of land model parameters is a major activity of the scientific subdomains that comprise land models \cite{kattge2020}.
-Even with parameters that have an empirical basis from field studies, uncertainty quantification is difficult due to the scale mismatch between site-level observations and land model grid cells zqz \cite{liu2021}.
+Even with parameters that have an empirical basis from field studies, uncertainty quantification is difficult due to the scale mismatch between site-level observations and land model grid cells \cite{liu2021}.
 As such, we cannot claim that the parameters are equivalently sampled, whereby parameter effects and parameter rankings could be subject to sampling asymmetries.

### 2025-04-15 ms/ms.tex
overleaf

@@
 Making centennial-scale projections of the cumulative terrestrial carbon sink has been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2006,friedlingstein2014,arora2020}, and limited efficacy of emergent constraints \cite{liu2024}.
-A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
+\add{Emissions trajectories and land management scenarios contribute some uncertainty, but the majority} \remove{A portion} of this uncertainty is \remove{irreducible,} inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
 Still, our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns \cite{worden2021,baldocchi2024}.
-The capability to efficiently ingest these data and improve simulation performance is especially valuable for actionable science \cite{cheng2023}.
+The capability to efficiently ingest these data and improve simulation performance \add{with improved parameters} is especially valuable for actionable science \cite{cheng2023}.
 However, given increasingly comprehensive land modeling systems, and the wide array of observational products, several technical hurdles exist that hinder effective model development and calibration.
@@
 The parameter range spreadsheet is provided within the Supplementary Information.
+
 In some cases parameters could not or should not be perturbed independently, because they feature inherent co-dependence.
 In the case of nitrogen fixation costs, we opted to perturb the parameters independently, but also as a group (`KCN'), to enforce a change in overall nitrogen limitation in lieu of switching from one uptake pathway to another.
-We likewise opted to perturb soil hydraulics directly in two distinct ways, first by directly perturbing the hydraulic properties (e.g.
+\add{For the grouped simulations, the given set of parameters were perturbed in tandem, either to their minimum or maximum values.} We likewise opted to perturb soil hydraulics directly in two distinct ways, first by directly perturbing the hydraulic properties (e.g.
 saturated hydraulic conductivity) and also \remove{then also} in a parallel set of simulations perturbing the soil texture (percent sand, silt, and clay) \add{and allowing the soil hydraulic properties to respond through the pedotransfer functions}.
+\add{Future experiments could utilize multifactor perturbation sampling strategies to resolve parameter interdependencies and tradeoffs, such as along major axes or ecosystem function} \cite{migliavacca2021}.
 
@@
 Conversely, canopy evaporation is governed by fewer parameters.
-Just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S3).
+Just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S\remove{3}\add{11}).
 In general, most of the parameter perturbations had a relatively small impact on any given output variable.
@@
 \caption{The eight most influential parameters \add{(ranked by absolute parameter effect)} on NEP (top row) in the control simulations (a), high CO$_2$ simulations (b), and the relative response to high CO$_2$ (c), as well as on evapotranspiration (ET, bottom row) in the control simulations (d), future climate simulations (e) and the relative response to future climate (f).
+The solid lines span the range between the two simulations for each parameter, with an open circle indicating the simulation where a parameter is set to its minimum value and a filled circle for its maximum.
+The lines are color-coded according to scientific domain.
+The dashed lines indicate the value in the simulation with default CLM5.1 parameters.
 The parameters that are most influential on present-day climate (a,d) can differ significantly from the parameters that control the response to forcing perturbations (c,f).}
@@
 \caption{The eight most influential parameters \add{(ranked by absolute parameter effect)} on leaf area index (LAI) within the control ensemble, globally and within three biomes.
-The solid lines span the range between the two simulations for each parameter, with an open circle indicating the simulation where a parameter is set to its minimum value and a filled circle for its maximum.
-The lines are color-coded according to scientific domain.
-The dashed lines indicate the LAI for the simulation with default CLM5.1 parameters.
-Parameter rankings vary by biome, with the global rankings seeming to reflect contributions from each of these three biomes.}
+\add{Lines and circles are as described in Figure 6.} Parameter rankings vary by biome, with the global rankings seeming to reflect contributions from each of these three biomes.}
 \label{fig:lai}
@@
 We found that many acclimation parameters, which were not as important for determining present-day evapotranspiration (ET), were among the most influential on the response of ET to future climate.
-This emphasizes the importance of testing models not only under mean-state conditions, but also under experimentally perturbed conditions under global change analogues \cite{wieder2019}.
+This emphasizes the value of testing models not only under mean-state conditions, but also under experimentally perturbed conditions under global change analogues \cite{wieder2019}.
+\add{These acclimation parameters likewise would not be constrained by observations of mean ET, indicating that alternative metrics may be required or that we may lack capacity to reduce projection uncertainties associated with these parameters.}
+
 Furthermore, because parameter values in CLM are time-invariant, plants must be able to survive pre-industrial conditions in order to be present for the later stages of a transient simulation, such as those performed for many model intercomparison experiments.

### 2025-04-15 ms/ms.tex
overleaf

@@
 saturated hydraulic conductivity) and also \remove{then also} in a parallel set of simulations perturbing the soil texture (percent sand, silt, and clay) \add{and allowing the soil hydraulic properties to respond through the pedotransfer functions}.
-\add{Future experiments could utilize multifactor perturbation sampling strategies to resolve parameter interdependencies and tradeoffs, such as along major axes or ecosystem function} \cite{migliavacca2021}.
+\add{Future experiments could utilize a multifactor perturbation sampling strategy to resolve parameter interdependencies and tradeoffs, such as along major axes of ecosystem function} \cite{migliavacca2021}.
 
@@
 
+\add{We opted to rank parameters by absolute parameter effect, given that perturbations while unequal in relative size, were defined with the same uncertainty range in mind.} 
 We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps} (Figure \ref{fig:ranges}).

### 2025-04-15 ms/ms.tex
moving figs to ms directory for james latex

--- 2025-04-09 ms/ms.tex
@@
 Making centennial-scale projections of the cumulative terrestrial carbon sink has been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2006,friedlingstein2014,arora2020}, and limited efficacy of emergent constraints \cite{liu2024}.
-A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
+Emissions trajectories and land management scenarios contribute some uncertainty, but the majority of this uncertainty is inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{lovenduski2017}.
 Still, our expectation is that some of this uncertainty is indeed reducible, if we can more effectively utilize the ongoing expansion of observational data sources from remote sensing, meteorological stations, flux towers, and field campaigns \cite{worden2021,baldocchi2024}.
-The capability to efficiently ingest these data and improve simulation performance is especially valuable for actionable science \cite{cheng2023}.
+The capability to efficiently ingest these data and improve simulation performance with improved parameters is especially valuable for actionable science \cite{cheng2023}.
 However, given increasingly comprehensive land modeling systems, and the wide array of observational products, several technical hurdles exist that hinder effective model development and calibration.
@@
 Understanding the range of outcomes arising from the many plausible parameter combinations is a critical step in robust uncertainty quantification.
-Parameter sensitivity tests can be used to gauge parametric uncertainty, and a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \add{new citation} \cite{dagon2020,baker2022,mcneall2024}.
+Parameter sensitivity tests can be used to gauge parametric uncertainty, and a collection of systematic parameter sensitivity tests across multiple parameters is often termed a Perturbed Parameter Ensemble (PPE, also referred to as Perturbed Physics Ensemble), with examples in the coupled \cite{murphy2004} and land-only contexts \cite{dagon2020,baker2022,mcneall2024}.
 One important application of PPEs has been assessing uncertainty in climate model projections \cite{murphy2004,sanderson2008,booth2012,hawkins2019,yamazaki2021,peatier2022,tett2022}.
-Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through various approaches like history matching, \add{which does not require assumptions regarding parameter distributions} \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024}, or Bayesian methods \add{new citations} \cite{ziehn2012,fer2018,cleary2021}, \add{or using genetic algorithms} \cite{bastrikov2018}.
+Beyond uncertainty quantification, PPEs also form a basis for automated model calibration through various approaches like history matching, which does not require assumptions regarding parameter distributions \cite{williamson2013,williamson2017,hourdin2020,couvreux2021,mcneall2024}, or Bayesian methods \cite{ziehn2012,fer2018,cleary2021}, or using genetic algorithms \cite{bastrikov2018}.
 Several promising avenues for model calibration are in development \cite{pinnington2020,cleary2021,alonso-gonzalez2022}, many of which require the construction of large PPEs \cite{qian2018}.
@@
 The process of parameter estimation for the complex land models typically embedded within Earth System Models is challenging on account of the intrinsic complexity of the heterogeneous land surface, the diversity and plasticity of plant and microbial life, and the multiplicity of domains that comprise the terrestrial biosphere (e.g, hydrology, snow, biogeochemistry, physiology, land management, etc.).
-The number of model parameters required to represent all the relevant processes is large and model simulations are relatively expensive, presenting a major barrier to robust \remove{objective} calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
+The number of model parameters required to represent all the relevant processes is large and model simulations are relatively expensive, presenting a major barrier to robust calibration, model interpretation, and development \cite{fisher2020,dagon2020}.
 Thorough assessment of the parametric sensitivity of land models in a manner that is repeatable, open, and integrated with ongoing code development is desirable but challenging \cite{hourdin2017,balaji2022}.
@@
 In this paper we present that dataset, describe how it was produced, and survey some potential applications.
-The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b,zarakas2024}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and \remove{uncertainties} \add{sensitivities} within CLM.
+The dataset has already demonstrated utility for diagnosing parameter effects \cite{cheng2023,yan2023a,yan2023b,zarakas2024}, while the software and modeling infrastructure has greatly expanded our capability to generate insights about parameter effects and sensitivities within CLM.
 
@@
 To recompose mapped output and compute global means, the output from the representative grid cell is substituted for all members of the cluster cohort.
+This naive interpolation will introduce errors that grow in relative magnitude as the domain of interest shrinks.
 
@@
 The remaining 66 parameters include a plant functional type (PFT) dimension, with the potential for different parameter values for each unique plant functional type.
-\add{For any given PFT parameter, all PFTs were perturbed in tandem (i.e.
+For any given PFT parameter, all PFTs were perturbed in tandem (i.e.
 all 16 PFTs to their maximum values and then to their minimum values), but with potentially different ranges for each PFT.
-Explicit parameter ranges for each parameter (and each PFT) are provided in a supplemental spreadsheet.} \remove{Not all parameters are currently utilizing this functionality, such as \textit{medlynintercept} where the default value is 100 $\mu$mol H$_2$O m$^{-2}$ s$^{-1}$ for all sixteen of CLM's natural vegetation PFTs.}
+Explicit parameter ranges for each parameter (and each PFT) are provided in a supplemental spreadsheet.
 
@@
 The parameter range spreadsheet is provided within the Supplementary Information.
+
 In some cases parameters could not or should not be perturbed independently, because they feature inherent co-dependence.
 In the case of nitrogen fixation costs, we opted to perturb the parameters independently, but also as a group (`KCN'), to enforce a change in overall nitrogen limitation in lieu of switching from one uptake pathway to another.
+For the grouped simulations, the given set of parameters were perturbed in tandem, either to their minimum or maximum values.
 We likewise opted to perturb soil hydraulics directly in two distinct ways, first by directly perturbing the hydraulic properties (e.g.
-saturated hydraulic conductivity) and also \remove{then also} in a parallel set of simulations perturbing the soil texture (percent sand, silt, and clay) \add{and allowing the soil hydraulic properties to respond through the pedotransfer functions}.
+saturated hydraulic conductivity) and also in a parallel set of simulations perturbing the soil texture (percent sand, silt, and clay) and allowing the soil hydraulic properties to respond through the pedotransfer functions.
+Future experiments could utilize a multifactor perturbation sampling strategy to resolve parameter interdependencies and tradeoffs, such as along major axes of ecosystem function \cite{migliavacca2021}.
 
@@
 Therefore the majority of simulations contain the erroneous parameter values, and may slightly overestimate GPP.
-\add{The interacting effect of this bug fix with all of the other parameter perturbations was not explored.} The bug has since been resolved in later model versions \add{and will be tested in future PPEs}.
+The interacting effect of this bug fix with all of the other parameter perturbations was not explored.
+The bug has since been resolved in later model versions and will be tested in future PPEs.
 
@@
 The GSWP3v1 reanalysis product (\url{http://hydro.iis.u-tokyo.ac.jp/GSWP3/}) served as our atmospheric forcing, as it is the default forcing data for CLM5 \cite{lawrence2019}.
-Our six forcing scenarios were: a 2010-era control climate, high and low CO$_2$, future and pre-industrial climate anomaly forcing, and \add{enhanced} nitrogen deposition (see Table \ref{tab:exps}).
+Our six forcing scenarios were: a 2010-era control climate, high and low CO$_2$, future and pre-industrial climate anomaly forcing, and enhanced nitrogen deposition (see Table \ref{tab:exps}).
 We applied climate and CO$_2$ anomalies independently, in order to disentangle their effects on parameter rankings.
@@
 Overall there are nine vegetated biomes and an additional ice sheet biome, where no vegetation is present (Supp Figures S9,10).
-\add{These biomes are somewhat arbitrary, and the biomes would vary slightly based on the climatology period, but they provide some coarse delineation that we feel has meaningful ecological significance and have found useful for analyzing this ensemble.}
+These biomes are somewhat arbitrary, and the biomes would vary slightly based on the climatology period, but they provide some coarse delineation that we feel has meaningful ecological significance and have found useful for analyzing this ensemble.
 
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/main/bar.png}
+\includegraphics[width=\textwidth]{bar.png}
 \caption{211 parameters were perturbed across the various domains of the land model.}
@@
 \end{figure}
+
+\subsection{Computational cost}
+
 Given the very large size of the CLM5-BGC parameter space (Figure \ref{fig:params}), and the relatively expensive computational cost of standard model configurations, a major goal of this project was to develop a fast configuration of CLM5-BGC that would allow for a large number of simulations.
@@
 \centering
-\includegraphics[width=30pc]{../figs/main/sims.png}
+\includegraphics[width=30pc]{sims.png}
 \caption{The approximate number of simulations afforded by 1 million core-hours on the Cheyenne supercomputer for a range of CLM configurations.
@@
 We had hoped to see a clear point of diminishing returns across a wide range of variables as we increase the number of clusters.
-Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, convergence slows and presents some small amplitude oscillations (Supp Figure S2).
+Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, improvement slows but does not saturate (Supp Figure S2).
 We chose 400 clusters, because it provides satisfactory performance across a range of important metrics, while affording a sufficient number of simulations to perform the full experiment within our computational budget.
+The main goal of the sparsegrid is to obtain a coarse understanding of parameter effects at low computational cost, which could later be improved by full grid simulations or alternative sparsegrids with more clusters.
+
 \begin{figure}[h]
 \centering
-\includegraphics[width=25pc]{../figs/main/sparsegrid_gpp.pdf}
+\includegraphics[width=25pc]{sparsegrid_gpp.pdf}
 \caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual gross primary production (GPP) across the last forty years of a transient CLM5.1 simulation.
@@
 
-This model configuration allowed for over \remove{2000} \add{2500} simulations across our six forcing scenarios.
-Overall, we found substantial impacts of parametric \remove{uncertainty} \add{sensitivity} on model behavior, in some cases exceeding the magnitude of scenario effects (Figure \ref{fig:ranges}).
+This model configuration allowed for over 2500 simulations across our six forcing scenarios.
+Overall, we found substantial impacts of parametric sensitivity on model behavior, in some cases exceeding the magnitude of scenario effects (Figure \ref{fig:ranges}).
 In the control scenario, the global sum of gross primary production (GPP) ranged from 103 to 158 PgC per year across all ensemble members, and net ecosystem production (NEP) ranged from 1.7 to 3.8 PgC per year.
@@
 
+\subsection{Main results}
+
 A small fraction of parameters tends to explain a large amount of the ensemble variance for any given variable or metric (Figure \ref{fig:variance}).
@@
 Conversely, canopy evaporation is governed by fewer parameters.
-Just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S3).
+Just four parameters (maximum leaf wetted fraction, liquid canopy storage, the turbulent transfer coefficient, and leaf characteristic length) explain upwards of 90\% of the canopy evaporation variance across all the forcing scenarios (Supp Fig S11).
 In general, most of the parameter perturbations had a relatively small impact on any given output variable.
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/main/ranges.png}
-\caption{\add{Global annual mean values for the default simulations (circles) and ensemble ranges (bars) for} (a) gross primary production (GPP), (b) net ecosystem production (NEP), \add{(c) leaf area index (LAI), and (d) evapotranspiration (ET)} across six forcing scenarios.
-\add{(e) Global LAI across the full set of parameter perturbation simulations under the future climate scenario.
+\includegraphics[width=\textwidth]{ranges.png}
+\caption{Global annual mean values for the default simulations (circles) and ensemble ranges (bars) for (a) gross primary production (GPP), (b) net ecosystem production (NEP), (c) leaf area index (LAI), and (d) evapotranspiration (ET) across six forcing scenarios.
+(e) Global LAI across the full set of parameter perturbation simulations under the future climate scenario.
 Most simulations fall within 5\% of the default LAI, with a small number of parameters generating larger effects.
-One perturbation, reducing sand heat capacity by 20\% caused a very low LAI.} \remove{Circles mark the CLM5.1 default simulation, and the bars span the ensemble ranges generated by individual parameter perturbations.}}
+One perturbation, reducing sand heat capacity by 20\% caused a very low LAI.
+}
 \label{fig:ranges}
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/main/variance.png}
+\includegraphics[width=\textwidth]{variance.png}
 \caption{Cumulative fraction of variance explained by the most influential parameters on gross primary production and canopy evaporation.
@@
 
+\subsection{Comparing scenarios and biomes}
+
+We opted to rank parameters by absolute parameter effect, given that perturbations while unequal in relative size, were defined with the same uncertainty range in mind.
 We repeated the full set of parameter perturbations across the six forcing scenarios in Table \ref{tab:exps} (Figure \ref{fig:ranges}).
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/main/deltas.png}
-\caption{The eight most influential parameters \add{(ranked by absolute parameter effect)} on NEP (top row) in the control simulations (a), high CO$_2$ simulations (b), and the relative response to high CO$_2$ (c), as well as on evapotranspiration (ET, bottom row) in the control simulations (d), future climate simulations (e) and the relative response to future climate (f).
+\includegraphics[width=\textwidth]{deltas.png}
+\caption{The eight most influential parameters (ranked by absolute parameter effect) on NEP (top row) in the control simulations (a), high CO$_2$ simulations (b), and the relative response to high CO$_2$ (c), as well as on evapotranspiration (ET, bottom row) in the control simulations (d), future climate simulations (e) and the relative response to future climate (f).
+The solid lines span the range between the two simulations for each parameter, with an open circle indicating the simulation where a parameter is set to its minimum value and a filled circle for its maximum.
+The lines are color-coded according to scientific domain.
+The dashed lines indicate the value in the simulation with default CLM5.1 parameters.
 The parameters that are most influential on present-day climate (a,d) can differ significantly from the parameters that control the response to forcing perturbations (c,f).}
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/main/lai_biome.pdf}
-\caption{The eight most influential parameters \add{(ranked by absolute parameter effect)} on leaf area index (LAI) within the control ensemble, globally and within three biomes.
-The solid lines span the range between the two simulations for each parameter, with an open circle indicating the simulation where a parameter is set to its minimum value and a filled circle for its maximum.
-The lines are color-coded according to scientific domain.
-The dashed lines indicate the LAI for the simulation with default CLM5.1 parameters.
+\includegraphics[width=\textwidth]{lai_biome.pdf}
+\caption{The eight most influential parameters (ranked by absolute parameter effect) on leaf area index (LAI) within the control ensemble, globally and within three biomes.
+Lines and circles are as described in Figure 6.
 Parameter rankings vary by biome, with the global rankings seeming to reflect contributions from each of these three biomes.}
@@
 \centering
-\includegraphics[width=\textwidth]{../figs/QRUNOFFabs_x_medlynslope_CTL2010.png}
+\includegraphics[width=\textwidth]{QRUNOFFabs_x_medlynslope_CTL2010.png}
 \caption{Map of the effect of perturbing the medlynslope stomatal conductance parameter on runoff within the control ensemble.
@@
 
+Our sparsegrid formulation allows us to run far fewer gridcells than a standard rectilinear grid, but at the cost of model fidelity.
+We visualized interannual variations in global GPP for different sparsegrids (Figure \ref{fig:sg}) to demonstrate the tradeoff between number of clusters and representational accuracy.
+A wider array of scoring metrics (Supp Fig S3, using the ILAMB2.5 scoring methodologies, Collier et al.
+2018) shows that scores continue to improve beyond 400 clusters (our choice for these experiments).
+For example bias scores for total carbon stocks are consistently lower than other variables and may require a larger number of clusters or an alternative clustering strategy that accounts for the outsized influence of high latitude grid cells.
+The sparsegrid configuration will generate a coarse understanding of parameter effects at low computational cost, that can be honed by iteratively increasing the number of clusters or by supplementing with full grid simulations.
+
 For this experiment we opted for a one-at-a-time perturbation strategy, testing a minimum and maximum value for each parameter.
@@
 We found that many acclimation parameters, which were not as important for determining present-day evapotranspiration (ET), were among the most influential on the response of ET to future climate.
-This emphasizes the importance of testing models not only under mean-state conditions, but also under experimentally perturbed conditions under global change analogues \cite{wieder2019}.
+This emphasizes the value of testing models not only under mean-state conditions, but also under experimentally perturbed conditions under global change analogues \cite{wieder2019}.
+These acclimation parameters likewise would not be constrained by observations of mean ET, indicating that alternative metrics may be required or that we may lack capacity to reduce projection uncertainties associated with these parameters.
+
 Furthermore, because parameter values in CLM are time-invariant, plants must be able to survive pre-industrial conditions in order to be present for the later stages of a transient simulation, such as those performed for many model intercomparison experiments.
@@
 A one-at-a-time PPE cannot capture parameter interactions, and our min/max sampling protocol precludes diagnosing non-linearities.
-As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty.
+As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty \cite{raoult2024}.
 A primary utility of our dataset is that we can diagnose parameter effects without the uncertainty contributed by an emulator or a regression model.
 As such, it is easy to diagnose which parameters are most influential on a given process.
-We have published a large set of ensemble diagnostic plots online \add{(url:}\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT/}\add{)}, which serves as a valuable enhancement to our model technical documentation.
+We have published a large set of ensemble diagnostic plots online (\url{https://webext.cgd.ucar.edu/I2000/PPEn11_OAAT/}), which serves as a valuable enhancement to our model technical documentation.
 Now, in addition to seeing the definition of a given parameter, and the relevant equations, a model user can easily investigate the magnitude and spatial patterns of its effects (e.g.
@@
 This could lead to biased estimates of water and energy fluxes, as atmospheric feedbacks can modify the state of the land surface itself \cite{lague2019}.
+Likewise, fire parameters, which were not among the most important parameters for global ET, NEP, or LAI, may still influence the coupled model when coupled fire emissions are allowed to modify aerosol loads.
 Additional work is ongoing to extend this work in a coupled model framework to quantify how much, where, and for which processes atmospheric feedbacks are likely to modulate land parameter effects substantially.
@@
 Our expectation is that relative parameter rankings are likely robust, but that the absolute value of parameter effects may be significantly biased relative to their effects in the coupled model.
+
+\section{Conclusion}
 
@@
 The software, datasets, and analysis tools generated to enable and utilize this experiment will greatly reduce the burden of generating future PPEs and parameter sensitivity experiments.
+Due to the large parameter space of comprehensive land models like CLM, reducing model computational costs can expand the scope of feasible parameter estimation problems.
 All of the code to generate this ensemble is available via github.

### 2025-05-29 ms/ms_trackchanges.tex

--- 2025-04-15 ms/ms.tex
@@
 To recompose mapped output and compute global means, the output from the representative grid cell is substituted for all members of the cluster cohort.
+\add{This naive interpolation will introduce errors that grow in relative magnitude as the domain of interest shrinks.}
 
@@
 We had hoped to see a clear point of diminishing returns across a wide range of variables as we increase the number of clusters.
-Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, convergence slows and presents some small amplitude oscillations (Supp Figure S2).
+Instead, for most variables, we observe a quick reduction of bias approaching 200 clusters, but after this point, improvement slows \remove{and presents some small amplitude oscillations} \add{but does not saturate} (Supp Figure S2).
 We chose 400 clusters, because it provides satisfactory performance across a range of important metrics, while affording a sufficient number of simulations to perform the full experiment within our computational budget.
+\add{The main goal of the sparsegrid is to obtain a coarse understanding of parameter effects at low computational cost, which could later be improved by full grid simulations or alternative sparsegrids with more clusters.}
+
 \begin{figure}[h]
@@
 Carrying out an extensive PPE increases the possibility of exposing unexpected model behavior, including unforeseen tipping points, brittle parameterizations, and/or bugs.
+
+\subsection{Main results}
 
@@
 \end{figure}
+
+\subsection{Comparing scenarios and biomes}
 
@@
 
+\add{Our sparsegrid formulation allows us to run far fewer gridcells than a standard rectilinear grid, but at the cost of model fidelity.
+We visualized interannual variations in global GPP for different sparsegrids} (Figure \ref{fig:sg}) \add{to demonstrate the tradeoff between number of clusters and representational accuracy.
+A wider array of scoring metrics (Supp Fig S3, using the ILAMB2.5 scoring methodologies, Collier et al.
+2018) shows that scores continue to improve beyond 400 clusters (our choice for these experiments).
+For example bias scores for total carbon stocks are consistently lower than other variables and may require a larger number of clusters or an alternative clustering strategy that accounts for the outsized influence of high latitude grid cells.
+The sparsegrid configuration will generate a coarse understanding of parameter effects at low computational cost, that can be honed by iteratively increasing the number of clusters or by supplementing with full grid simulations.}
+
 For this experiment we opted for a one-at-a-time perturbation strategy, testing a minimum and maximum value for each parameter.
@@
 A one-at-a-time PPE cannot capture parameter interactions, and our min/max sampling protocol precludes diagnosing non-linearities.
-As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty.
+As such, this dataset will be insufficient for most calibration activities or for estimating overall parametric uncertainty \add{(new citation)} \cite{raoult2024}.
 A primary utility of our dataset is that we can diagnose parameter effects without the uncertainty contributed by an emulator or a regression model.
@@
 This could lead to biased estimates of water and energy fluxes, as atmospheric feedbacks can modify the state of the land surface itself \cite{lague2019}.
-Additional work is ongoing to extend this work in a coupled model framework to quantify how much, where, and for which processes atmospheric feedbacks are likely to modulate land parameter effects substantially.
+\add{Likewise, fire parameters, which were not among the most important parameters for global ET, NEP, or LAI, may still influence the coupled model when coupled fire emissions are allowed to modify aerosol loads.} Additional work is ongoing to extend this work in a coupled model framework to quantify how much, where, and for which processes atmospheric feedbacks are likely to modulate land parameter effects substantially.
 Furthermore our experiments are conducted at nominal 2-degree resolution, certain processes subject to land heterogeneity may respond differently at higher resolution, such as permafrost depth in the Arctic \cite{schickhoff2024}.
 Our expectation is that relative parameter rankings are likely robust, but that the absolute value of parameter effects may be significantly biased relative to their effects in the coupled model.
+
+\section{Conclusion}
 
@@
 The software, datasets, and analysis tools generated to enable and utilize this experiment will greatly reduce the burden of generating future PPEs and parameter sensitivity experiments.
+\add{Due to the large parameter space of comprehensive land models like CLM, reducing model computational costs can expand the scope of feasible parameter estimation problems.} 
 All of the code to generate this ensemble is available via github.

### 2025-05-29 ms/supp.tex

@@
 \includegraphics[width=60pc]{../figs/supp/ilamb_lines.png}
-\caption{The convergence of several scoring metrics with increasing sparsegrid resolution across a variety of CLM variables.
-All scores tend towards 1 as the number of sparsegrid clusters (k) approaches the 5666 gridcells for the native resolution of our 2-degree simulations.
+\caption{The convergence of several scoring metrics (based on the ILAMB2.5 benchmarking software, Collier et al.
+2018) with increasing sparsegrid resolution across a variety of CLM variables.
+All scores tend towards 1 as the number of sparsegrid clusters (k) approaches the 5666 grid cells for the native resolution of our 2-degree simulations.
 }
